Efficient coding explains neural response homeostasis and stimulus-specific adaptation

Version 1

Abstract

Abstract

In the absence of adaptation, the average firing rate of neurons would rise or drop when changes in the environment make their preferred stimuli more or less prevalent. However, by adjusting the responsiveness of neurons, adaptation can yield firing rate homeostasis and stabilise the average rates of neurons at fixed levels, despite changes in stimulus statistics. In sensory cortex, adaptation is typically also stimulus specific, in that neurons reduce their responsiveness to over-represented stimuli, but maintain or even increase their responsiveness to stimuli far from over-represented ones. Here, we present a normative explanation of firing rate homeostasis grounded in the efficient coding principle, showing that this homeostasis yields an optimal trade-off between coding fidelity and the metabolic cost of neural firing. Unlike previous efficient coding theories, we formulate the problem in a computation-agnostic manner, enabling our framework to apply far from the sensory periphery. We then apply this general framework to Distributed Distributional Codes, a specific computational theory of neural representations serving Bayesian inference. We demonstrate how homeostatic coding, combined with such Bayesian neural representations, provides a normative explanation for stimulus-specific adaptation, widely observed across the brain, and how this coding scheme can be accomplished by divisive normalisation with adaptive weights. Further, we develop a model within this combined framework, and by fitting it to previously published experimental data, quantitatively account for measures of stimulus-specific and homeostatic adaption in the primary visual cortex.

Introduction

Neural population responses, and thus the computations on sensory inputs represented by them, are corrupted by noise. The extent of this corruption can, however, be modulated by changing neural gains. When the gain of a neuron increases, the strength of its response noise, or trial-to-trial variability, typically grows sublinearly and more slowly than its average response; for example, for Poisson-like firing, response noise grows as the square root of the mean response. The neuron can therefore increase its signal-to-noise ratio by increasing its gain or responsiveness. However, this increase in coding fidelity comes at the cost of an elevated average firing rate, and thus higher metabolic energy expenditure. From a normative perspective, coding fidelity and metabolic cost are thus two conflicting forces.

In this article, we ask what is the optimal way of adjusting neural gains in order to combat noise with minimal metabolic cost, irrespective of the computations represented by the neural population. As we will see, the answer depends on the prevailing stimulus statistics. We will thus address the following more specific question: given a neural population with arbitrary tuning curve shapes and noise distribution, how should neurons optimally adjust their gains depending on the stimulus statistics prevailing in the environment?

To address this question, we use efficient coding theory as our normative framework (Attneave, 1954; Nadal and Parga, 1999; Laughlin, 1981; Barlow, 2012; Linsker, 1988; Ganguli and Simoncelli, 2014; Wei and Stocker, 2015; Atick and Redlich, 1990). Efficient coding theories formalise the problem of optimal coding subject to biological constraints and costs, and by finding the optimal solution make predictions about the behavior of nervous systems (which are postulated to have been approximately optimised, e.g., via natural selection). Concretely, the Infomax Principle (Linsker, 1988; Atick and Redlich, 1990; Ganguli and Simoncelli, 2014; Wei and Stocker, 2015) states that sensory systems optimise the mutual information between a noisy neural representation and an external stimulus (or alternatively, the ideal noise-free result of computations on external stimuli), subject to metabolic constraints. In order to do so, the neural system should exploit the statistics of stimuli in its local environment (Simoncelli and Olshausen, 2001). Efficient coding theories can therefore predict how sensory systems should optimally adapt to changes in environmental stimulus statistics.

Suppose an environmental shift makes a stimulus feature more prevalent. Neurons sensitive to that feature will then see their preferred feature more often, and thus their average firing rate will initially increase. However, typically, the neurons do gradually adapt by reducing their responsiveness or gain (Solomon and Kohn, 2014; Clifford et al., 2007; Benucci et al., 2013). A special case of such adaptation is firing rate homeostasis (Desai, 2003; Turrigiano and Nelson, 2004; Maffei and Turrigiano, 2008; Hengen et al., 2013), in which adaptation brings the average firing rate back to the level prior to the environmental shift (Fig. 1). Thus, under firing rate homeostasis, neuronal populations maintain a constant stimulus-averaged firing rate in spite of changes to the environment (Benucci et al., 2013; Hengen et al., 2013; Maffei and Turrigiano, 2008).

Homeostatic adaptation via gain modulation (cartoon example).

Each column of panels corresponds to an environment as defined by its stimulus distribution (bottom row). The top panels show the post-adaptation tuning curves of the same four neurons in these two environments. Neurons adapt only by independently adjusting their gains, i.e., by scaling their tuning curves up or down (correspondingly, tuning curves of the same color in the left and right columns, which belong to the same neuron, are scaled versions of each other and have identical shape). This adjustment is such that each neuron maintains the same stimulus-averaged firing rate in both environments. This is shown by the bar plots in the insets of the top row panels: the bar plot in the left column shows the average rates of the four neurons in the first environment, while the two bar plots in the right column show the average rates of these neurons before and after gain adaptation to the new statistics in the second environment. After adaptation the average rates return to their level before the environmental shift.

There are multiple levels at which homeostatic adaptation can be observed. Firstly, there is population homeostasis, in which the stimulus-average firing rate of an entire population of neurons remains constant, without individual neurons necessarily holding their rates constant (Slomowitz et al., 2015; Sanzeni et al., 2023). Secondly, there is what we term cluster homeostasis. In this form of homeostasis, stimulus-average firing rate of groups or clusters of neurons with similar stimulus tuning remains stable under environmental shifts (Benucci et al., 2013), but the firing rate of individual neurons within a cluster can change. Lastly, in the strongest form, homeostasis can occur at the level of individual neurons, in which case the firing rate of each individual neuron is kept constant under changes in environment statistics (Marder and Prinz, 2003). Previous normative explanations for firing rate homeostasis often focus on the necessity of preventing the network from becoming hypo- or hyperactive (Turrigiano and Nelson, 2004; Maffei and Turrigiano, 2008; Keck et al., 2013; Hengen et al., 2013). Although this might explain homeostasis at the population level, it does not adequately explain homeostasis at a more fine-grained level. As we will show in this paper, the answer to the normative question that we posed above provides an explanation for firing rate homeostasis at such a finer level. Specifically, we will show that, for a wide variety of neural noise models (and under conditions that we argue do obtain in the visual cortex), optimal adjustment of neural gains to combat the effect of coding noise predicts firing rate homeostasis at the level of clusters of similarly-tuned neurons.

Having shown the optimality of homeostatic codes without making assumptions on the nature of the neural representation and computation, we apply our framework to Distributed Distributional Codes (DDC) (Vertes and Sahani, 2018), as a specific computational theory for neural implementation of Bayesian inference. Combining DDC theory (which determines tuning curve shapes) with our normative results (on optimal adjustments of neural gains) yields what we call homeostatic DDC. We will show that homeostatic DDC predict the typical finding that sensory adaptations are stimulus specific (Kohn, 2007; Schwartz et al., 2007). In stimulus specific adaptation (SSA), the suppression of response is greater for test stimuli that are closer to an over-represented adaptor stimulus than for stimuli further away (to which the neuron may even respond more strongly after adaptation). In particular, this causes a repulsion of tuning curves away from the adaptor (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013). We will show that homeostatic DDC are able to account for SSA effects which cannot be fully accounted for in previous efficient coding frameworks (Wei and Stocker, 2015; Ganguli and Simoncelli, 2014; Snow et al., 2016), and we will use them to quantitatively model homeostatic SSA as observed in V1, by re-analysing previously published data (Benucci et al., 2013). This model relies on a special class of homeostatic DDC which we term Bayes-ratio coding. We show that Bayes-ratio coding has attractive computational properties: it can be propagated between populations without synaptic weight adjustments, and it can be achieved by divisive normalisation with adaptive weights (Carandini and Heeger, 2012; Westrick et al., 2016).

We start the next section by introducing our efficient coding framework for addressing the normative question posed above. We show numerically that our framework predicts that the firing rate of clusters of similarly tuned neurons should remain constant despite shifts in environmental stimulus statistics, with individual neurons free to shuffle their average rates. This result is shown to hold across a diverse set of noise models, demonstrating the robustness of the optimality of homeostasis to the precise nature of neural response noise. We additionally demonstrate that our framework can account for a wide distribution of stimulus-averaged single-neuron firing rates as observed in cortex. We then provide an analytic argument that demonstrates why, within a certain parameter regime, homeostasis arises from our optimisation problem, and show that cortical areas, in particular the primary visual cortex (V1), are likely to be within this parameter regime. We then numerically validate the quality of our analytic approximations to the optimal solution. Lastly, we apply our theory to DDC representational codes, showing how homeostatic DDC can account for stimulus specific adaptation effects observed experimentally.

Results

Theoretical framework

The central question we seek to address is how the presence of response noise affects optimal neural coding strategies. Specifically, we consider how a neural population should distribute activity in order to best mitigate the effect of neural noise on coding fidelity, subject to metabolic constraints, irrespective of the computations it performs on sensory inputs (which we do not optimise, but take as given). We consider a population of K neuronal units, responding to the (possibly high-dimensional) stimulus s. We assume that our population engages in rate coding using time bins of a fixed duration, and denote the vector of joint population spike counts in a coding interval by n = (n1, …, nK). Single-trial neural responses, n, are taken to be noisy emissions from neural tuning curves, h(s), according to a noise distributions n |s ∼Pnoise(n|s), subject to 𝔼[n|s] = h(s). We consider a number of different noise models, including correlated and power-law noise, showing that our main results are robust to specific assumptions about the nature of noise in neural systems. The sensory environment is characterised by a stimulus distribution P (s); accordingly, an “environmental shift” corresponds to a change in P (s) (making certain stimuli more or less prevalent).

We are interested in how changes in the stimulus distribution, P, affect the responsiveness of population neurons. We therefore adopt a shape-amplitude decomposition of the neural tuning curves. The tuning curve of the a-th unit, ha(s), is factorised into a representational curve, Ωa(s), and a gain, ga: Fig. 2A demonstrates the effect of changing the gain while keeping the representational curve constant. We take the Ωa to encode the computations represented by the neural population, and thus to be determined by those computational goals; we therefore treat them as given, and do not optimise them within our framework. Importantly, we do not make any assumptions on the shape of the representational curve: Ωa can be any complex (e.g., multi-modal, discontinuous) function of the possibly high-dimensional stimulus, and can thus represent any computation. In this respect, our treatment is more general than other efficient coding frameworks (e.g., Ganguli and Simoncelli (2014)), which place tight constraints on the shape and configuration of the tuning curves (see the Discussion for a more detailed comparison of our approach with those studies). In particular, this generality enables our theory to apply to populations located deep in the processing pathway, and not just to primary sensory neurons.

The effect of changing the gain on tuning curves and different components of the objective function.

(A) As the gain, g, of a unit is increased, the shape of its tuning curve remains the same, but all firing rates are scaled up. (The cartoon shows a one-dimensional example, but our theory applies to general tuning curve shapes and joint configurations of population tuning curves, on high-dimensional stimulus spaces.) (B) Cartoon representation of the efficient coding objective function. Optimal neural gains maximise an objective function, ℒ, which is the weighted difference between mutual information, I, capturing coding fidelity, and metabolic cost, E, given by average population firing rate.

Our framework, relating environmental stimuli to neural tuning curves and finally to noise responses, is summarised by the diagram in Fig. 3.

Diagrammatic representation of the generative model underlying the theoretical framework.

The environment gives rise to a stimulus distribution P from which the stimulus s is drawn. Conceptually, the brain performs some computation on s, the result of which are represented by neurons via their representational curves Ωa(s). These are multiplied by adaptive gains, ga, to yields the actual tuning curves ha(s). The single-trial neural responses are noisy emissions based on ha(s).

We theorise that units adapt their gains to maximise an objective function that trades off the metabolic cost of neural activity with the information conveyed by the responses (Levy and Baxter, 1996; Ganguli and Simoncelli, 2014). Informally As is common in efficient coding frameworks (see e.g., Atick and Redlich (1990); Ganguli and Simoncelli (2014)) we assume the main contributor to metabolic cost is the energy cost of emitting action potentials. Thus we take the metabolic cost term in the objective to be proportional to , the total average spike count fired by the population. On the other hand, a natural and common choice (up to approximations) for the coding fidelity is the mutual information between the stimulus and response, I(n; s). However, since in general the analytic maximisation of mutual information is intractable, in the tradition of efficient coding theory (Brunel and Nadal, 1998; Ganguli and Simoncelli, 2014; Linsker, 1988), we will instead optimise an approximate surrogate for mutual information. We decompose the mutual information as I(s; n) = H[n] −H[n|s]. The marginal entropy term H[n] can be upper bounded by the entropy of a Gaussian with the same covariance, Substituting the right hand side for the marginal entropy term in the mutual information, gives us an upper bound for the latter. Using this upper bound on mutual information as our quantification of “coding fidelity”, yields the objective function where the constant µ > 0 controls the information-energy trade-off. This trade-off is illustrated in Fig. 2B. To summarise, we assume that the optimal gains maximize the objective Eq. (3) Note that because the Ωa depend deterministically on the stimulus s, and in turn fully determine the statistics of na, we have the identity I(n; s) = I(n; Ω), which also holds for the upper bound on the mutual information that we are using. This means that the coding fidelity term in the objective function can equivalently be interpreted, without any reference to the low-level stimulus, as (an upper bound on) the mutual information between the population’s noisy output and its ideal, noise-free outputs, Ω, as specified by the computational goals of the circuit.

Optimal adaptation of gains leads to rate homeostasis across a variety of noise models

We now investigate the consequences of optimal gain adaptation, according to Eqs. (3)–(4), on the statistics of population response. As our hypothesis is that gains adapt in order to optimally mitigate the effect of neural noise on coding fidelity (with minimal energy expenditure), ideally we would like our conclusions to be robust to the model of noise. We thus consider a broad family of noise models which allow for general noise correlation structure, as well as different (sub- or super-Poissonian) scalings of noise strength with mean response. Specifically, we assume that conditioned on the stimulus, the noisy population response, n, has a normal distribution, with mean h(s), and covariance matrix where h(s) is the diagonal matrix with the components of h(s) on its diagonal (more generally, we adopt the convention that the non-bold and index-free version of a bold symbol, representing a vector, denotes a diagonal matrix made from that vector), σ scales the overall strength of noise, and 0 < α < 1 is a parameter that controls the power-law scaling of noise power with mean response h(s). Finally, Σ(s) is the noise correlation matrix which can depend on the stimulus, but is independent of the gains. The noise model Eq. (5) can equivalently be characterised by (1) the noise variance of spike count na is given by σ2ha(s)2α, and (2) the noise correlation coefficient between na and nb is given by Σab(s), independently of gains. Thus the strength (standard deviation) of noise scales sublinearly as gain to the power α. Note that the case σ = 2α = 1 corresponds to Poisson-like noise with variance of na equal to its mean; in App. B.2, we show that, under mild conditions, a true Poisson noise model (with no noise correlations) leads to the same results as we derive below for the Gaussian noise model, Eq. (5), with σ = 2α = 1 and Σ(s) = I.

In general, the objective ℒ, Eq. (3), depends tacitly on the stimulus distribution P (s) and the tuning curves Ω(s). As shown in App. B.1, for our class of noise models, Eq. (5), ℒ depends on P and Ω only through the following collection of parameters:

The units’ pre-modulated average response (i.e., the average spike count before multiplication by the adaptive gain ga) These are related to the actual average responses, ra, via ra = 𝔼 [ha(s)] = gaωa.

The coefficients of variation of Ωa(s), which we denote by CVa.

The stimulus-averaged signal correlations, i.e., the Pearson correlation coefficients between Ωa(s)’s. We denote the matrix of these signal correlations by ρ.

The matrix W with elements defined by W can be viewed as a stimulus-averaged normalised noise covariance matrix (note that for α = 1/2 this matrix is a correlation matrix, in the sense that it is positive-definite and diagonal elements equal to 1.)

Note that these quantities depend on the stimulus distribution P and the functions Ωa. Note also that, except for ωa, the other three sets of quantities are invariant with respect to rescalings of the representational curves, and therefore can be equivalently defined in terms of the actual tuning curves, ha, rather than Ωa. Thus, CVa’s are, equivalently, the coefficients of variation of the neurons’ trial-averaged responses, and ρab’s) are the correlation coefficients between pairs of trial-averaged responses (hence “signal correlation”).

As shown in App. B.1, in terms of the above quantities, the objective function, Eq. (3), is given by where, in line with the convention introduced above, CV, ω, and g denote K × K diagonal matrices with CVa, ωa, and ga on their diagonals, respectively. We have obtained the expression Eq. (8) for the objective for the case of a Gaussian noise model, Eq. (5). This noise model is very general, in the sense that it can capture any power-law relationship between the mean and variance of the response (given a stimulus) through α, and any noise strength and correlation structure through σ2 and Σ(s). However, spike counts are discrete rather than continuous, a commonly used noise model for which is the Poisson distribution. In App. B.2, we demonstrate that, under appropriate conditions, for an independent Poisson noise model the objective Eq. (3) is approximately given by Eq. (8) with α = 1/2 and with σ2W replaced by the identity matrix.

We now present the results of numerical optimisation of the objective Eq. (8) for different noise distributions within the general family of noise models, Eq. (5). To capture the notion of adaptation under environmental shift, we simulated a family of environment models smoothly parameterised by ν ∈ [0, 1]. An environment corresponds to a stimulus density P (s), which, as we have just seen, affects the objective only via ω, ρ, W, and CV. We will therefore specify environments through these four quantities. For simplicity, we fixed , σ2, and µ at 10, 1, and 5/(1 ∈ − α), in all environments. (Below, in Sec. 2.5, we will discuss the biological basis for these parameter choices.) On the other hand, ωa and ρ depended smoothly on the environment parameter ν, as follows. For each a = 1, …, K, ωa(ν = 0) and ωa(ν = 1) are drawn independently from a wide distribution (specifically the beta distribution Beta(6, 1)). The value of ωa(ν) at intermediate values of ν were then obtained by linear interpolation between the independently sampled boundary values. This method for sampling environments is illustrated in Fig. 4. Similarly, to construct the matrix ρ(ν), we first constructed ρ(ν = 0) and ρ(ν = 1) with independent random structures and smoothly interpolated between them (see Sec. 4.1 for the details). We did this in such a way that, for all ν, the correlation matrices, ρ(ν), were derived from covariance matrices with a 1/n power-law eigenspectrum (i.e., the ranked eigenvalues of the covariance matrix fall off inversely with their rank), in line with the findings of Stringer et al. (2019) in the primary visual cortex. Finally, the stimulus-averaged noise correlation matrix, W, is determined by our choice of the noise model subfamily used in each simulation. We now present simulation results for three 1-parameter subfamilies of the general family of noise models, Eq. (5), in turn.

The structure of pre-modulated average population responses in the simulated environments.

We simulated a one-parameter family of environments parameterised by ν ∈ [0, 1]. In the environments on the two extremes, for each unit, the pre-modulated responses ωa(ν = 0) and ωa(ν = 1) are randomly and independently sampled from a Beta(1, 6) distribution, with density shown on the left and right panels. For intermediate environments (i.e., for 0 < ν < 1), ωa(ν) is then obtained by linearly interpolating these values (middle panel). The middle panel shows 10000 such interpolations, with every 500th interpolation (ordered by initial value) coloured black.

Uncorrelated power-law noise

Consider first the case of no average noise correlations, in the sense that W = I (in terms of the objective function, Eq. (8), this is mathematically equivalent to having zero stimulus-conditioned noise correlations, i.e., Σ(s) = 1, but with CV’s redefined not to denote the standard coefficient of variation, but rather where ; for Poisson-like scaling, α = 1/2, this reduces to the standard coefficient of variation). We investigated the adaptation behavior of optimally gain-modulated population responses as a function of the noise scaling exponent α. Fig. 5A shows the scaling of spike count variance as a function of mean spike count for different values of α. By construction, in the absence of adaptation, an environmental shift (ν = 0 → ν = 1) results in a wide range of average firing rates (Fig. 5C, middle panel). But following adaptation, the firing rates become tightly concentrated around a single value (Fig. 5C, bottom panel), which matches the value they were concentrated around in the original environment (Fig. 5C, top panel). Thus, adaptation results in (1) uniformisation of average firing rates across the population in a fixed environment, and (2) firing rate homeostasis, where following adaptation the mean firing rates of the units returns to their value before the environmental shift. These same uniformisation and homeostatic effects are seen for other values of α (Fig. 5B and Supp. Fig. 14). In Fig. 5B we quantify deviation from perfect homeostasis as the average % difference in the units’ mean firing rates between the initial environment (ν = 0) and environments at increasing ν. Even under super-poissonian noise (α = 0.75 in Fig. 5B), the average relative change in firing rates between environments never gets above 3%. Fig. 5B additionally shows that the slower spike count variance grows with mean spike count (i.e., the smaller α is) the more precise the firing rate homeostasis.

Optimal gains under uncorrelated power-law noise lead to homeostasis of rates.

(A) Illustration of power-law noise. The trial-to-trial variance of spike counts, as a function of the trial-averaged spike count, is shown for different values of the noise scaling parameter α. (C) Distribution of average firing rates before and after a discrete environmental shift from ν = 0 to ν = 1 for the case α = 1/2. Top: the histogram of firing rates adapted to the original environment (ν = 0) before the shift, as determined by the optimal gains Eq. (4) in this environment, . Middle: the histogram of firing rates immediately following the environmental shift to ν = 1, but before gain adaptation; these are proportional to . Bottom: the histogram of firing rates after adaptation to the new environment, proportional to . We have normalised firing rates such that the pre-adaptation firing rate distribution has mean 1. (B) Deviation from firing rate homeostasis for different values of α as a function of environmental shift. For each environment (parametrised by ν), we compute the optimal gains and find the adapted firing rate under these gains. The average relative deviation of the (post-adaptation) firing rates from their value before the environmental shift, , is plotted as a function of ν.

Uniform noise correlations

In the second sub-family of noise models we allow for noise correlations, but consider only Poisson-like scaling of noise power (i.e., α = 1/2). More precisely, we assume that the effective noise correlation coefficients between all pairs of units (corresponding to off-diagonal elements of the matrix W, Eq. (7)) are the same, equal to p (Fig. 6A) (the positivity of the noise correlation matrix, W, requires that ; assuming large K, the right side of the inequality goes to zero; hence we only simulated cases with positive p). As shown in Fig. 6C, optimal gain adaptation leads to uniformisation and homeostasis of mean firing rates in this noise model as well. In fact, in the presence of uniform noise correlations, homeostasis is even stronger compared to the previous noise model with zero noise correlations; in this case, mean relative deviations in average firing rates never exceed 1% (Fig. 6B). Further, as noise correlations get stronger (i.e., as p increases) we see tighter homeostasis.

Firing rate homeostasis under Poissonian noise with uniform noise correlations.

(A) Illustration of an effective noise correlation matrix with uniform correlation coefficients (off-diagonal elements). Panels B and C have the same format as in Fig. 5.

Aligned correlated noise

In the final noise model we considered, we once again fixed α = 1/2 (Poisson-like scaling), but allowed for heterogeneous effective noise correlation coefficients. Specifically, we took the effective noise correlation matrix W to have approximately the same eigenbasis as ρ, the signal correlation matrix —hence “aligned noise”—, but with a different eigenvalue spectrum. As described above, for all noise sub-families, ρ(ν) was obtained by normalising a covariance matrix with 1/n spectrum (corresponding to the findings of Stringer et al. (2019) in V1); in the current case, we obtained W (ν) by normalising a covariance matrix with the same eigenbasis but with a 1/nγ spectrum (i.e., with the n-th largest eigenvalue of W scaling as 1/nγ); see Fig. 7A, and Sec. 4.1 for further details. Fig. 7C demonstrates that, as with the other noise sub-families, optimal gain adaptation in the presence of aligned correlated noise also leads to homeostasis and uniformisation of mean firing rates (see Supp. Fig. 16 for rate histograms for other values of γ). Across different values of γ, homeostasis is again tighter compared to the zero noise correlation case, with the average firing rate shift never exceeding 1% (Fig. 5B). Further, with the exception of γ = 1, we see more homeostasis for higher values of γ, corresponding to lower dimensional noise. The special case of γ = 1 corresponds to so-called information-limiting noise (Moreno-Bote et al., 2014; Kanitscheider et al., 2015), where noise and signal correlation structures are fully aligned: W = ρ. In this case, homeostasis is perfect, and the mean relative error in firing rates is 0. We provide an analytic proof that in this case we have perfect homeostasis and uniformisation in App. B.5, and discuss this situation further in Sec. 2.5 and Sec. 2.6.

Firing rate homeostasis under Poissonian, aligned noise.

(A) Eigenspectra of noise covariance matrix for the aligned noise model for various values of γ; in this model, the n-th eigenvalue is proportional to 1/nγ. Panels B and C have the same format as in Fig. 5.

We have shown that, according to our efficient coding framework, optimal adaptation of response gains (with the goal of combating neural noise at minimal metabolic cost) can robustly account for the firing rate homeostasis of individual units within a population, despite environmental shifts, under a diverse family of noise models. To shed light on these results, in Sec. 2.4, we analytically investigate sufficient conditions on the parameters, Eq. (6)–(7), of our model under which we expect homeostasis to emerge. Following that, in Sec. 2.5, we will show that biological estimates for these parameters obtained in cortex (which are determined by natural stimulus statistics, as well as the stimulus dependence of high-dimensional population responses) are consistent with these conditions.

In addition to homeostasis, our framework also predicts uniformisation of average firing rates across the population, in any given environment. By contrast, cortical neurons display a wide range of average firing rates spanning a few orders of magnitude (Buzsáki and Mizuseki, 2014). However, we can interpret the units in our preceding numerical experiments not as single neurons, but as clusters of similarly tuned neurons. In this interpretation, uniformisation occurs at the cluster level, with potentially a wide distribution of single-cell average firing rates within each cluster. In the next section, we show that our model is indeed consistent with such an interpretation. In particular, in the case of uncorrelated noise with Poisson-like mean-variance scaling, we show that a simple extension of our model gives rise to both diversity of firing rates as seen in cortex as well as homeostasis and uniformisation at the level of clusters of similarly tuned neurons.

A clustered population accounts for the diversity of firing rates seen in cortex

A cortical neuron can have very similar response tuning to other neurons in the same area (particularly with nearby cells). We will call a collection of neurons which have similar (but not necessarily identical) representational curves a cluster. To gain intuition and allow for analytic solutions, we will henceforth adopt a toy model in which the N neurons in the population are sorted into K such clusters. In this toy model, neurons within a cluster have very similar stimulus tuning, but neurons in different clusters are tuned differently; the model thus exaggerates a more realistic situation in which there is a more gradual transition from similar to dissimilar tuning curves within the entire population. More concretely, we will assume that the representational curves of the neurons, indexed by i, belonging to cluster a, are small perturbations to the same representational curve characterising that cluster; i.e., where ϵ is a small parameter controlling the deviation from perfect within-cluster similarity. Below, hatted symbols denote variables defined analogously to Eq. (6)–(7) but for single neurons, with Ωa(s) replaced by .

We assume single-neuron gains adapt to optimise an objective ℒpop, defined analogously to the objective ℒ, Eq. (8), but for individual neurons rather than clusters. In the case of uncorrelated noise with Poisson-like scaling, this objective is given by Where are the mean firing rates of individual neurons, given by ; note that maximising ℒpop in terms of the single-neuron gains is equivalent to maximising it in terms of . In App. B.3, we show that this objective can be expanded in ϵ as Here, the leading order term, ℒ (r = ωg), is a cluster-level objective and is gi(ven) by Eq. (8), now viewed as a function of the cluster rates ra, i.e., the totalLmean firing rate of neurons in a cluster; thus where Ca is the set of neurons in cluster a. Note that this term is a function only of the cluster firing rates. Thus, for ϵ = 0 (perfect similarity within clusters), the efficient coding objective function is blind to the precise distribution of single-cell firing rates, , among the neurons of a cluster, as long as the total rate of the cluster is held fixed. This is because for clusters of identically-tuned and independently-firing neurons with Poisson-like noise, the cluster can be considered as a single coherent unit whose firing rate is given by the sum of the firing rates of its neurons. Thus at zeroth order, the distribution of individual neuron rates are free and undetermined by the optimisation, as long as they give rise to optimal cluster rates. At small but nonzero ϵ, is still dominated by ℒ; we thus expect this term to play the dominant role in determining the cluster rates, with terms of order ϵ2 and higher having negligible effect. However, the perturbation term breaks the invariance of the loss with respect to changes in the distribution of single neuron gains (or rates) within clusters.

We will therefore approximate the maximisation of the total objective function ℒpop as follows. We first specify the cluster rates by maximising ℒ (r), obtaining the optimal cluster rates . We then specify the rates of individual neurons within the population by maximising the leading correction term in the objective, , subject to the constraint that the single-neuron rates in each cluster sum to that cluster’s optimal rate. We reinterpret the optimisation in the previous sections to correspond to the first stage of this two-stage optimisation, with population units there corresponding to clusters of similarly tuned neurons. As we have seen in Sec. 2.2, optimisation of the cluster-level objective, ℒ, can result in homeostasis and uniformisation across clusters.

As for the distribution of individual neuron rates, we show in App. B.3.4 (see the derivation of Eq. (91)) that, in the parameter regime (see Secs. 2.4 and 2.5) in which cluster-level homeostasis and uniformisation occurs, the perturbation objective ℒpert decomposes into a sum of objectives over different clusters. Thus, in the second stage of optimisation the problem decouples across clusters, and (as shown in App. B.3.4) the optimal single-neuron rates in cluster a maximise the objective subject to the constraints Here ΔΩi is defined to be the centered, zero-mean version of δΩi (we can interpret ΔΩi as the direction of the perturbed tuning curve in excess of the cluster tuning curve Ωa): Note that the optimisation problem, Eqs. (12)–(13), is a quadratic program.

We solved the optimisation problem Eqs. (12)–(13) numerically for 10,000 randomly generated covariance matrices (note that different instances of the optimisation can be thought of as characterising different clusters within the same population). In general, the covariance matrix, (which defines this optimization problem), depends on the stimulus distribution, the cluster tuning curve, and properties (e.g., degree of smoothness) of the intra-cluster variations in single-cell tuning curves. In our simulations, instead of making specific arbitrary choices for these different factors, we used a minimal toy model where the covariance matrix was generated as the matrix of inner products of k random vectors in a D-dimensional space (see Sec. 4.2 for details). In App. B.3.5 we show that D can be interpreted as the dimensionality of the space of independent perturbations to the cluster-wide tuning curve Ωa which vary significantly on the portion of stimulus space to which the cluster tuning curve responds (below, we will refer to D as the effective tuning curve dimension).

In Fig. 8 (see also Supp. Fig. 18) we show the resulting firing rate distributions for neurons aggregated across all clusters (corresponding to the different optimisations), for different values of the cluster size k and effective tuning curve dimension, D, of within-cluster tuning curve variations. Firstly, note that, depending on these parameters, a significant fraction of neurons can be silent, i.e., with zero firing rate in the optimal solution; the figure shows the rate distributions for the active neurons, as well as their percentage of all neurons. Our theory therefore predicts that a significant fraction of cortical neurons are silent in any given sensory environment, and that which neurons are silent can shuffle following shifts in environmental stimulus statistics (specifically shifts that result in significant changes in the signal correlation structure of the cluster neurons). Secondly, note that the distributions span multiple orders of magnitude, in agreement with empirical observations in the cortex (see e.g. Slomowitz et al. (2015); Hengen et al. (2013); Maffei and Turrigiano (2008)), and is approximately log-normal (albeit with a skewed tail towards small log-rates), also consistent with empirical findings (Buzsáki and Mizuseki, 2014). Lastly, for fixed k, as the effective tuning curve dimension, D, increases, the fraction of silent neurons decreases (Supp. Fig. 18).

The distribution of single neuron firing rates is consistent with empirical observations of approximately log-normal rates.

Each panel shows the distribution of log firing rates of the active single neurons for different choices of cluster size k and effective tuning curve dimension, D, of the space of within-cluster tuning curve variations (see the main text for the definition, and Sec. 4.2 for further details). The distributions shown are for the active neurons with nonzero firing rate, with their percentage of the total population given in each panel (purple text). For all simulations we fixed the cluster firing rate to µ = k × 5 Hz, such that the mean rate of single neurons is 5 Hz on average. The plotted densities were obtained by averaging over 10, 000 optimisations of Eq. (12) based on different random draws of Cov(ΔΩ).

To summarise, we have shown that the objective ℒ considered in Sec. 2.2, reinterpreted to correspond to the cluster level, can arise from optimisation of a similar objective, ℒpop, defined at the level of single neurons (Eqs. (10)–(11)). Furthermore, the correction term to the cluster-level objective, ℒpert, serves to break the symmetry within clusters, generating a diverse range of firing rates which matches the distribution of cortical single-neuron firing rates. In the rest of the paper, we will pursue the problem at the coarser level of clusters.

Analytical insight on the optimality of homeostasis

The numerical results of Sec. 2.2 showed that, across a range of neural noise models, optimal gain modulation for combating noise can give rise to firing rate homeostasis and uniformisation across neural clusters defined based on similarity of stimulus tuning. To shed light on these results, we analytically investigated how and in what parameter regimes these results arise.

We found that homeostasis and uniformisation emerge in the regime where the matrix is small, according to some notion of matrix norm; here β = 2 − 2α. The intuitive meaning of this matrix is that its inverse, Δ−1, characterises the structure of signal-to-noise ratio (SNR) in the space of cluster responses (before gain modulation). Therefore the above condition can be roughly interpreted as the requirement that signal-to-noise ratio is strong. We will discuss this condition, and the constraints it imposes on various model parameters, further in Sec. 2.5, and will provide evidence that it is likely to hold in cortex. Here, we focus on its consequences.

We developed a perturbative expansion in Δ, to obtain approximate analytical solutions for the cluster gains that maximise our objective ℒ, Eq. (8) (see App. B.4). In the first approximation, i.e., to zeroth order in Δ, this expansion yields We denote this approximate solution by . To see the significance of this result, note that the average response of cluster a, i.e., its stimulus-averaged total spike count is given by 𝔼 [ha(s)] = gaωa. Thus, in the leading approximation, the stimulus-average spike count of cluster a is given by the constant . Thus, since βµ is constant across both clusters and environments, the zeroth order solution yields homeostasis and uniformisation of cluster firing rates.

We also calculated the first order correction to this homeostatic solution. This is given by which (assuming Δaa are small) yields approximate homeostasis, with small deviations depending on the SNR structure, as captured by Δaa, and how that structure shifts between environments.

Improvements to the zeroth-order homeostatic solution can also be obtained in a different manner: by finding the best configuration of cluster gains, , that, by construction, are guaranteed to yield homeostasis and uniformisation. As we saw above, the latter arise if and only if gaωa is constant (across clusters and environments). We denote this a priori unknown constant by χ. In other words, we let under which the average spike count of clusters in the coding interval is given by χ. The variable χ is then chosen to optimise the expectation of ℒ, Eq. (8), across a relevant family of environments (note that we need to optimise the average objective across environments, since otherwise the optimal χ will in general depend on the environment, invalidating homeostasis). With such an optimal χ, ghom will, by definition, perform better than g(0) in terms of our objective function. Further, it can be shown (see App. B.7) that χ < βµ, i.e., the mean cluster firing rate under the optimal homeostatic solution is strictly smaller than that predicted by g(0), and that the average spike-count of the entire population of clusters is approximately the same under this solution as it is under g(1). In Sec. 2.6, we numerically compare the performance of g(0), g(1), and ghom, and find that in many cases, the homeostatic ghom performs close to the (numerically computed) true optimal gains.

Finally, we note the following exact analytical result regarding homeostasis that holds even for non-small Δ. It follows form the structure of the objective function, Eq. (8), that, irrespective of how pre-modulation mean rates, ωa, change between two stimulus environments, as long as Δ and W, the effective noise covariance, remain invariant across the environments, optimal gains imply exact homeostasis of each cluster’s average firing rate across the two environments. However, unless Δ is small, in general the optimal solution does not result in uniformization of firing rates across clusters.

Conditions for the validity of the homeostatic solution hold in cortex

In this section, we further examine the conditions under which the homeostatic solution, Eq. (16), provides a good approximation to the optimal gains, and, based on empirical estimates of relevant quantities, discuss whether these conditions obtain in cortical populations. As noted above, the general mathematical condition is that the matrix Δ, defined in Eq. (15), is in some sense small (recall that through its inverse, Δ represents the structure of signal-to-noise ratio —before gain modulation— in the coding space). We can make this statement more concrete as follows. The homeostatic solution, and the first order correction to it, Eq. (17), arise from a perturbative asymptotic expansion in the small Δ. When the leading order correction to the (zeroth-order) homeostatic solution becomes non-negligible, it indicates the breakdown of this expansion. From Eq. (17) we can thus reformulate the condition for the validity of the homeostatic approximation as for all a. Before proceeding to empirical estimates, let us first delineate the dependence of Δ (or its diagonal elements) on different model parameters. Firstly, note that Δ scales linearly with σ2; thus, if noise is scaled up, so is Δ. Next, consider increasing the noise scaling exponent α. This decreases (βµ)β provided α ≤1 − e/2µ (which holds in our simulations above). Moreover, increasing α increases the diagonal elements of W (see Eq. (7)). Intuitively, this increases the magnitude of W. Through these two mechanisms, increasing α will generally increase the size of Δ. Next, note that Δ has an inverse relationship with the coefficients of variation, CV, which increase as neurons become more stimulus selective; accordingly, the more selective the tuning curves, the smaller the Δ. Lastly, Δ has a complex dependency on the spectra of both CVρCV and W as well as the alignment of their eigenbases (which represent the geometric structure of signal and noise in the neural coding space, respectively). In Sec. 2.6, we will numerically investigate these dependencies, and their effects on the validity of the homeostatic approximation.

Translated to biological quantities, it follows from the above dependencies that the smallness of Δ (or Δaa) requires that (1) the cluster firing rates are sufficiently high, (2) noise variance scales sufficiently slowly with mean response, (3) individual clusters are sufficiently selective in responding to stimuli, and (4) the neural representation has a high-dimensional geometry in the population response space. We now review the available biological data on these variables in cortical populations.

High firing rate

We first estimate βµ which enters the prefactor in Eq. (19). We see from the zeroth order solution Eq. (16) that βµ is approximately the average total spike count of a cluster over the rate-coding time interval. Condition Eq. (19) thus requires that the stimulus-averaged firing rate of all clusters are sufficiently high. A wide range of mean firing rates for individual neurons have been reported in cortex. Here we focus on firing rates in rodent V1 during free behaviour (which tends to be lower compared to rates in cat or monkey cortex). Reported values tend to range from 0.4 Hz (Greenberg et al., 2008) to 14 Hz (Parker et al., 2022), depending on the cortical layer or area or the brain state, with other values lying in a tighter range of 4-7 Hz (Hengen et al., 2013; Szuts et al., 2011; Torrado Pacheco et al., 2019). Therefore, for rodent V1, we take a mean firing rate of 5 Hz to be a reasonable rough estimate. Assuming a coding interval of 0.1 seconds, and a cluster size of k = 20 (as a guess for the number of similarly coding V1 neurons), we obtain βµ ∼ 10. (Note that for lower or higher single neuron firing rates, a similar value of µ could be achieved by correspondingly scaling cluster sizes up or down.)

Sufficiently slow scaling of noise variance

We can obtain an estimate for the prefactor σ2/(βµ)β, by fitting the mean-variance relationship for cluster spike counts: variance = σ2mean2−β. A number of papers have fit such a relationship (Shadlen and Newsome, 1998; Gershon et al., 1998; Moshitch and Nelken, 2014; Koyama, 2015). We adjusted the values reported by those papers for our chosen coding interval length and cluster size (see Sec. 4.3, the plot in Fig. 13, and the values of β and σ2 thereby obtained). These (together with the estimate βµ ≈ 10) yield values for σ2/(βµ)β ranging from 0.05 (Shadlen and Newsome, 1998) to 0.31 (Koyama, 2015), with median value 0.14. Note that this is close to the value for Poissonian noise (σ2 = 1, β = 1), which is 0.1.

Sufficiently selective responses

The coefficient of variation CV can be seen as a measure of neural selectivity or sparseness of neural responses. To see this, consider a toy model in which the cluster responds at a fixed level to a fraction p of stimuli and is silent otherwise. In this case, CV2 = (1 −p)/p ≈ 1/p for small p. Our condition therefore requires that neurons are sufficiently selective in their responses and respond only to a small fraction of stimuli. Lennie (2003) places the fraction of simultaneously active neurons (which we use as a proxy for the response probability of a single cluster) at under 5%. For our toy model, this yields the estimate CV2 ≈ 20. The Bernoulli distribution in the toy model is particularly sparse, and so we take CV2 ≈ 10 as a more conservative estimate.

High-dimensional signal geometry

Assuming, for simplicity, that the coefficients of variation are approximately the same for all clusters, we see that the expression for Δaa are proportional to the diagonal elements of ρ−1W. To estimate the latter, we will assume that these elements are all comparable in size (This is expected to be valid when the eigenbasis of ρ−1W is not aligned with the standard (cluster) basis, which, in turn, corresponds to distributed coding), and will therefore estimate their average value, which is given by the normalised trace . By the invariance of trace, this can equally be characterised as the mean eigenvalue of ρ−1W. We will estimate this trace for two sub-cases of interest.

Firstly, consider the case of zero noise correlations, i.e., W = I. In this case, the trace is the mean of [ρ−1]aa. Recall that ρ is the signal correlation matrix of neuron clusters. For a correlation matrix, it can be shown that [ρ−1]aa ≥ 1, with equality if and only if cluster a has zero signal correlation with every other cluster. [ρ−1]aa can thus be seen as a measure of the extent to which cluster a shares its representation with other clusters, i.e., as a measure of representational redundancy. Many traditional efficient coding accounts predict zero signal correlation between neurons, at least in the low noise limit (Barlow, 2012; Atick and Redlich, 1990; Nadal and Parga, 1994, 1999), providing an additional normative justification for low signal correlations. However, complete lack of signal correlations is not necessary for our condition to hold; we merely require a sufficiently slow decay of the eigenvalue spectrum of the signal correlation matrix. This condition is geometrically equivalent to neural responses forming a representation with high intrinsic dimensionality. Stringer et al. (Stringer et al., 2019) found that, the signal covariance matrix of mouse V1 neurons responding to natural stimuli possesses an approximately 1/n spectrum. In this case, for K large, we obtain the estimate [ρ−1]aa ≈ tr(ρ−1)/K ≈ ln(K)/2 (see App. B.6). Given the slow logarithmic increase with K, we expect that our condition can hold even for very large neural populations. For example, suppose we take the entire human V1 as our neural population, to obtain an arguably generous upper bound for K. This contains roughly 1.5 × 108 neurons (Wandell, 1995), leading to 7.5 × 106 clusters of 20 neurons, and therefore an average value of [ρ−1]aa just under 8. This, together with our other estimates above, yields the following overall estimate for Δaa We now consider another sub-case of interest, that of so-called information-limiting noise where directions of strong signal variations align with those of strong noise variation (Moreno-Bote et al., 2014; Kanitscheider et al., 2015). Specifically, we consider the extreme case W = ρ, i.e., when noise and signal correlations are perfectly aligned. This case was considered numerically in Sec. 2.2, Fig. 7, and we saw that in this case the optimal solution displayed perfect homeostasis and uniformisation of rates across clusters. Moreover, in this case, ρ−1W = I, and so the normalised trace tr(ρ−1W) does not scale with K at all. Note that this is an extreme of signal-noise alignment. In Sec. 2.2 we also considered less extreme forms of alignment, in which W shares the eigen-bases with ρ but has a 1/nγ spectrum (as opposed to the 1/n spectrum for ρ). We show in App. B.6 that, for 0 < γ < 1 (relatively high-dimensional and weakly correlated noise), we have . On the other hand, for γ > 1 (relatively low-dimensional and strongly correlated noise), tr(ρ−1W) in fact decays like K− min(γ−1,1) as K grows. Thus, in this case, the estimate of Δaa is (potentially significantly) smaller than that of Eq. (20). Accordingly, even partial signal-noise alignment (especially for correlated, relatively low-dimensional noise) can strongly reduce the size of Δ, encouraging homeostatic coding.

The above analysis makes it clear when we should expect homeostasis in general – when cluster firing rates are high, the noise scaling of cluster responses is sufficiently slow, responses are highly selective, and signal correlation structure corresponds to a high-dimensional geometry (Stringer et al., 2019), possibly with information-limiting noise correlations. On the other hand, when these conditions are violated, the optimal gain configuration can potentially deviate strongly from the homeostatic solution. Moreover, given the above estimates, we thus see that in V1 and possibly other cortical areas, we can expect the leading corrections to the homeostatic solution (see Eq. (17) and Eq. (20)) to not exceed 10%, in relative terms. Indeed, as we will show in the Sec. 2.6 using numerical validations, for values of the above parameters in their estimated biological range, the homeostatic solution provides a very good approximation to the optimal solution of Eq. (3).

Numerical comparison of homeostatic gains to optimal gains

We now return to the numerical simulations performed in Sec. 2.2, to compare the performance and accuracy of the different approximate solutions to optimally adapted gains, discussed in Sec. 2.4, with those of the optimal solution of Eq. (8). For each of the different noise models considered in that section we compare the performance of the homeostatic gains and (where χ is chosen as detailed in App. B.7), as well as the first-order correction solution g(1) (see Eq. (16)–(18)) to the numerically optimised gains gopt. To examine the adaptation properties of these gains, we once again work with the sequence of environments indexed by ν ∈ [0, 1], with the same specifications as in the numerical simulations of Sec. 2.2.

To measure the accuracy of the approximations to the optimal gains, for each environment ν, we calculated the mean relative errors for each of the approximate solutions, denoted by gapp(ν), with gopt(ν), the numerically optimised solution in environment ν. To measure the performance of an approximate solution, we use the following relative improvement measure, defined by which can be interpreted as the improvement in the objective ℒ (·; ν) (the efficient coding objective Eq. (8) according to the statistics of environment ν) achieved by the adaptive gains gapp(ν) over the unadapted gains from the original environment gapp(0), relative to the improvement obtained by the optimally adaptive gains gopt(ν). The better the approximation the closer Capp(ν) will be to 1. Note that, in general, the relative improvement worsens when ν goes to 0 (see Fig. 9B, D and F); this follows from the definition of C(ν) and the fact that when environmental change is weak, using gains optimized in the original environment without adapting them in the new environment is superior to adaptive but only approximately optimal gains. We now summarise the results for the different noise models discussed in Sec. 2.2 (see that section for the detailed specification of the three noise models).

Accuracy of homeostatic approximation to optimal gains.

(A) The relative error, Eq. (21), of different approximate solutions for the optimal gains, averaged across environments (the standard deviations of these values across environments were negligible relative to the average), as a function of the noise scaling parameter α. To give a sense of the scale of variation of the optimal gains across environments, the black line shows the relative change in the optimal gains between the most extreme environments (a measure of effect size),. (B) Relative improvement in the objective for the homeostatic approximation ghom as measured by Eq. (22). Panels C and E (panels F and D) are the same as panel A (panel B), but for the uniformly correlated noise and the aligned noise models, respectively.

Uncorrelated power-law noise

First, we consider the uncorrelated noise model with general power-law scaling of noise strength with mean responses. In this case, the homeostatic solution ghom, and the first order correction, g(1), had very low (< 3%) relative errors for all noise scaling exponents, α, while the relative error in g(0) becomes large for significantly super-Poissonian noise scaling, i.e., α = 0.75 (Fig. 9A). (For the case of uncorrelated power-law noise model, as well as the aligned correlated noise model, we used an analytical approximation to the optimal χ (see App. B.7) used in the homeostatic approximation , Eq. (18). This means that the performance measures in the plots of Fig. 9B and F are lower-bounds for the performance of with the truly optimal χ.) The relative improvement measure Chom(ν) for the homeostatic solution is also close to one, the optimal value, for all α, except for 0.75 (Fig. 9B). For the other approximate solutions too (Supp. Fig. 17A,B), the relative improvement decreases with increasing noise scaling. This is consistent with our analysis, since increasing α increases the size of Δ (see Sec. 2.4) and hence the influence of terms of higher order in Δ.

Uniform noise correlation coefficients

We now examine the effect of noise correlations on the accuracy of the approximate homeostatic solutions, first within the noise family with uniform noise correlation coefficients, denoted by p, and Poisson-like scaling (α = 1/2). For all p, the relative deviation of the homeostatic solution ghom from optimal gains was below 1% (Fig. 9C), while g(0) and g(1) were less accurate overall. Note that the accuracy of the homeostatic solutions ghom and g(0) increases with p, suggesting that positive noise correlations improve the optimality of homeostatic gains. The same trend can be seen in the relative improvement on the objective, Chom, for homeostatic gain adaptation, which increases with p (Fig. 9D).

Aligned noise

Lastly, we considered the algined noise model, in which noise has Poisson-like scaling (i.e., α = 1/2), and is correlated across clusters with the noise and signal covariance matrices sharing the same eigenbasis. In this noise family, the parameter, γ, controls the scaling of the eigenvalues of the noise covariance matrix with their rank, n, as 1/nγ. As in the previous cases, for all γ, the homeostatic solution ghom was very close to the optimal gains gopt (Fig. 9E), and preformed nearly equally (Fig. 9F). The case γ = 1, which corresponds to perfect alignment between signal and noise correlations, W = ρ, is particularly interesting. As noted in the previous section, this case corresponds to that of so-called information-limiting noise correlations (Moreno-Bote et al., 2014; Kanitscheider et al., 2015). As we show in App. B.5, in this case the optimal gains, without any approximation, lead to homeostasis and uniformisation of cluster firing rates. Thus, in this case gopt = ghom exactly, leading to zero relative error (Fig. 9E) for ghom, and perfect relative improvement Chom (Fig. 9F). For other cases, we can see that for increasing γ (at least for the values we simulated), which corresponds to more correlated and lower-dimensional noise (as noise power falls off faster as a function of eigenvalue rank), the homeostatic gains ghom tend to perform better.

Our numerical simulations across the three noise subfamilies (with varying noise power scaling, noise correlations, and noise spectrum) demonstrate that homeostatic strategies can - robustly and across a wide range of noise distributions - combat coding noise by effectively navigating the trade-off between energetic costs and coding fidelity. Finally, note that adaptation in ghom uses only information local to each neuron, i.e., its firing rate. This implies that good performance does not require the use of complex regulation mechanisms which take into account how the encoding of each cluster relates to the population at large; homeostatic regulation, implemented by purely local mechanisms, is sufficient.

Synaptic normalisation allows for propagation of homeostasis between populations

So far, we have considered homeostatic coding from a purely normative perspective, showing with both theoretical arguments and numerical simulations that homeostasis efficiently trades-off between coding fidelity and energetic costs. In this section, we begin to address the question of biological plausibility; specifically, what biological mechanisms are necessary for the homeostatic coding regime to be propagated between neural populations. Specifically, We will show that for the case of a linear feedforward network implementing the representation curves, there is a connection between homeostatic coding and synaptic scaling (Turrigiano, 2008).

Consider two neural populations, with an upstream population providing feedforward input to a downstream population. We will suppose that the upstream population is engaged in homeostatic coding, and ask what is necessary for the downstream population to be also engaged in homeostatic coding. We will denote the tuning and representational curves of the upsteam population by and , respectively, where i = 1 (i = 2) for the upstream (downstream) population. We drop corrections to the zeroth order solution for optimal gains, and work within the homeostatic coding regime; that is we assume the gains of the upstream population are given by .

The downstream population of tuning curves will be given by for m = 1, …, M. Let Wma be the synaptic weight from neuron a in the upstream population to neuron m in the downstream population. Working in a linearized rate model, this gives us that the tuning curve of neuron m is Suppose that, as dictated by the computational goals of the circuit, the downstream population has representational curves for m = 1, …, M. We similarly write each representational curve as a linear combination of the upstream cluster representational curves, We call these weights wma the computational weights, since they are determined by the computational goals of the downstream population. In line with our approach so far, we treat these computational goals, and hence also the computational weights, as given (i.e., set independently of optimal gain adaptation).

We asked how the synaptic weights Wma should depend on the computational weights wma, and how they should adapt as stimulus statistics change in order for the downstream population to also engage in homeostatic coding. In App. B.8 we find that homeostatic coding in the downstream population is guaranteed if the synaptic weights are related to the computational weights via where as before. Note that the total synaptic weight received by a downstream neuron is thus constant. In other words, this scheme requires the normalization of the total synaptic weights received by each downstream neuron to be constant, independently of adjustments of the gains of the pre-synaptic neurons or of possible changes in the computational weights (different normalisation factors are equivalent to different values of χ for different populations; differences in the optimal choice of χ may arise from e.g., different noise correlation statistics or different rate coding intervals between populations). Thus, homeostatic coding, applied sequentially to two populations, provides an additional normative interpretation of synaptic normalisation, in which synapses onto a neuron are jointly scaled to keep total input weights constant. This may be the computational reason for the synaptic scaling observed in certain studies of firing rate homeostasis (Turrigiano et al., 1998; Turrigiano, 2008) that occurs over relatively long timescales.

Homeostatic noisy DDCs

Up to this point we have made no assumptions about the nature of cortical representations, beyond rate coding (and the condition described by Eq. (19)), and thus our framework has been independent of the computational goals of the circuit. We now apply our framework to a specific theory of neural representation and computation, namely the distributive distributional code (DDC) (Vertes and Sahani, 2018), introducing what we term a homeostatic DDC. We will show below, in Sec. 2.10, that specific forms of homeostatic DDC are able to account for stimulus-specific adaptation effects, observed in V1 (Benucci et al., 2013), which have been difficult to account for under other efficient coding frameworks. While there is ample behavioural and psychophysical evidence for the hypothesis that animal perception and decision making approximate optimal Bayesian probabilistic inference (Kersten et al., 2004; Stocker and Simoncelli, 2006; van Bergen et al., 2015; Geurts et al., 2022), our knowledge of the neural implementation of computations serving Bayesian inference (such as computations of posterior distributions and posterior expectations) are rudimentary at best. The DDC framework is one proposal for neural implementations of Bayesian computations underlying perception. In a DDC model, neural responses directly encode posterior expectations of the inferred latent causes of sensory inputs according to an internal generative model. This internal model mathematically relates the latent causes or latent variables, which we denote by z, to the sensory inputs or stimuli, s, via a family of conditional distributions, f (s|z), as well as a prior distribution, π(z), over the latent variables. The sensory system is assumed to implement a so-called recognition model corresponding to this generative model. Namely, given a stimulus s, the task of the sensory system is to invert the generative process by calculating and representing the posterior distribution of latent variables given the current sensory input (see the schema in Fig. 10).

Schema of the generative and recognition models underlying a DDC.

According to the internal generative model, sensory inputs, s, are caused by latent causes, z, that occur in the environment according to a prior distribution π(z). A conditional distribution f (s|z), the so-called likelihood function, describes how the latent causes generate or give rise to the sensory inputs s. The task of the brain is to invert this generative process by inferring the latent causes based on the current sensory input, which is done by computing the posterior distribution Π(z|s).

In a DDC, neural responses are specifically postulated to directly represent the posterior expectations of the latent variables, z, or rather, the posterior expectations of a sufficiently rich (and fixed) class of so-called kernel functions of z. Thus, each neuron, say neuron a, is assigned to a so-called kernel function of the latent variables ϕa(z), and its response represents the posterior expectation 𝔼 [ϕa(z) |s]. Here, we add two assumptions to this standard DDC scheme. First, we assume that the actual single-trial response, na, of the neuron is a noisy version of 𝔼 [ϕa(z) |s], but such that the trial-averaged tuning curve of the neuron is 𝔼 [ϕa(z)|s], up to a constant of proportionality. Second, we assume that these constants of proportionality are adaptive and are given by the optimal gains of our efficient coding framework, which we assume are the homeostatic gains given by . In the language of previous sections, this is equivalent to saying that we assume the representational curve of a neuron with kernel function, ϕa(z), is given by Ωa(s) = 𝔼 [ϕa(z) |s]. Under optimal homeostatic gains, our noisy homeostatic DDC predicts the following relationship between the neural tuning curves and Bayesian posterior and prior expectations of ϕa(z): Thus, this homeostatic DDC framework predicts that the neural tuning curves are the ratio of the posterior to the prior expectations of kernel functions. In deriving this result we have used the tower property to replace the trial-averaged posterior expectation 𝔼 [𝔼 [ϕa(z) |s]] with the prior expectation 𝔼 [ϕa(z)] under the internal model; strictly speaking, this is true under an ideal-observer internal model in which the generative internal model (as specified by f (s|z) and π(z)) provides a bona fide, objective description of the statistics of sensory stimuli s. In Sec. 2.10 below, we will relax this assumption and discuss the consequences of using a non-ideal internal model instead.

Bayes-ratio coding

We now consider a special case of the homeostatic DDC in which the kernel functions are delta functions, ϕa(z) = δ(z− za). Here za is a point in latent variable space corresponding to neuron a’s preferred latent variable value. In this case, the homeostatic DDC becomes where in the last equality we used Bayes’ rule: Π(z|s) = f (s|z)π(z)/P (s). Thus, in this case, the tuning curve of a neuron is the ratio of the posterior distribution to the prior distribution both evaluated at that neuron’s preferred latent variable value. Accordingly, we call this coding scheme Bayes-ratio coding.

Bayes-ratio coding can be achieved via divisive normalisation with adaptive weights. Divisive normalisation has been dubbed a canonical neural operation (Carandini and Heeger, 2012; Westrick et al., 2016), as it appears in multiple brain regions serving different computational goals. Given the feedforward inputs, Fa(s), the divisive normalisation model computes the response (and thus the tuning curve) of neuron a as Here, wb are a collection of so-called normalisation weights, and α ≥ 0 and n ≥ 1 are constants. As we now show, Bayes-ratio coding can be naturally implemented by a divisive normalisation model in which n = 1. In the divisive normalisation implementation of Bayes-ratio coding, the feed-forward inputs, Fb(s), are given by the likelihood function, f (s|zb), of the DDC’s internal generative model, while the normalisation weights wb encode the prior probabilities, wb = π(zb)δzb (the volume element, δzb, is chosen such that the latent variable space is the disjoint union of volumes of size δzb each containing their corresponding sample point zb). We then obtain which after substitution in Eq. (28) yields Taking the limit as α → 0, we obtain Bayes-ratio coding Eq. (27). Therefore, provided α is small compared to P (s), divisive normalisation can be used to approximate Bayes-ratio coding. Not only does this result show that implementing Bayes-ratio coding is biologically plausible, it also provides a normative computational interpretation to both the feedforward inputs and the normalisation weights of the divisive normalization model, as the internal generative model’s likelihoods and prior probabilities, respectively. Note also that the normalisation weights of this model are adaptive, and vary between environments with different latent variable marginal distributions, π(z).

Finally, we consider the propagation of Bayes-ratio coding between populations, by using the scheme discussed in Sec. 2.7 for propagation of homeostatic codes. Suppose we now have a hierarchical generative model z(2) → z(1) → s, specified by a prior distribution π(2)(z(2)) and generative likelihood functions f (2)(z(1) | z(2)) and f(1)(s | z(1)). Then consider a recognition model corresponding to this generative model, comprised of a feedforward network with two populations or layers that implement a Bayes-ratio encoding of the posterior distributions of the hierarchical latent variables z(1), and z(2), respectively, conditional on the stimulus s. Let us assume that the first (upstream) layer is implementing Bayes-ratio coding for the lower-level variables z(1) Then, as we show in App. B.9, a simple linear propagation of rates according to Eq. (23) with synaptic weights given by would implement Bayes-ratio coding for z(2) in the downstream layer. This result is significant for three reasons. Firstly, the synaptic weights make no reference to the prior distribution, π(2), over z(2). There is therefore no need to adapt the weights when the environment changes, if the change only affects the statistics of the higher-level causes z(2) but not the generative process. Secondly, while the downstream population, is part of the recognition model and (effectively) represents the posterior distribution of z(2), the synaptic weights, Eq. (30), needed to implement this representation only require knowledge of the generative probabilities (specifically those given by f (2)). Finally, we can see by induction that this scheme can be propagated through many layers, each representing posterior probabilities of higher-level latent variables further up a generative hierarchy.

Homeostatic DDCs account for stimulus specific adaptation

We found that homeostatic DDCs can provide a normative account of stimulus specific adaptation (SSA). Here, we start by showing this in the special case of Bayes-ratio coding, which is mathematically simpler and provides good intuitions for a more general subset of homeostatic DDCs. We then discuss this more general case and build a homeostatic DDC model of empirically observed SSA in the primary visual cortex (V1) (Benucci et al., 2013).

According to Eq. (27), for Bayes-ratio codes the tuning curve of neuron a is given by ha(s) = χf (s|za)/P (s). Suppose now that the stimulus distribution in the environment, P (s), changes due to a change in the statistics of the latent causes, π(z), while the causal processes linking the latent causes and observed stimuli, as captured by f (s|z), remain stable. In this case, by Eq. (27), the tuning curve of all neurons are modified by a multiplicative factor given by the ratio of the P (s) in the old environment to that in the new environment. This will lead to a multiplicative suppression of the response of all neurons to stimuli that are over-represented in the new environment relative to the old one, in a way that is independent of the neurons’ identity or preferred stimulus. Such a suppression thus constitutes a pure form of stimulus specific adaptation. In particular, neurons do not reduce their responsiveness to stimuli that are not over-represented in the new environment. Thus, tuning curves are suppressed only near over-represented stimuli (and may in fact be enhanced near under-represented stimuli), leading to a repulsion of tuning curve peak from over-represented stimuli. This repulsion is a typical manifestation of stimulus specific adaptation; in V1, for example, orientation tuning curves display a repulsion from an over-represented orientation (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013).

Heretofore we have implicitly made a so-called ideal observer assumption by assuming that the internal generative model underlying the DDC or Bayes-ratio code is a veridical model of the stimulus distribution in the environment. We now provide a significant generalization of the above results to the more general and realistic case of a non-ideal-observer internal model. As we will see, in this case, the effects of adaptation on tuning curves are captured by both a stimulus-specific factor as well as a neuron-specific factor. For a non-ideal observer model, there will be a discrepancy between the environment’s true stimulus distribution, PE(s), and the marginal stimulus distribution predicted by the internal generative model, i.e., In App. B.10, we demonstrate that the Bayes-ratio tuning curves in this case are given by (instead of Eq. (27)) where Note that, for ideal-observer internal models, PE = PI, in which case the integral yields Fa = 1 due to the normalization of f (s| z), and we recover Eq. (27). As we can see, the tuning curves neatly decompose into a “base tuning curve” χf (s|za) multiplied by a stimulus-specific factor PI (s)−1 and a neuron-specific factor . Consequently, if we consider a pair of environments indexed by ν = 0, 1, and assume the likelihood function of the internal model does not change between environments, the transformation of the tuning curves due to adaptation will be given by Eq. (34), Thus, adaptation from environment ν = 0 to environment ν = 1 causes the tuning curves to transform via multiplication by a stimulus specific factor PI (s; ν = 0)/PI (s; ν = 1) and a neuron specific factor Fa(ν = 0)/Fa(ν = 1). Now suppose an environmental change results in an increase in the prevalence of an adaptor stimulus. As long as the internal model is a sufficiently good (if not perfect) model of PE(s), it should adapt its prior distribution, π(z), in a way that results in an increase of its predicted stimulus distribution, PI (s), around the adaptor stimulus. This results in a stimulus specific factor that goes below 1 around the adaptor s, thus causing suppression and repulsion of tuning curves away from the adaptor. Now consider a neuron whose preferred stimulus (roughly represented by za) is close to the adaptor stimulus. We further make the assumption that the change in the internal model, in response to the environmental change, is sufficiently conservative such that it leads to a smaller increase in PI (s) around the adaptor, compared with the increase in the veridical PE(s) (this is a reasonable assumption especially for artificially extreme adaptation protocols used in experiments). In this case, we expect the ratio PE(s)/PI (s) to increase in the support of f (s|za), which (by our assumption about the preferred stimulus za) is near the adaptor. According to Eq. (33) this will lead to an increase in Fa, and hence a suppressive neuron-specific factor Fa(v = 0)/Fa(v = 1) for neurons with preferred stimulus near the adaptor. In accordance with this picture, Benucci et al. (2013) found that homeostatic and stimulus-specific adaptation in V1 can indeed be modelled via separable stimulus-specific and neuron-specific factors.

Note that the assumption of constancy of the likelihood across the two environments need not be framed as a veridical assumption about the objective environmental change, but rather as an assumption about the inductive biases of the internal generative model, according to which it tends to model changes in the observed stimulus distribution, PE(s), as primarily resulting from changes in the statistics of the latent causes in the environment (which the model captures in π(z)), rather than in the causal process itself, as modelled by f (s|z). As long as this is good enough an assumption to result in a PI (s) which changes in similar ways to PE(s) (and in particular exhibits an increase around adaptor stimuli), the conclusions reached above will hold.

In the case of DDC codes that use more general kernel functions, other than the delta functions underlying Bayes-ratio codes, we arrive at an analogous transformation, given by Eq. (138) in App. B.10. However, in this case, neuron-specific and stimulus-specific effects are mixed, and the wider the DDC kernel the stronger the mixing. Nevertheless, for kernels that are unimodal and sufficiently narrow, we expect an approximate factorisation of the effect of adaptation into a stimulus-specific and a neuron-specific suppression factor, as seen in Eq. (34).

We now apply this result to the experiments performed by Benucci et al. (2013). These experiments examined the effects of adaptation on orientation-tuned neurons in the cat primary visual cortex. Anaesthetised cats were shown a sequence of oriented gratings chosen randomly from a distribution that had an increased prevalence (3 to 5 times more likely) of one particular orientation (arbitrarily defined to be to 0°). A control group was exposed to a uniform distribution of gratings. After about 2 seconds (on the order of 50 stimulus presentations), V1 tuning curves had adapted, and exhibited both the suppressive and repulsive effects mentioned above.

To model these findings we constructed a homeostatic DDC model as follows, and fit it to the data from Benucci et al. (2013). We took the model’s stimulus and latent variable spaces to be the circular orientation space [−90, 90), and we took the model’s generative likelihoods to be given by translation invariant functions f (s| z) = f (s− z), proportional to the Gaussian distribution with standard deviation σf, normalised over the circle (more rigorously, we ought to use a wrapped (periodic) normal distribution; however, since the standard deviations are small compared to the length of the circle, the normalisation constant is approximately 1, and we can treat the density as a normal Gaussian). Likewise, we take the DDC kernel functions ϕa(z) = ϕ (z− za) proportional to the Gaussian distribution with standard deviation σϕ.

The distributions of stimulus orientations used in the experiment is highly artificial, in the way they jump discontinuously near the adaptor orientation. The internal prior distribution employed by the brain is more likely to be smooth, reflecting the brain’s inductive bias adapted to natural environments. Thus in our model we chose the internal prior to be a smooth distribution. The orientation distribution used in the experiment can be understood as a mixture of a uniform distribution, and a spike concentrated at the adaptor (Fig. 11A, blue curve). To obtain the smooth internal prior in our model, we replace the spike component of the experimental orientation distribution with a Gaussian distribution centred at the adaptor with standard deviation σπ (Fig. 11A, red curve). Thus, our model has only three free parameters, σf, σϕ and σπ, i.e., the widths of the generative likelihood, kernel functions, and the internal orientation prior, respectively.

Components of the model for stimulus-specific adaptation in V1.

(A) The distribution of stimulus orientations used in the experiments (Benucci et al., 2013), for an adaptor probability of 30% is shown in blue. We assume V1’s internal model use continuous prior distributions and thus we used instead a mixture of the uniform distribution with a smooth Gaussian, with standard deviation σπ, centered at the adaptor orientation (red curve). (B) The baseline tuning curves, i.e., the tuning curves adapted to the uniform orientation distribution. The blue curve was obtained by averaging the experimentally measured tuning curves adapted to the uniform orientation distribution, after centering those curves at 0°. The model’s baseline tuning curves (red) are given by the convolution of the Gaussian likelihood function, with width σf, of the postulated internal generative model, and the Gaussian DDC kernel, with width σϕ, and thus are themselves Gaussian with width . We fit this quantity to the experimentally observed baseline curves as described in Sec. 4.4.

Using the dataset from Benucci et al. (2013), we found the baseline tuning curves, i.e., the tuning curves of the population adapted to the uniform orientation distribution, by assuming that neural response is a function only of the difference between the preferred orientation and stimulus orientation (see Sec. 4.4). We then calculate the changes in the neurons’ preferred orientations (due to adaptation and the resulting tuning curve repulsion) in each of the experimental conditions (corresponding to different levels of over-representation of the adaptor stimulus). We then fit our model’s three parameters to both the unadapted tuning curves (Fig. 11B) and the changes in preferred orientation (see Sec. 4.4 for further details). The best fit values for the model parameters were found to be σπ = 18°, σf = 14.2°, and σϕ = 20°. The predictions of our model are compared to experimental data in Fig. 12. We found that the adaptive tuning curves within our model display suppression and repulsion away from the adaptor stimulus (Fig. 12A,D). In both the experimental data and our model, unadapted tuning curves exhibit a heightened average firing rate near the adaptor stimulus (Fig. 12B,E, blue curves). However, adaptation suppresses responses near the adaptor (Fig. 12A,D, red curves) to just the right degree to return the firing rates to the value before the environmental shift, leading to homeostasis of firing rates (as demonstrated by the uniformity of the red curves in Fig. 12B,E). Lastly, our model recapitulates the repulsion of preferred orientations found experimentally (Fig. 12C,F). Repulsion here is reflected by the change in preferred orientation having the same sign as the difference between the pre-adaptation preferred orientation and the adaptor orientation. As seen, in both the experiment and the model, repulsion is strongest for neurons with pre-adaptation preferred orientation about 30° from the adaptor.

A homeostatic DDC model accounts for the observations of stimulus-specific adaptation in V1 Benucci et al. (2013).

Panels A-C (left column) and E-F (right column) correspond to the experimental results and the predictions of our model, respectively. (A, D) The tuning curves for the adapted (red) and unadapted (blue) populations, averaged across experimental conditions. (B, E) The trial-averaged firing rates of the adapted (red) and unadapted (blue) populations exposed to the non-uniform stimulus ensemble. The trial-averaged population responses to the uniform stimulus ensemble were normalised to 1. (C, F) The repulsion of preferred orientations, obtained from the average tuning curves in panel A, as a function of the deviation of the neurons’ preferred orientation from the adaptor orientation.

In sum, the application of our homeostatic coding framework to a specific DDC model of Bayesian representations (which approximates what we have termed Bayes-ratio coding) explains quantitatively the empirical observations of both stimulus-specific and homeostatic adaptation in V1 (Benucci et al., 2013).

Discussion

We developed a theory of optimal gain modulation for combating noise in neural representations serving arbitrary computations. We demonstrated that — when mean neural firing rates are not too small; neurons are sufficiently selective and sparsely activated; and their responses form a high-dimensional geometry — the trade-off between coding fidelity and metabolic cost is optimised by gains that react to shifts in environmental stimulus statistics to yield firing rate homeostasis. Specifically, our theory predicts the homeostasis of the mean firing rate for clusters of similarly tuned neurons, while, at the same time, it accounts for the optimality of a broad distribution of firing rates within such clusters. By examining empirical estimates of parameters characterising cortical representations, we argued that the conditions necessary for the homeostatic adaptation to be optimal are expected to hold in the visual cortex and potentially other cortical areas. We further validated our approximation by demonstrating numerically that it performs well compared to the calculated optimal gains. Having developed a normative theory of neural homeostasis divorced from the computational aims of the neural population, we next showed how our theory can account for stimulus specific adaptation when coupled with distributed distributional codes (DDC). In particular, we focused on a special case of homeostatic DDC codes which we termed Bayes-ratio coding and showed that this model can account for stimulus specific adaptation effects empirically observed in the primary visual cortex. In the following, we will situate our work within the broader context of efficient coding theory, Bayesian neural representations, and alternative theories of stimulus specific adaptation. We will also discuss the shortcomings of the present work and possible directions for extension.

Within efficient coding theory, our work is closely related to that of Ganguli and Simoncelli (2014), but differs from it in important ways. Ganguli and Simoncelli (2014) considered a population of tuning curves, defined on a one-dimensional stimulus space, which was homogeneous and translationally invariant up to a warping (i.e., reparametrization) of the stimulus space and possible variations in gain. Their normative framework then optimised the neural gains as well as the stimulus warping (i.e., the reparametrization mapping from the original stimulus parameter to one over which the tuning curve population is homogeneous) to maximise a similar objective function to ours. Specifically, their objective employed the Fisher Information lower bound on the mutual information (Brunel and Nadal, 1998), as opposed to the Gaussian upper bound we used; our choice of the Gaussian upper bound was motivated by increased analytic tractability in the multidimensional stimulus case covered by our theory. In the case of unimodal (and one-dimensional) tuning curves, they showed that the optimal warping amounts to a concentration of tuning curves around stimulus values that are over-represented in the environment. On the other hand, in their setting, the optimal gains stay constant and do not adapt. Nevertheless, their solution also exhibits homeostasis of single-neuron firing rates.

Our framework extends and generalises that of Ganguli and Simoncelli (2014) in different ways. In their framework, the tuning curves are one-dimensional, and rigid constraints (most importantly, either unimodal or sigmoidal tuning curves, and homogeneity up to stimulus reparametrization) are placed on the shape and arrangement of the tuning curve population. In contrast, our framework is applicable to tuning curves with heterogeneous (in particular multi-modal) shapes and arbitrary arrangement over a high-dimensional stimulus space; our homeostatic solution only requires sufficiently high signal-to-noise ratio (which is also the condition for the tightness of the mutual information lower-bound employed in Ganguli and Simoncelli (2014)). On the other hand, we only optimised the neural gains, and not the shapes or arrangement of the tuning curves, as we assumed the latter are determined by the computational goals of the circuit and not by the aim of optimally combating coding noise; in this way, our framework is agnostic to the computational goals of the population. Additionally, Ganguli and Simoncelli (2014) only considered the case of uncorrelated Poissonian noise. In contrast, our framework can handle correlated noise with general (non-Poissonian) power-law scaling. Furthermore, we analytically calculated first-order corrections to the homeostatic solution that arises in the high signal-to-noise ratio limit.

As noted above, the theory of Ganguli and Simoncelli (2014) predicts that tuning curves optimised to provide an efficient code under a stimulus distribution with over-represented stimuli cluster around those stimuli. This has been used to explain asymmetries in neural selectivities, such as the over-abundance of V1 neurons with preferred orientations near cardinal orientations, which are more abundant in natural scenes; or more generally, the aggregation of tuning curves around stimuli that are more prevalent in the natural environment. Many of these asymmetries have likely resulted from adaptation at very long (e.g., ontogenetic, or possibly even evolutionary) timescales; however, long-term exposure on the order of few minutes to adaptor stimuli has also been found to result in the attraction of tuning curves towards adaptor stimuli (Ghisovan et al., 2009). This is in contrast to the repulsive effects seen in short-term adaptation (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013). It is likely that the effects of long-term exposure are the results of mechanisms which operate in parallel to those underlying short-term adaptation, with both types co-existing at different timescales. We can therefore interpret Ganguli and Simoncelli (2014) as a model of adaptation to natural stimulus statistics at long timescales. Our results obtained here are applicable to adaptation on shorter timescales (Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013). In particular, when married with Bayesian theories of neural representation, our framework predicts repulsive effects around an adaptor stimulus, as shown in Sec. 2.10.

Westrick et al. (2016) also propose a model of the experimental findings of Benucci et al. (2013). Their model (which is not explicitly normative or based on efficient coding) uses divisive normalisation (Carandini and Heeger, 2012) with adaptive weights to achieve homeostasis and stimulus specific repulsion. As discussed above in Sec. 2.9, Bayes-ratio coding (a special case of homeostatic DDCs, which we showed can account for the findings of Benucci et al. (2013)) can be accomplished by such a divisive normalisation scheme. Our framework therefore yields a normative interpretation of the model of Westrick et al. (2016), and links divisive normalisation with Bayesian representations and efficient coding theory. In particular, our theory provides an interpretation of the normalization weights of the divisive normalization model as the internal prior over latent causes of sensory inputs. This prior should naturally adapt as stimulus statistics change across environments. Additionally, the feedforward inputs to the divisive normalisation model are interpreted as the likelihood of stimuli given latent causes.

Snow et al. (2016) developed two normative models of temporal adaptation effects in V1, based on statistics of dynamic natural visual scenes (natural movies). These models are based on generative models of natural movies in the class of mixtures of Gaussian scale mixture (MGSM) distributions. In MGSMs the outputs of linear oriented filters applied to video frames are assumed to be Gaussian variables, multiplied by positive scale variables. These scale variables can be shared within pools of filter outputs at different times (video frames) and orientations. The two models of Snow et al. (2016) differ according to which pools of filter outputs are able to share scale variables. In both models, V1 neural responses are assumed to represent the inferred Gaussian latent variables of the corresponding MGSM. While each model was able to account for some of the findings of Benucci et al. (2013), neither one could account for all. In particular, each of their two models could only account for either stimulus-specific or neuron-specific adaptation factors found by Benucci et al. (2013). In particular, only the model that accounted for neuron-specific adaptation was able to produce firing rate homeostasis. As we showed in Sec. 2.10, our framework, which combines homeostatic gain modulation with the DDC theory of representation, robustly accounts both for stimulus-specific adaptation and firing rate homeostasis. Furthermore, when there is a discrepancy between the internal model’s stimulus prior and the true environmental stimulus distribution, it additionally accounts for a neuron-specific adaptation factor as well (see Eq. (34)).

Our approach has relied on a number of assumptions and simplifications; relaxing or generalizing these assumptions provide opportunities for future research. Firstly, our toy model (see Sec. 2.3) imposed a cluster structure on the neural tuning curves, whereby neurons within a cluster are similarly tuned, while neurons in different clusters have distinct tuning. We found that total firings rates are equalized across clusters (Sec. 2.2), but the optimal rates of individual neurons within clusters span a broad range (Sec. 2.3). This results from the fact that the efficient coding objective function depends sharply on total cluster firing rates, but changes only slightly when the partition of the cluster firing rate among the cluster’s neurons changes. In reality, tuning similarity has a more graded distribution. As the distinction between clusters with different tuning (or similarity of tuning within clusters) weakens, higher order terms in our expansion of the objective Eq. (11) cannot be neglected, and hence the justification for dividing the problem into two separate, cluster-level vs. within-cluster, problems breaks down. However, arguably, in a more realistic model with a more graded variation in neural signal correlations, the optimization landscape will still contain shallow directions (i.e., directions over which the objective changes slowly), corresponding to partitions of firing among similarly tuned neurons, and non-shallow directions corresponding to total firings of such groups of neurons; if so, we would expect a similar solution to emerge, whereby similarly tuned neurons display heterogeneous firing rates, with the total firing rate of such a group of neurons displaying approximate homeostasis.

Secondly, we showed analytically that, at the level of cluster responses, homeostasis emerges universally in the high signal-to-noise regime. In general, firing rate homeostasis is no longer optimal in the low signal-to-noise ratio limit, as correction terms to our solutions become large. However, as we argued in Sec. 2.5, based on empirical values we expect the signal-to-noise ratio within cortex to be sufficiently high for our results to hold. Generally, within efficient coding theory (except for simple cases such as linear Gaussian models), the regime of arbitrary signal-to-noise ratio is analytically intractable. Many approaches are therefore limited to the regime of high signal-to-noise ratio. For example, the Fisher Information Lower Bound used frequently within efficient coding theory breaks down outside of this regime (Yarrow et al., 2012; Bethge et al., 2002; Wei and Stocker, 2016). Thus, inevitably, exploring the low signal-to-noise ratio regime would require numerical simulation. However, such numerical simulations require concrete models of signal and noise structure. The manner in which optimal coding deviates from homeostatic coding will not be universal across these models, but depend on each model’s specific details. This limits our ability to draw general conclusions from numerical simulations, which is why we have chosen not to pursue that strategy extensively here.

Thirdly, our analysis here considered a specific class of noise models. However, these noise models include those of particular biological relevance, such as information-limiting noise correlations (Moreno-Bote et al., 2014; Kanitscheider et al., 2015), signatures of which have been found in the cortex (Rumyantsev et al., 2020), and power-law variability (Goris et al., 2014) (see Sec. 2.6). We note that our homeostatic solution is particularly well suited to the case of information-limiting noise correlations, performing optimally in that case (see App. B.5). The core ideas of our framework and derivation could potentially be applied to an even wider class of noise models.

Finally, up to section 2.8, we make no assumptions about what the neural population is attempting to represent via the representational curves. We then apply our framework to a specific theory of Bayesian encoding (namely DDCs), and develop the new idea of Bayes-ratio coding. There is therefore an opportunity to apply our framework to other theories of neural representation, and in particular to alternative theories of Bayesian representations, such as Probabilistic Population Codes (PPCs) (Beck et al., 2007).

In summary, we showed that homeostatic coding can arise from optimal gain modulation for fighting corruption by noise in neural representations. Based on this coding scheme, we derived a novel implementation of probabilistic inference in neural populations, known as Bayes-ratio coding, which can be achieved by divisive normalisation with adaptive weights to invert generative probabilistic models in an adaptive manner. This coding scheme accounts for adaptation effects that are otherwise hard to explain exclusively based on efficient coding theory. These contributions provide important connections between Bayesian representation, efficient coding theory, and neural adaptation.

Methods

Specification of environments for simulations

To generate the K× K signal correlation matrix, ρ(ν), for the different environments in our numerical simulations, in Sec. 2.2, we first generate a covariance matrix Σ(ν), and let ρ(ν) be the corresponding correlation matrix. The covariance matrix Σ(ν) is in turn constructed such that it has (1) the same eigenspectrum for all ν, with eigenvalues that decrease inversely with rank, as observed in V1 (Stringer et al., 2019), and (2) an eigenbasis that varies smoothly with ν. Thus we defined Σ(ν) = U (ν)Λ1U (ν), where Λ1 = diag(1, 1/2, 1/3, …, 1/K) and U (ν) is an orthogonal matrix that was generated as follows. We randomly and independently sample two K × K random iid Gaussian matrices R0, R1 ∼ 𝒩NK×K(0, IK×K) and obtain symmetric matrices and . As is well known, the eigen-basis (represented by an orthogonal matrix) of such a random Gaussian matrix is distributed uniformly (i.e., according to the corresponding Haar measure) over the orthogonal group O(K). For the extreme environments, ν = 0 and 1, we let U (ν) be the matrix of eigenvectors of S0 and S0, when ordered according to eigenvalue rank. For 0 < ν < 1, we let U (ν) be the matrix of the eigenvectors (again ordered according to eigenvalue rank) of the interpolated symmetric matrix S(ν) = (1−ν)S0 + νS1. Thus, for the extreme environments the eigenbases of the covariance matrices Σ(0) and Σ(1) are sampled independently and uniformly (i.e., from the Haar measure), and the eigenbases of Σ(ν) for intermediate environments (i.e., for 0 < ν < 1) smoothly interpolate between these.

In the case of simulations for aligned noise, we also generate a noise correlation matrix W (ν) which has an approximately 1/nγ spectrum. Ideally, W and ρ would have the same eigen-basis. However, this is impossible, since W and ρ are both correlation matrices. Instead, we generate W (ν) by normalising the positive definite matrix U (ν)ΛγU (ν)T where Λγ = diag(1, 1/2γ, 1/3γ, …, 1/Kγ)

Within-cluster optimisation problem

In Sec. 2.3 we showed that in our clustered population model, the firing rates of individual neurons within a cluster optimise the objective Eq. (12). For a cluster of k neurons, this problem is defined by the k× k covariance matrix . In our simulations, we generated this matrix as the the Gram matrix of k random D-dimensional vectors with isotropic Gaussians distributions, i.e., , where δw ∼ 𝒩 (0, ID). (A justification for our construction of the covariance matrix, and an interpretation of the dimensionality D, can be found in App. B.3.5.) Additionally, we fixed the cluster rate to µ = k × 5Hz, such that the average firing rate per single neuron is 5Hz, and fixed σ2 = 1 (corresponding to a Fano factor of 1). For various values of k and D, we used the above method to generate 10, 000 samples of the covariance matrix Cov(ΔΩ). For each such sample, we then solved the problem (12–13) numerically using the cvxpy package in Python (Diamond and Boyd, 2016). Neurons with firing rates less than 10−4Hz had their firing rate set to 0Hz. We aggregated the set of nonzero rates obtained from each of these optimisations to create the histograms in Fig. 8.

Estimating noise scaling parameters from biological data

In this section we discuss biological estimates for the parameters of the general form of mean-variance relationship of spike count cluster responses that we have assumed in our noise models, Four papers estimate this scaled power-law relationship: the findings of those papers are summarised in the first five columns of Table 1.

Note that the time interval of these papers differs from the our assumed coding interval of 0.1s. We therefore must adjust the reported values to be for an interval of this size. We do this by assuming temporal homogeneity over the timescales of interest; specifically we assume that mean and variance are linear in the time interval. We must also adjust the data to account for the fact that our responses are for clusters of k = 20 similarly tuned neurons. We do this by assuming uncorrelated response noise within each cluster, making mean and variance both linear functions of cluster size. Our adjustments do not affect the value of β, but do change the value of the pre-factor . Specifically, we perform the transformation: where δt is the reported time interval, and Δt is our coding interval, Δt = 0.1s. Fig. 13 shows the adjusted values on a scatter plot. Using the adjusted values, σ2, and assuming that µβ = 10 (to maintain a firing rate of 5Hz per neuron), we can find the value of σ2/(βµ)β. These are reported in the furthest right column of Table 1.

Empirically estimated values of the noise scaling parameter β and the base noise-level σ2, after adjusting for coding interval duration and cluster size using the formula Eq. (35). The dashed lines show the values for Poissonian noise (σ2 = 1, β = 1).

Fitting a homeostatic DDC to Benucci et al. 2013

The data we obtained from Benucci et al. (2013) comprised 11 datasets. In each dataset, the neural population was clustered into 12 groups based on preferred orientation. These neural populations were then exposed (in different contexts) to a distribution of 6 to 12 oriented gratings. This distribution was either uniform or biased; in the latter case, one particular grating (arbitrarily defined as 0°) had higher prevalence, by either 30%, 35%, 40%, or 50%. We discarded all datasets with a 50% prevalence, since (Benucci et al., 2013) report that homeostasis was not obtained in this case, and they similarly discarded it from their main analysis. After exposure to the distribution of gratings, the responses (i.e., spike count) of each cluster to a test set of 20 oriented gratings were then measured. Following Benucci et al. (2013) (see that reference for details), we normalized responses via an affine transform so that after exposure to the uniform stimulus ensemble, the normalised response ranged from 0 to 1.

For each dataset, we obtained the average tuning curve in the uniform ensemble context (Fig. 11B, blue) we used a cubic spline to interpolate and up-sample all the population tuning curves in that context. These were then centered at 0° and averaged across populations with different preferred orientations, yielding a single average tuning curve. To fit the resultant curve with a Gaussian (Fig. 11B, red), the width of this curve at height e−2 was divided by 4 to obtain the standard deviation of the Gaussian. In subsequent fitting of the model, was constrained to equal this standard deviation. This reduces the number of free parameters to 2, which we choose to be σπ and σf. We found the best value of these parameters by fitting the curves for repulsion of tuning curves (adaptation-induced change in preferred orientation vs. pre-adaptation preferred orientation – Fig. 12C and F) between the model and experimental data. Specifically, we performed a grid search over values of σπ and σf and chose the pair of values which minimised the sum of the absolute value of the difference between the empirical and model curves.

We obtained the empirical repulsion curves (Fig. 12C) as follows. For each preferred orientation cluster and context (i.e., exposure to a uniform or biased ensemble), a smoothing kernel was applied before using cubic interpolation to generate up-sampled tuning curves. The smoothing kernel was applied to ensure that none of the tuning curves were multimodal (multimodality was a minor effect — reasonably attributable to finite-trial averaging noise — but could have nevertheless introduced relatively large noise in the estimated preferred orientations). The argmax of these tuning curves was found to give the preferred orientation. The preferred orientation was then compared across conditions (uniform ensemble vs biased ensemble) to give a change in preferred orientation. These were then averaged across contexts with the same adaptor probability to obtain the repulsion curve for each adaptor probability.

eLife Assessment

This work derives a valuable general theory unifying theories of efficient information transmission in the brain with population homeostasis. The general theory provides an explanation for firing rate homeostasis at the level of neural clusters with firing rate heterogeneity within clusters. Applying this theory to the primary visual cortex, the authors present solid evidence that accounts for stimulus-specific and neuron-specific adaptation.

Reviewer #1 (Public review):

This work derives a general theory of optimal gain modulation in neural populations. It demonstrates that population homeostasis is a consequence of optimal modulation for information maximization with noisy neurons. The developed theory is then applied to the distributed distributional code (DDC) model of the primary visual cortex to demonstrate that homeostatic DDCs can account for stimulus-specific adaptation.

What I consider to be the most important contribution of this work is the unification of efficient information transmission in neural populations with population homeostasis. The former is an established theoretical framework, and the latter is a well-known empirical phenomenon - the relationship between them has never been fully clarified. I consider this work to be an interesting and relevant step in that direction.

The theory proposed in the paper is rigorous and the analysis is thorough. The manuscript begins with a general mathematical setting to identify normative solutions to the problem of information maximization. It then gradually builds towards questions about approximate solutions, neural implementation and plausibility of these solutions, applications of the theory to specific models of neural computation (DDC), and finally comparisons to experimental data in V1. Such a connection of different levels of abstraction is an obvious strength of this work.

Overall I find this contribution interesting and assess it positively. At the same time, I have three major points of criticism, which I believe the authors should address. I list them below, followed by a number of more specific comments and feedback.

Major comments:

(1) Interpretation of key results and relationship between different parts of the manuscript. The manuscript begins with an information-transmission ansatz which is described as "independent of the computational goal" (e.g. p. 17). While information theory indeed is not concerned with what quantity is being encoded (e.g. whether it is sensory periphery or hippocampus), the goal of the studied system is to *transmit* the largest amount of bits about the input in the presence of noise. In my view, this does not make the proposed framework "independent of the computational goal". Furthermore, the derived theory is then applied to a DDC model which proposes a very specific solution to inference problems. The relationship between information transmission and inference is deep and nuanced. Because the writing is very dense, it is quite hard to understand how the information transmission framework developed in the first part applies to the inference problem. How does the neural coding diagram in Figure 3 map onto the inference diagram in Figure 10? How does the problem of information transmission under constraints from the first part of the manuscript become an inference problem with DDCs? I am certain that authors have good answers to these questions - but they should be explained much better.

(2) Clarity of writing for an interdisciplinary audience. I do not believe that in its current form, the manuscript is accessible to a broader, interdisciplinary audience such as eLife readers. The writing is very dense and technical, which I believe unnecessarily obscures the key results of this study.

(3) Positioning within the context of the field and relationship to prior work. While the proposed theory is interesting and timely, the manuscript omits multiple closely related results which in my view should be discussed in relationship to the current work. In particular:

A number of recent studies propose normative criteria for gain modulation in populations:

- Duong, L., Simoncelli, E., Chklovskii, D. and Lipshutz, D., 2024. Adaptive whitening with fast gain modulation and slow synaptic plasticity. Advances in Neural Information Processing Systems

- Tring, E., Dipoppa, M. and Ringach, D.L., 2023. A power law describes the magnitude of adaptation in neural populations of primary visual cortex. Nature Communications, 14(1), p.8366.

- Młynarski, W. and Tkačik, G., 2022. Efficient coding theory of dynamic attentional modulation. PLoS Biology

- Haimerl, C., Ruff, D.A., Cohen, M.R., Savin, C. and Simoncelli, E.P., 2023. Targeted V1 co-modulation supports task-adaptive sensory decisions. Nature Communications

- The Ganguli and Simoncelli framework has been extended to a multivariate case and analyzed for a generalized class of error measures:

- Yerxa, T.E., Kee, E., DeWeese, M.R. and Cooper, E.A., 2020. Efficient sensory coding of multidimensional stimuli. PLoS Computational Biology

- Wang, Z., Stocker, A.A. and Lee, D.D., 2016. Efficient neural codes that minimize LP reconstruction error. Neural Computation, 28(12),

More detailed comments and feedback:

(1) I believe that this work offers the possibility to address an important question about novelty responses in the cortex (e.g. Homann et al, 2021 PNAS). Are they encoding novelty per-se, or are they inefficient responses of a not-yet-adapted population? Perhaps it's worth speculating about.

(2) Clustering in populations - typically in efficient coding studies, tuning curve distributions are a consequence of input statistics, constraints, and optimality criteria. Here the authors introduce randomly perturbed curves for each cluster - how to interpret that in light of the efficient coding theory? This links to a more general aspect of this work - it does not specify how to find optimal tuning curves, just how to modulate them (already addressed in the discussion).

(3) Figure 8 - where do Hz come from as physical units? As I understand there are no physical units in simulations.

(4) Inference with DDCs in changing environments. To perform efficient inference in a dynamically changing environment (as considered here), an ideal observer needs some form of posterior-prior updating. Where does that enter here?

(5) Page 6 - "We did this in such a way that, for all ν, the correlation matrices, ρ(ν), were derived from covariance matrices with a 1/n power-law eigenspectrum (i.e., the ranked eigenvalues of the covariance matrix fall off inversely with their rank), in line with the findings of Stringer et al. (2019) in the primary visual cortex." This is a very specific assumption, taken from a study of a specific brain region - how does it relate to the generality of the approach?

Reviewer #2 (Public review):

Summary:

Using the theory of efficient coding, the authors study how neural gains may be adjusted to optimize coding by noisy neural populations while minimizing metabolic costs. The manuscript first presents mathematical results for the general case where the computational goals of the neural population are not specified (the computation is implicit in the assumed tuning curves) and then develops the theory for a specific probabilistic coding scheme. The general theory provides an explanation for firing rate homeostasis at the level of neural clusters with firing rate heterogeneity within clusters, and the specific application further captures stimulus-specific and neuron-specific adaptation in the visual cortex.

The mathematical derivations, simulations, and application to visual cortex data are solid as far as I can tell.

In the current format, the significance is difficult to assess fully: the manuscript is a bit sprawling, in the first half the general theory is lengthy and technical, and then in the second half a few phenomena are addressed without a clear relation between them (rate homeostasis, rate heterogeneity, synaptic homeostasis, V1 adaptation, divisive normalization), requiring several ad-hoc choices and assumptions.

Strengths:

The problem of efficient coding is a long-standing and important one. This manuscript contributes to that field by proposing a theory of efficient coding through gain adjustments, independent of the computational goals of the system. The main result is a normative explanation for firing rate homeostasis at the level of neural clusters (groups of neurons that perform a similar computation) with firing rate heterogeneity within each cluster. Both phenomena are widely observed, and reconciling them under one theory is important.

The mathematical derivations are thorough as far as I can tell. Although the model of neural activity is artificial, the authors make sure to include many aspects of cortical physiology, while also keeping the models quite general.

Section 2.5 derives the conditions in which homeostasis would be near-optimal in the cortex, which appear to be consistent with many empirical observations in V1. This indicates that homeostasis in V1 might be indeed close to the optimal solution to code efficiently in the face of noise.

The application to the data of Benucci et al 2013 is the first to offer a normative explanation of stimulus-specific and neuron-specific adaptation in V1.

Weaknesses:

The novelty and significance of the work are not presented clearly. The relation to other theoretical work, particularly Ganguli and Simoncelli and other efficient coding theories, is explained in the Discussion but perhaps would be better placed in the Introduction, to motivate some of the many choices of the mathematical models used here.

The manuscript is very hard to read as is, it almost feels like this could be two different papers. The first half seems like a standalone document, detailing the general theory with interesting results on homeostasis and optimal coding. The second half, from Section 2.7 on, presents a series of specific applications that appear somewhat disconnected, are not very clearly motivated nor pursued in-depth, and require ad-hoc assumptions.

For instance, it is unclear if the main significant finding is the role of homeostasis in the general theory or the demonstration that homeostatic DDC with Bayes Ratio coding captures V1 adaptation phenomena. It would be helpful to clarify if this is being proposed as a new/better computational model of V1 compared to other existing models.

Early on in the manuscript (Section 2.1), the theory is presented as general in terms of the stimulus dimensionality and brain area, but then it is only demonstrated for orientation coding in V1.

The manuscript relies on a specific response noise model, with arbitrary tuning curves. Using a population model with arbitrary tuning curves and noise covariance matrix, as the basis for a study of coding optimality, is problematic because not all combinations of tuning curves and covariances are achievable by neural circuits (e.g. https://pubmed.ncbi.nlm.nih.gov/27145916/ )

The paper Benucci et al 2013 shows that homeostasis holds for some stimulus distributions, but not others i.e. when the 'adapter' is present too often. This manuscript, like the Benucci paper, discards those datasets. But from a theoretical standpoint, it seems important to consider why that would be the case, and if it can be predicted by the theory proposed here.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1(Public Review):

Major comments:

(1) Interpretation of key results and relationship between different parts of the manuscript. The manuscript begins with an information-transmission ansatz which is described as ”independent of the computational goal” (e.g. p. 17). While information theory indeed is not concerned with what quantity is being encoded (e.g. whether it is sensory periphery or hippocampus), the goal of the studied system is to *transmit* the largest amount of bits about the input in the presence of noise. In my view, this does not make the proposed framework ”independent of the computational goal”. Furthermore, the derived theory is then applied to a DDC model which proposes a very specific solution to inference problems. The relationship between information transmission and inference is deep and nuanced. Because the writing is very dense, it is quite hard to understand how the information transmission framework developed in the first part applies to the inference problem. How does the neural coding diagram in Figure 3 map onto the inference diagram in Figure 10? How does the problem of information transmission under constraints from the first part of the manuscript become an inference problem with DDCs? I am certain that authors have good answers to these questions - but they should be explained much better.

We are very thankful to the reviewer for highlighting the potential confusion surrounding these issues, in particular the relationship between the two halves of the paper – which was previously exacerbated by the length of the paper. We have now added further explanations at different points within the manuscript to better disentangle these issues and clarify our key assumptions. We have also significantly cut the length of the paper by moving more technical discussions to the Methods or Appendices. We will summarise these changes here and also clarify the rationale for our approach and point out potential disagreements with the reviewer.

Key to our approach is that we indeed do not assume the entire goal of the studied neural system (whether part of the sensory system or not) is to transmit the largest amount of information about the stimulus input (in the presence of noise). In fact, general computations, including the inference of latent causes of inputs, often require filtering out or ignoring some information in the sensory input. It is thus not plausible that tuning curves in general (i.e. in an arbitrary part of the nervous system) are optimised solely with regards to the criterion of information transmission. Accordingly we do not assume they are entirely optimised for that purpose. However, we do make a key assumption or hypothesis (which like any hypothesis might turn out to be partly or entirely wrong): that (1) a minimal feature of the tuning curve (its scale or gain) is entirely free to be optimised for the aim of information transmission (or more precisely the goal of combating the detrimental effect of neural noise on coding fidelity), (2) other aspects of the population tuning curve structure (i.e. the shape of individual tuning curves and their arrangement across the population) are determined by (other) computational goals beyond efficient coding. (Conceptually, this is akin to the modularization between indispensible error correction and general computations in a digital computer, and the need for the former to be performed in a manner that is agnostic as to the computations performed.) We have added two paragraphs in the manuscript which present the above rationale and our key hypothesis or assumption. The first of these was added to the (second paragraph of the) Introduction section, and the second is a new paragraph following Eq. 1 (which is about the gain-shape decomposition of the tuning curves, and the optimisation of the former based on efficient coding) of Results.

Our paper can be divided into two parts. In the first part, we develop a general, computationally agnostic (in the above sense, just as in the digital computer example), efficient coding theory. In the second part, we apply that theory to a specific form of computation, namely the DDC framework for Bayesian inference. The latter theory now determines the tuning curve shapes. When combined with the results of the first part (which dictate the tuning curve scale or gain according to efficient coding theory), this “homeostatic DDC” model makes full predictions for the tuning curves (i.e., both scale and shape) and how they should adapt to stimulus statistics.

So to summarise, it is not the case that the problem of information transmission (or rather mitigating the effect noise on coding fidelity under metabolic constraints), dealt with in the first part, has become a problem of Bayesian inference. But rather, the dictates of efficient coding for optimal gains for coding fidelity (under constraints) have been applied to and combined with a computational theory of inference.

We have added new expository text before and after Eq. 17 in Sec. 2.7 (at the beginning of the second part of the paper on homeostatic DDCs) to again make the connection with the first part and the rationale for its combination with the original DDC framework more clear.

With the changes outlined above, we believe and hope the connection between the two parts (which we agree with the reviewer, was indeed rather obscure previously) has been adequately clarified.

(2) Clarity of writing for an interdisciplinary audience. I do not believe that in its current form, the manuscript is accessible to a broader, interdisciplinary audience such as eLife readers. The writing is very dense and technical, which I believe unnecessarily obscures the key results of this study.

We thank the reviewer for this comment. We have taken several steps to improve the accessibility of this work for an interdisciplinary audience. Firstly, several sections containing dense, mathematical writing have now been moved into appendices or the Methods section, out from the main text; in their place we have made efforts to convey the core of the results, and to providing intuitions, without going into unnecessary technical detail. Secondly, we have added additional figures to help illustrate key concepts or assumptions (see Fig. 1B clarifying the conceptual approach to efficient coding and homeostatic adaptation, and Fig. 8A describing the clustered population). Lastly, we have made sure to refer back to the names of symbols more often, so as to make the analysis easier to follow for a reader with an experimental background.

(3) Positioning within the context of the field and relationship to prior work. While the proposed theory is interesting and timely, the manuscript omits multiple closely related results which in my view should be discussed in relationship to the current work. In particular, a number of recent studies propose normative criteria for gain modulation in populations: • Duong, L., Simoncelli, E., Chklovskii, D. and Lipshutz, D., 2024. Adaptive whitening with fast gain modulation and slow synaptic plasticity. Advances in Neural Information Processing Systems

Tring, E., Dipoppa, M. and Ringach, D.L., 2023. A power law describes the magnitude of adaptation in neural populations of primary visual cortex. Nature Communications, 14(1), p.8366.

Ml ynarski, W. and Tkaˇcik, G., 2022. Efficient coding theory of dynamic attentional modulation. PLoS Biology

Haimerl, C., Ruff, D.A., Cohen, M.R., Savin, C. and Simoncelli, E.P., 2023. Targeted V1 co-modulation supports task-adaptive sensory decisions. Nature Communications • The Ganguli and Simoncelli framework has been extended to a multivariate case and analyzed for a generalized class of error measures:

Yerxa, T.E., Kee, E., DeWeese, M.R. and Cooper, E.A., 2020. Efficient sensory coding of multidimensional stimuli. PLoS Computational Biology

Wang, Z., Stocker, A.A. and Lee, D.D., 2016. Efficient neural codes that minimize LP reconstruction error. Neural Computation, 28(12),

We thank the reviewer again for bringing these works to our attention. For each, we explain whether we chose to include them in our Discussion section, and why.

(1) Duong et al. (2024): We decided not to discuss this manuscript, as our assessment is that it is very relevant to our work. That study starts with the assumption that the goal of the sensory system under study is to whiten the signal covariance matrix, which is not the assumption we start with. A mechanistic ingredient (but not the only one) in their approach is gain modulation. However, in their case it is the gains of computationally auxiliary inhibitory neurons that is modulated and not (as in our case) the gain the (excitatory) coding neurons (i.e. those which encode information about the stimulus and whose response covariance is whitened). These key distinction make the connection with our work quite loose and we did not discuss this work.

(2) Tring et al. (2023): We have added a discussion of the results of this paper and its relationship to the results of our work and that of Benucci et al. This appears in the 7th paragraph of the Discussion. This study is indeed highly relevant to our paper, as it essentially replicates the Benucci et al. experiment, this time in awake mice (rather than anesthetised cats). However, in contrast to the resul‘ts of Benucci et al., Tring et al. do not find firing rate homeostasis in mouse V1. A second, remarkable finding of Tring et al. is that adaptation mainly changes the scale of the population response vector, and only minimally affects its direction. While Tring et al. do not portray it as such, this behaviour amounts to pure stimulus-specific adaptation without the neuron-specific factor found in the Benucci et al. results (see Eq. 24 of our manuscript). As we discuss in our manuscript, when our homeostatic DDC model is based on an ideal-observer generative model, it also displays pure stimulus-specific adaptation with no neuronal factor. Our final model for Benucci’s data did contain a neural factor, because we used a non-ideal observer DDC (in particular, we assumed a smoother prior distribution over orientations compared to the distribution used in the experiment - which has a very sharp peak – as it is more natural given the inductive biases we expect in the brain). The resultant neural factor suppresses the tuning curves tuned to the adaptor stimulus. Interestingly, when gain adaptation is incomplete, and happens to a weaker degree compared to what is necessary for firing rate homeostasis, an additional neural factor emerges that is greater than one for neurons tuned to the adaptor stimulus. These two multiplicative neural factors can approximately cancel each other; such a theory would thus predict both deviation from homeostasis and approximately pure stimulus-specific adaptation. We plan to explore this possibility in future work.

(3) Ml ynarski and Tkaˇcik (2022): We are now citing and discussing this work in the Discussion (penultimate paragraph), in the context of a possible future direction, namely extending our framework to cover the dynamics of adaptation (via a dynamic efficient gain modulation and dynamic inference). We have noted there that Mlynarski have used such a framework (which while similar has key technical differences with our approach) based on a task-dependent efficient coding objective to model top-down attentional modulation. By contrast, we have studied bottom-up and task-independent adaptation, and it would be interesting to extend our framework and develop a model to make predictions for the temporal dynamics of such adaptation.

(4) Haimerl et al. (2023): We have elected not to include this work within our discussion either, as we do not believe it is sufficiently relevant to our work to warrant inclusion. Although this paper also considers gain modulation of neural activity, the setting and the aims of the theoretical work and the empirical phenomena it is applied to are very different from our case in various ways. Most importantly, this paper is not offering a normative account of gain modulation; rather, gain modulation is used as a mechanism for enabling fast adaptive readouts of task relevant information.

(5) Yerxa et al. (2020): We have now included a discussion of this paper in our Discussion section. Note that, even though this study generalises the Ganguli and Simoncelli framework to higher diemsnions, just like that paper it still places strict requirements (which are arguably even more stringent in higher dimensions) on the form of the tuning curves in the population, viz. that there exists a differentiable transform of the stimulus space which renders these unimodal curves completely homogeneous (i.e., of the same shape, and placed regularly and with uniform density).

(6) Wang et al. (2016): We have included this paper in our discussion as well. As above, this paper does not consider general tuning curves, and places the same constraint on their shape and arrangement as in Ganguli and Simoncelli paper.

More detailed comments and feedback:

(1) I believe that this work offers the possibility to address an important question about novelty responses in the cortex (e.g. Homann et al, 2021 PNAS). Are they encoding novelty per-se, or are they inefficient responses of a not-yet-adapted population? Perhaps it’s worth speculating about.

We are not sure why the relatively large responses to “novel” or odd-ball stimuli should be considered inefficient or unadapted: in the context in which those stimuli are infrequent odd-balls (and thus novel or surprising when occurring), efficient coding theory would indeed typically predict a large response compared to the (relatively suppressed) responses to frequently occurring stimuli. Of course, if the statistics change and the odd-ball stimulus now becomes frequent, adaptation should occur and would be expected to suppress responses to this stimulus. As to the question of whether (large) responses to infrequent stimuli can or should be characterised as novelty responses: this is partly an interpretational or semantic issue – unless it is grounded in knowledge of how downstream populations use this type of coding in V1, which could then provide a basis for solidly linking them to detection of novelty per se. In short, our theory, could be applied to Homann et al.’s data, but we consider that beyond the scope of the current paper.

(2) Clustering in populations - typically in efficient coding studies, tuning curve distributions are a consequence of input statistics, constraints, and optimality criteria. Here the authors introduce randomly perturbed curves for each cluster - how to interpret that in light of the efficient coding theory? This links to a more general aspect of this work - it does not specify how to find optimal tuning curves, just how to modulate them (already addressed in the discussion).

We begin by addressing the reviewer’s more general concern regarding the fact that our theory does not address the problem of finding optimal tuning curves, only that of modulating them optimally. As we expound within the updated version of the paper (see the newly expanded 3rd paragraph in Sec. 2.1 and the expanded 2nd paragraph in Introduction), it is not plausible that the sole function of sensory systems, and neural circuits more generally, is the transmission of information. There are many other computational tasks which must be performed by the system, such as the inference of the latent causes of sensory inputs. For many such tasks, it is not even desirable to have complete transmission of information about the external stimulus, since a substantial portion of that information is not important for the task at hand, and must be discarded. For example, such discarding of information is the basis of invariant representations that occur, e.g., in higher visual areas. So we recognise that tuning curve shapes are in general dictated and shaped by computational goals beyond transmission of information or error correction. As such, we have remained agnostic as to the computational goals of neural systems and therefore the shape of the tuning curve. We have made the assumption and adopted the postulate that those computational goals determine the shape of the tuning curves, leaving the gains to be adjuted freely for the purpose of mitigating the effect noise on coding fidelity (this is similar to how error correction is done in computers independendently of the computations performed). by assuming that those computational goals are captured adequately by the shape of tuning curves, this leaves us free to optimise the gains of those curves for purely information theoretic objectives. Finally, we note that the case where the tuning curve shapes are additionally optimised for information transmission is a special case of our more general approach. For further discussion, see the updated version of our introduction.

We now turn to our choice to model clusters using random perturbations. This is, of course, a toy model for clustering tuning curves within a population. With this toy model we are attempting to capture the important aspects of tuning curve clusters within the population while not over-complicating the simulations. Within any neural population, there will be tuning curves that are similar; however, such curves will inevitably be heterogeneous, as opposed to completely identical. Thus, when we cluster together similar curves there will be an “average” cluster tuning curve (found by, e.g., normalising all individual curves and taking the average), which all other tuning curves within the cluster are deviations from. The random perturbations we apply are our attempt to capture these deviations. However, note that the perturbations are not fully random, but instead have an “effective dimensionality” which we vary over. By giving the perturbations an effective dimensionality, we aim to capture the fact that deviations from the average cluster tuning curve may not be fully random, and may display some structure.

(3) Figure 8 - where do Hz come from as physical units? As I understand there are no physical units in simulations.

We have clarified this within the figure caption. The within-cluster optimisation problem requires maximising a quadratic program subject to a constraint on the total mean spike count of the cluster. The objective for the quadratic program is however mathematically homogeneous. So we can scale the variables and parameters in a consistent to be in units of Hz – i.e., turn them into mean firing rates, instead of mean spike counts, with an assumption on the length of the coding time interval. We fix this cluster firing rate to be k × 5 Hz, so that the average single-neuron firing rate is 5 Hz (based on empirical estimates – see our Sec. 2.5). This agrees with our choice of µ in our simulations (i.e., µ = 10) if we assume a coding interval of 0.1 seconds.

(4) Inference with DDCs in changing environments. To perform efficient inference in a dynamically changing environment (as considered here), an ideal observer needs some form of posterior-prior updating. Where does that enter here?

A shortcoming of our theory, in its current form, is that it applies only to the system in “steady-state”, without specifying the dynamics of how adaptation temporlly evolves (we assume the enrivonment has periods of relative stability that are of relatively long duration compared to the dynamical timescales of adaptation, and consider the properties of the well-adapted steady state population). Thus our efficient coding theory (which predicts homeostatic adaptation under the outlined conditions) is silent on the time-course over which homeostasis occurs. Likewise, the DDC theory (in its original formulation in Vertes & Sahani) is silent on dynamic updating of posteriors and considers only static inference with a fixed internal model. We have now discuss a new future directoin in the Discussion (where we cite the work of Mlynarski and Tkacik) to point out that our theory can in principle be extended (based on dynamic inference and efficient coding) to account for the dynamics of attention, but this is beyond the scope of the current work.

(5) Page 6 - ”We did this in such a way that, for all , the correlation matrices, (), were derived from covariance matrices with a 1/n power-law eigenspectrum (i.e., the ranked eigenvalues of the covariance matrix fall off inversely with their rank), in line with the findings of Stringer et al. (2019) in the primary visual cortex.” This is a very specific assumption, taken from a study of a specific brain region - how does it relate to the generality of the approach?

Our efficient coding framework has been formulated without relying on any specific assumptions about the form of the (signal or noise) correlation matrices in cortex. The homeostatic solution to this efficient coding problem, however, emerges under certain conditions. But, as we demonstrate in our discussion of the analytic solutions to our efficient coding objective and the conditions necessary for the validity of the homeostatic solution, we expect homeostasis to arise whenever the signal geometry is sufficiently high-dimensional (among other conditions). By this we mean that the fall-off of the eigenvalues of the signal correlation matrix must be sufficiently slow. Thus, a fall-off in the eigenvalue spectrum slower than 1/n would favor homeostasis even more than our results. If the fall off was faster, then whether or not (and to what degree) firing rate homeostasis becomes suboptimal depends on factors such as the fastness of the fall-off and also the size of the population. Thus (1) rate homeostasis does not require the specific 1/n spectrum, but that spectrum is consistent with the conditions for optimality of rate homeostasis, (2) in our simulations we had to make a specific choice, and relying on empirical observations in V1 was of course a well-justified choice (moreover, as far as we are aware, there have been no other studies that have characterised the spectrum of the signal covariance matrix in response to natural stimuli, based on large population recordings).

Reviewer #2 (Public Review):

Strengths:

The problem of efficient coding is a long-standing and important one. This manuscript contributes to that field by proposing a theory of efficient coding through gain adjustments, independent of the computational goals of the system. The main result is a normative explanation for firing rate homeostasis at the level of neural clusters (groups of neurons that perform a similar computation) with firing rate heterogeneity within each cluster. Both phenomena are widely observed, and reconciling them under one theory is important.

The mathematical derivations are thorough as far as I can tell. Although the model of neural activity is artificial, the authors make sure to include many aspects of cortical physiology, while also keeping the models quite general.

Section 2.5 derives the conditions in which homeostasis would be near-optimal in the cortex, which appear to be consistent with many empirical observations in V1. This indicates that homeostasis in V1 might be indeed close to the optimal solution to code efficiently in the face of noise.

The application to the data of Benucci et al 2013 is the first to offer a normative explanation of stimulus-specific and neuron-specific adaptation in V1.

We thank the reviewer for these assessments.

Weaknesses:

The novelty and significance of the work are not presented clearly. The relation to other theoretical work, particularly Ganguli and Simoncelli and other efficient coding theories, is explained in the Discussion but perhaps would be better placed in the Introduction, to motivate some of the many choices of the mathematical models used here.

We thank the reviewer for this comment; we have updated our introduction to make clearer the relationship between this work and previous works within efficient coding theory. Please see the expanded 2nd paragraph of Introduction which gives a short account of previous efficient coding theories and now situates our work and differentiates it more clearly from past work.

The manuscript is very hard to read as is, it almost feels like this could be two different papers. The first half seems like a standalone document, detailing the general theory with interesting results on homeostasis and optimal coding. The second half, from Section 2.7 on, presents a series of specific applications that appear somewhat disconnected, are not very clearly motivated nor pursued in-depth, and require ad-hoc assumptions.

We thank the reviewer for this suggestion. The reviewer is right to note that our paper contains both the exposition of a general efficient coding theory framework in addition to applications of that framework. Following your advice we have implemented the following changes. (1) significantly shortened or entirely moved some of the less central results in the second half of Results, to the Methods or appendices (this includes the entire former section 2.7 and significant shortening of the section on implementation of Bayes ratio coding by divisive normalisation). (2) We have added a new figure (Fig 1B) and two long pieces of text to the (2nd paragraph of) Introduction, after Eq. (1), and in Sec. 2.7 (introducing homeostatic DDCs) to more clearly explain and clarify the assumptions underlying our efficient coding theory, and its connection with the second half of the Results (i.e. application to DDC theory of Bayesian inference), and better motivate why we consider the homeostatic DDC.

For instance, it is unclear if the main significant finding is the role of homeostasis in the general theory or the demonstration that homeostatic DDC with Bayes Ratio coding captures V1 adaptation phenomena. It would be helpful to clarify if this is being proposed as a new/better computational model of V1 compared to other existing models.

We see the central contribution of our work as not just that homeostasis arises as a result of an efficient coding objective, but also that this homeostasis is sufficient to explain V1 adaptation phenomena - in particular, stimulus specific adaptation (SSA) - when paired with an existing theory of neural representation, the DDC (itself applied to orientation coding in V1). Homeostatic adaptation alone does not explain SSA; nor do DDCs. However, when the two are combined they provide an explanation for SSA. This finding is significant, as it unifies two forms of adaptation (SSA and homeostatic adaptation) whose relationship was not previously appreciated. Our field does not currently have a standard model of V1, and we do not claim to have provided one either; rather, different models have captured different phenomena in V1, and we have done so for homeostatic SSA in V1.

Early on in the manuscript (Section 2.1), the theory is presented as general in terms of the stimulus dimensionality and brain area, but then it is only demonstrated for orientation coding in V1.

The efficient coding theory developed in Section 2 is indeed general throughout, we make no assumptions regarding the shape of the tuning curves or the dimensionality of the stimulus. Further, our demonstrations of the efficient coding theory through numerical simulations - make assumptions only about the form of the signal and noise covariance matrices. When we later turn our attention away from the general case, our choice to focus on orientation coding in V1 was motivated by empirical results demonstrating a co-occurrence of neural homeostasis and stimulus specific adaptation in V1.

The manuscript relies on a specific response noise model, with arbitrary tuning curves. Using a population model with arbitrary tuning curves and noise covariance matrix, as the basis for a study of coding optimality, is problematic because not all combinations of tuning curves and covariances are achievable by neural circuits (e.g. https://pubmed.ncbi.nlm.nih.gov/27145916/ )

First, to clarify, our theory allows for complete generality of neural tuning curve shapes, and assumes a broad family of noise models (which, while not completely arbitrary, includes cases of biological relevance and/or models commonly used in the theoretical literature). Within this class of noise covariance models, we have shown numerical results for different values for different parameters of the noise covariance model, but more importantly, have analytically outlined the general properties and requirements on noise strength and structure (and its relationship to tuning curves and signal structure) under which homeostatic adaptation would be optimal. Regarding the point that not all combinations of tuning curves and noise covariances occur in biology or are achievable by neural circuits: (1) If we are guessing correctly the specific point of the reviewer’s reference to the review paper by Kohn et al. 2016, we have in fact prominently discussed the case of information limiting noise which corresponds to a specific relationship between signal structure (as determined by tuning curves) and noise structure (as specified by the noise covariance matrix). Our family of noise models include that biologically relevant case and we have indeed paid it particular attention in our simulations and discussions (see discussion of Fig. 7 in Sec. 2.3, and that of aligned noise in Sec. 2.5). (2) As for the more general or abstract point that not all combinations of noise covariance and tuning curve structures are achievable by neural circuits, we can make the following comments. First, in lieu of a full theoretical or empirical understanding of the achievable combinations (which does not exist), we have outlined conditions for homeostatic adaptations under a broad class of noise models and arbitrary tuning curves. If some combinations within this class are not realised in biology, that does not invalidate the theoretical results, as the latter have been derived under more general conditions, which nevertheless include combinations that do occur in biology and are achievable by neural circuits (which, as pointed out, include the important case of aligned noise and signal structure – as reviewed in Kohn et al.– to which we have paid particular attention).

The paper Benucci et al 2013 shows that homeostasis holds for some stimulus distributions, but not others i.e. when the ’adapter’ is present too often. This manuscript, like the Benucci paper, discards those datasets. But from a theoretical standpoint, it seems important to consider why that would be the case, and if it can be predicted by the theory proposed here.

The theory we provide predicts that, under certain (specified) conditions, we ought to see deviation from exact homeostatic results; indeed, we provide a first order approximation to the optimal gains in this case which quantifies such deviations when they are small. However, unfortunately the form of this deviation depends on a precise choice of stimulus statistics (e.g. the signal correlation matrix, the noise correlation matrix averaged over all stimulus space, and other stimulus statistics), in contrasts to the universality of the homeostatic solution, when it is a valid approximation. In our model of Benucci et al.’s experiment, we restrict to a simple one-dimensional stimulus space (corresponding to orientated gratings), without specifying neural responses to all stimuli; as such, we are not immediately able to make predictions about whether the homeostatic failure can be predicted using the specific form of deviation from homeostasis. However, we acknowledge that this is a weakness of our analysis, and that a more complete investigation would address this question. For reasons of space, we elected not to pursue this further. We have added a paragraph to our Discussion (8th paragraph) explaining this.

Reviewer#1 (Recommendations for the authors):

(1) To make the article more accessible I would suggest the following:

(a) Include a few more illustrations or diagrams that demonstrate key concepts: adaptationof an entire population, clustering within a population, different sources of noise, inference with homeostatic DDCs, etc.

We thank the reviewer for this suggestion - we have added an additional figure in (Figure 8, Panel A) to explain the concept of clustering within a population. We also added a new panel to Figure 1 (Figure 1B) which we hope will clarify the conceptual postulate underlying our efficient coding framework and its link to the second half of the paper.

(b) Within the text refer to names of quantities much more often, rather than relying onlyon mathematical symbols (e.g. w,r,Ω, etc).

We thank the reviewer for the suggestion; we have updated the text accordingly and believe this has improved the clarity of the exposition.

(2) It is hard to distill which components of the considered theory are crucial to reproducing the experimental observations in Figure 12. Is it the homeostatic modulation, efficient coding, DDCs, or any combination of those or all of them necessary to reproduce the experiment? I believe this could be explained much better, also with an audience of experimentalists in mind.

We have updated the text to provide additional clarity on this matter (see the pointers to these changes and additions in the revised manuscript, given above in response to your first comment). In particular, reproducing the experimental results requires combining DDCs with homeostatic modulation – with the latter a consequence of our efficient coding theory, and not an independent ingredient or assumption.

(3) It would be good to comment on how sensitive the results are to the assumptions made, parameter values, etc. For example: do conclusions depend on statistics of neural responses in simulated environments? Do they generalize for different values of the constraint µ? This could be addressed in the discussion / supplementary material.

This issue is already discussed extensively within the text - see Sec. 2.4, Analytical insight on the optimality of homeostasis, and Sec. 2.5, Conditions for the validity of the homeostatic solution to hold in cortex. In these sections, we outline that - provided a certain parameter combination is small - we expect the homeostatic result to hold. Accordingly, we anticipate that our numerical results will generalise to any settings in which that parameter combination remains small.

(4) How many neurons/units were used for simulations?

We apologies for omitting this detail; we used 10,000 units for our simulations. We have edited both the main text and the methods section to reflect this.

(5) Typos etc: a) Figure 5 caption - the order of panels B and C is switched. b) Figure 6A - I suggest adding a colorbar.

Thank you. We have relabelled the panels B and C in the appropriate figures so that the ordering in the figure caption is correct. We feel that a colourbar in figure 6A would be unnecessary, since we are only trying to convey the concept of uniform correlations, rather than any particular value for the correlations; as such we have elected not to add a colourbar. We have, however, added a more explicit explanation of this cartoon matrix in the figure caption, by referring to the colors of diagonal vs off-diagonal elements.

Reviewer#2 (Recommendations for the authors):

The text on page 10, with the perturbation analysis, could be moved to a supplement, leaving here only the intuition.

We thank the reviewer for this suggestion; we have moved much of the argument into the appendix so as to not distract the reader with unnecessary technical details.

Text before eq. 12 “...in cluster a maximize the objective...” should be ‘minimize’?

The cluster objective as written is indeed maximised, as stated in the text. Note that, in the revised manuscript, this argument has been moved to an appendix to reduce the density of mathematics in the main text.

Top of page 25 “S0 and S0” should be “S0 and S1”?

Thank you, we have corrected the manuscript accordingly.

Version 2

Abstract

Abstract

In the absence of adaptation, the average firing rate of neurons would rise or drop when changes in the environment make their preferred stimuli more or less prevalent. However, by adjusting the responsiveness of neurons, adaptation can yield firing rate homeostasis and stabilise the average rates of neurons at fixed levels, despite changes in stimulus statistics. In sensory cortex, adaptation is typically also stimulus specific, in that neurons reduce their responsiveness to over-represented stimuli, but maintain or even increase their responsiveness to stimuli far from over-represented ones. Here, we present a normative explanation of firing rate homeostasis grounded in the efficient coding principle. Specifically, we show that this homeostasis can arise when neurons adapt their responsiveness to optimally mitigate the effect of neural noise on population coding fidelity, at minimal metabolic cost. Unlike previous efficient coding theories, we formulate the problem in a computation-agnostic manner, enabling our framework to apply far from the sensory periphery. We then apply this general framework to Distributed Distributional Codes, a specific computational theory of neural representations serving Bayesian inference. We demonstrate how homeostatic coding, combined with such Bayesian neural representations, provides a normative explanation for stimulus-specific adaptation, widely observed across the brain, and how this coding scheme can be accomplished by divisive normalisation with adaptive weights. Further, we develop a model within this combined framework, and, by fitting it to previously published experimental data, quantitatively account for measures of stimulus-specific and homeostatic adaption in the primary visual cortex.

Introduction

Neural population responses, and thus the computations on sensory inputs represented by them, are corrupted by noise. The extent of this corruption can, however, be modulated by changing neural gains. When the gain of a neuron increases, the strength of its response noise, or trial-to-trial variability, typically grows sublinearly and more slowly than its average response; for example, for Poisson-like firing, response noise grows as the square root of the mean response. The neuron can therefore increase its signal-to-noise ratio by increasing its gain or responsiveness. However, this increase in coding fidelity comes at the cost of an elevated average firing rate, and thus higher metabolic energy expenditure. From a normative perspective, coding fidelity and metabolic cost are thus two conflicting forces.

We start this study by asking what is the optimal way of adjusting neural gains in order to combat noise with minimal metabolic cost, irrespective of the computations represented by the neural population. As we will see, the answer depends on the prevailing stimulus statistics. We will thus address the following more specific question: given a neural population with arbitrary tuning curve shapes and general noise distribution, how should neurons optimally adjust their gains depending on the stimulus statistics prevailing in the environment? To address this question, we use efficient coding theory as our normative framework (Attneave, 1954; Nadal and Parga, 1999; Laughlin, 1981; Barlow, 2012; Linsker, 1988; Ganguli and Simoncelli, 2014; Wei and Stocker, 2015; Atick and Redlich, 1990). Efficient coding theories formalise the problem of optimal coding subject to biological constraints and costs, and by finding the optimal solution make predictions about the behavior of nervous systems (which are postulated to have been approximately optimised, e.g., via natural selection). Concretely, the Infomax Principle (Linsker, 1988; Atick and Redlich, 1990; Ganguli and Simoncelli, 2014; Wei and Stocker, 2015) states that sensory systems optimise the mutual information between a noisy neural representation and an external stimulus, subject to metabolic constraints, e.g., on the energy cost of spiking activity. To that end, the neural system should exploit the statistics of stimuli in its local environment (Simoncelli and Olshausen, 2001). Efficient coding theories can therefore predict how sensory systems should optimally adapt to changes in environmental stimulus statistics. There is, however, a key difference between the approach we will adopt and classic applications of the efficient coding principle. In most applications, this principle has been applied to systems located early in the sensory stream, and information transmission is taken to be the exclusive goal of the system. Accordingly, the entire input-output transformations performed by the system is assumed to have been optimised (under metabolic or anatomical constraints) purely for this goal. Based on this assumption, such theories typically make predictions about patterns of neural selectivity, including the shapes and arrangement of neural tuning curves. However, for general brain regions located deeper within the sensory-motor pathway, information transmission is far from an adequate charactrisation of the system’s computational goals. For example, many neural systems perform computations which require systematically discarding information contained in the stimulus input. The discarded information may correspond to nuisance variables, or be irrelevant to the behavioural tasks faced by the organism. Here, we make the key assumption that the shapes and arrangement of tuning curves are determined by computational goals — beyond mitigating the effect of noise on coding fidelity — to which our efficient coding framework remains agnostic. We therefore do not optimise these aspects of the population code and allow them to be arbitrary. On the other hand, we assume the neural gains (which set the scale of the tuning curves) are adapted and optimised purely for the purpose of fighting noise corruption at minimal metabolic cost, regardless of the system’s computational goals. Conceptually, this division of labour is akin to the separation, in digital computers, between error correction (which is indispensable in practice) and general computations, and the need for the former to be performed in a manner agnostic to the ongoing computations.

Suppose an environmental shift makes a stimulus feature more prevalent. Neurons sensitive to that feature will then see their preferred feature more often, and thus their average firing rate will initially increase. However, typically, those neurons gradually adapt by reducing their responsiveness or gain (Solomon and Kohn, 2014; Clifford et al., 2007; Benucci et al., 2013). Such gain adaptation can lead to firing rate homeostasis (Desai, 2003; Turrigiano and Nelson, 2004; Maffei and Turrigiano, 2008; Hengen et al., 2013), whereby adaptation brings the average firing rate back to the level prior to the environmental shift (Fig. 1A). Thus, under such homeostatic adaptation, neuronal populations maintain a constant stimulus-averaged firing rate in spite of changes to the environment (Benucci et al., 2013; Hengen et al., 2013; Maffei and Turrigiano, 2008).

Firing rate homeostasis via gain adaptation (cartoon examples).

(A) Gain adaptation after a shift in environmental statistics, without changes in tuning curve shapes. The bottom row shows the stimulus distributions in the first (left column) and the second(middle and right columns) environments. The top row shows the tuning curves of the same four neurons (main plots) and their mean firing rates (insets) when adapted to the first (left) and the second (right) environments, as well as immediately after the transition, but before adaptation to the new stimulus statistics (middle). Immediately after the transition the mean firing rates differ from their values in the first environment (middle and left insets). Neurons adapt to the new environment solely by independently adjusting their gains, i.e., by scaling their tuning curves up or down (correspondingly, the tuning curves of the same color in the middle and right columns — which belong to the same neuron — are scaled versions of each other and have identical shape), to a degree appropriate for returning their stimulus averaged firing rates to the level before the environmental shift (right and left insets). (B) Gain adaptation after a change in tuning curve shapes, due, e.g., to a change in the system’s computational goals. The plots have the same format as in (A). In this case, the stimulus distribution (bottom row) does not change between the two contexts, but the tuning curves have different shapes in the second context (top row, second and third columns vs. the first). Again, the average firing rates change immediately after the transition (middle vs. left insets), but the neurons adapt their gains to the right degree to return their rates to the level before the context changed (right inset).

There are multiple levels at which homeostatic adaptation can be observed. Firstly, there is population homeostasis, in which the stimulus-average firing rate of an entire population of neurons remains constant, without individual neurons necessarily holding their rates constant (Slomowitz et al., 2015; Sanzeni et al., 2023). Secondly, there is what we term cluster homeostasis. In this form of homeostasis, stimulus-average firing rate of groups or clusters of neurons with similar stimulus tuning remains stable under environmental shifts (Benucci et al., 2013), but the firing rate of individual neurons within a cluster can change. Lastly, in the strongest form, homeostasis can occur at the level of individual neurons, in which case the firing rate of each individual neuron is kept constant under changes in environment statistics (Marder and Prinz, 2003). Previous normative explanations for firing rate homeostasis often focus on the necessity of preventing the network from becoming hypo- or hyperactive (Turrigiano and Nelson, 2004; Maffei and Turrigiano, 2008; Keck et al., 2013; Hengen et al., 2013). Although this might explain homeostasis at the population level, it does not adequately explain homeostasis at a more fine-grained level. As we will show in this paper, the answer to the normative question that we posed above provides an explanation for firing rate homeostasis at such a finer level. Specifically, we will show that, for a wide variety of neural noise models (and under conditions that we argue do obtain in the visual cortex), optimal adjustment of neural gains to combat the effect of coding noise predicts firing rate homeostasis at the level of clusters of similarly-tuned neurons.

Having shown the optimality of homeostatic gain adaptation regardless of the computational goals of the population (which we have assumed determine its tuning curve shapes), we then apply our framework to Distributed Distributional Codes (DDC) (Vertes and Sahani, 2018), as a specific computational theory for neural implementation of Bayesian inference. Combining DDC theory (which determines tuning curve shapes) with our normative results (on optimal adjustments of neural gains) yields what we call homeostatic DDC. We will show that homeostatic DDCs explain the typical finding that sensory adaptations are stimulus specific (Kohn, 2007; Schwartz et al., 2007; Taaseh et al., 2011). In stimulus specific adaptation (SSA), the suppression of response is greater for test stimuli that are closer to an over-represented adaptor stimulus than for stimuli further away (to which the neuron may even respond more strongly after adaptation). In particular, this causes a repulsion of tuning curves away from the adaptor (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013). We will show that homeostatic DDC are able to account for SSA effects which cannot be fully accounted for in previous efficient coding frameworks (Wei and Stocker, 2015; Ganguli and Simoncelli, 2014; Snow et al., 2016), and we will use them to quantitatively model homeostatic SSA as observed in V1, by re-analysing previously published data (Benucci et al., 2013). This model relies on a special class of homeostatic DDC which we term Bayes-ratio coding. We show that Bayes-ratio coding has attractive computational properties: it can be propagated between populations without synaptic weight adjustments, and it can be achieved by divisive normalisation (which has been called a canonical neural operation) with adaptive weights (Carandini and Heeger, 2012; Westrick et al., 2016).

We start the next section by introducing our efficient coding framework for addressing the normative question posed above. We show numerically that our framework predicts that the firing rate of clusters of similarly tuned neurons should remain constant despite shifts in environmental stimulus statistics, with individual neurons free to shuffle their average rates. This result is shown to hold across a diverse set of noise models, demonstrating the robustness of the optimality of homeostasis to the precise nature of neural response noise. We additionally demonstrate that our framework can account for a wide distribution of stimulus-averaged single-neuron firing rates as observed in cortex. We then provide an analytic argument that demonstrates why, within a certain parameter regime, homeostasis arises from our optimisation problem, and show that cortical areas, in particular the primary visual cortex (V1), are likely to be within this parameter regime. We then numerically validate the quality of our analytic approximations to the optimal solution. Lastly, we apply our theory to DDC representational codes, showing how homeostatic DDC can account for stimulus specific adaptation effects observed experimentally.

Results

Theoretical framework

The central question we seek to address is how the presence of response noise affects optimal neural coding strategies. Specifically, we consider how a neural population should distribute activity in order to best mitigate the effect of neural noise on coding fidelity, subject to metabolic constraints, irrespective of the computations it performs on sensory inputs (which we do not optimise, but take as given). We consider a population of K neuronal units (We will use “units” instead of “neurons” to refer to the tuning curves because we will apply the theoretical framework to both cases where the units correspond to single neurons or clusters of similarly tuned neurons.), responding to the (possibly high-dimensional) stimulus s according to their tuning curves h(s). More precisely, we assume that our population engages in rate coding using time bins of a fixed duration, and denote the vector of joint population spike counts in a coding interval by n = (n1, …, nK). Single-trial neural responses, n, are taken to be noisy emissions from the neural tuning curves, h(s), according to a noise distributions n|s ~ Pnoise(n|s), subject to the constraint that the trial-averaged response 𝔼[n | s] equals h(s). We consider a number of different noise models, including correlated and power-law noise, showing that our main results are robust to specific assumptions about the nature of noise in neural systems. The sensory environment is characterised by a stimulus distribution P (s); accordingly, an “environmental shift” corresponds to a change in P (s), making certain stimuli more or less prevalent.

The central assumption of our efficient coding framework is that, while the shapes of neural tuning curves are determined by the computational goals of the circuit, the gains of those tuning curves are adapted purely for the purpose of fighting noise corruption at minimal metabolic cost, regardless of the computational goals. To this end, we adopt a shape-amplitude decomposition of the neural tuning curves. The tuning curve of the a-th unit, ha(s), is factorised into a representational curve, Ωa(s), and a gain, ga: Fig. 3A demonstrates the effect of changing the gain while keeping the representational curve, and therefore the tuning curve shape, fixed (Note that adjusting the gains does not mathematically restrict the computations which can be performed by downstream populations, as adjustments in gains can in principle — and if needed — be compensated for by reciprocal adjustments in readout weights. More formally, if a function, F, of the input s, can be computed from the unmodulated Ωa(s), say by a linear combination F (s) ≈ ∑ a waΩa(s), then it can equally be represented as a linear combination of the gain-modulated tuning curves, , with the adjusted weights .). The representational curve can be thought of as a normalised tuning curve; in fact, to make the definition of gain unambiguous, we define Ωa(s) to be the tuning curve normalised by its peak value, i.e., Ωa(s) ≡ ha(s)/ max ha (Fig. 3A inset). Correspondingly, the gain of a tuning curve is simply its maximum value, ga = maxs ha(s).

We take the shape of the tuning curve, captured by Ωa, to encode the computations represented by the neural population, and thus to be determined by those computational goals. These computational goals, and hence the Ωa may change across environments or contexts (e.g., corresponding to different tasks), but our efficient coding theory will not dictate how they should change (see Fig. 1B, and Section 2.7 for an example). Rather, we treat the representational curves, Ωa, as given, and do not optimise them within our efficient coding framework. In this respect, we deviate from the assumption of many classic studies based on the efficient coding principle, wherein information transmission is taken to be the sole computational goal of the system (usually located early in the sensory stream), and therefore the input-output transformation performed by the system is assumed to have been optimised purely for this purpose. General computations, however, often involve discarding information contained in the stimulus, that, e.g., constitute nuisance variables or may be irrelevant to behavioural tasks; this is the rationale for taking the Ωa as given and not optimising them purely based on information transmission criteria. In line with this rationale, we do not make any prior assumptions about or put any constraints on the shapes of the representational curves: the Ωa can be any complex (e.g., multi-modal, discontinuous, or heterogeneous) set of functions of the possibly high-dimensional stimulus, and can thus represent any computation. This makes our treatment more generally applicable than most other efficient coding frameworks which place tight constraints on the shapes and configuration of the tuning curves (see the Discussion for a more detailed comparison of our approach with those studies). In particular, this generality enables our theory to apply to populations located deep in the processing pathway, and not just to primary sensory neurons. For example, the representational curves, Ωa, could map the normalised response of face-detecting neurons in the inferotemporal cortex as a function of the stimulus, s, which could be taken as the high-dimensional raster of RGB values in an image shown to the animal.

Our framework, relating environmental stimuli to neural tuning curves and finally to noisy neural responses, and the respective computational roles of the gains and representational curves, is summarised by the diagram in Fig. 2.

The assumptions and the generative model underlying the theoretical framework.

The environment gives rise to a stimulus distribution P from which the stimulus s is drawn. Conceptually, the brain performs some computation on s, the result of which are represented by neurons via their normalised tuning curves, Ωa(s), which we thus refer to as representational curves. These are multiplied by adaptive gains, ga, to yields the actual tuning curves ha(s). The single-trial neural responses are noisy emissions based on ha(s). We assume the only function of gain modulation is to mitigate the eventual corruption of computational results by neural noise, subject to metabolic energy constraints.

We theorise that units adapt their gains to maximise an objective function that trades off the metabolic cost of neural activity with the information conveyed by the responses (Levy and Baxter, 1996; Ganguli and Simoncelli, 2014). Informally As is common in efficient coding frameworks (see e.g., Atick and Redlich (1990); Ganguli and Simoncelli (2014)) we assume the main contributor to metabolic cost is the energy cost of emitting action potentials. Thus we take the metabolic cost term in the objective to be proportional to , the total average spike count fired by the population. On the other hand, a natural and common choice (up to approximations) for the coding fidelity is the mutual information between the stimulus and response, I(n; s). However, since in general the analytic maximisation of mutual information is intractable, in the tradition of efficient coding theory (Brunel and Nadal, 1998; Ganguli and Simoncelli, 2014; Linsker, 1988), we will instead optimise an approximate surrogate for mutual information. We decompose the mutual information as I(s; n) = H[n] − H[n | s]. The marginal entropy term H[n] can be upper bounded by the entropy of a Gaussian with the same covariance, Substituting the right hand side for the marginal entropy term in the mutual information, gives us an upper bound for the latter. Using this upper bound on mutual information as our quantification of “coding fidelity”, yields the objective function where the constant µ > 0 controls the information-energy trade-off. This trade-off is illustrated in Fig. 3B. To summarise, we assume that the optimal gains maximise the objective Eq. (4)

The effect of changing the gain on tuning curves and different components of the objective function.

(A) As the gain, g, of a neuron is increased, the shape of its tuning curve remains the same, but all firing rates are scaled up. The inset shows the gain-independent representational curve, which we define as the tuning curve normalised to have a peak value of 1. (This cartoon shows a one-dimensional example, but our theory applies to general tuning curve shapes and joint configurations of population tuning curves, on high-dimensional stimulus spaces.) (B) Cartoon representation of the efficient coding objective function. Optimal neural gains maximise an objective function, ℒ, which is the weighted difference between mutual information, I, capturing coding fidelity, and metabolic cost, E, given by average population firing rate.

Note that because the Ωa depend deterministically on the stimulus s, and in turn fully determine the statistics of na, we have the mutual information identity I(n; s) = I(n; Ω) (this identity also holds for the upper bound on the mutual information that we are using). This means that the coding fidelity term in the objective function can equivalently be interpreted, without any reference to the low-level stimulus, as (an upper bound on) the mutual information between the population’s noisy output and its ideal, noise-free outputs, Ω, as specified by the computational goals of the circuit.

Optimal adaptation of gains leads to rate homeostasis across a variety of noise models

We now investigate the consequences of optimal gain adaptation, according to Eqs. (4)–(5), for the statistics of population response. As our hypothesis is that gains adapt in order to optimally mitigate the effect of neural noise on coding fidelity (with minimal energy expenditure), ideally we would like our conclusions to be robust to the model of noise. We thus consider a broad family of noise models which allow for general noise correlation structure, as well as different (sub- or super-Poissonian) scalings of noise strength with mean response. Specifically, we assume that conditioned on the stimulus, the noisy population response, n, has a normal distribution, with mean h(s), and stimulus-dependent covariance matrix where h(s) is the diagonal matrix with the components of the tuning curve vector, h(s), on its diagonal (more generally, we adopt the convention that the non-bold and index-free version of a bold symbol, representing a vector, denotes a diagonal matrix made from that vector), the noise strength σ scales the overall strength of noise, and 0 < α < 1 is a parameter that controls the power-law scaling of noise power with mean response h(s). Finally, Σ(s) is the noise correlation matrix which can depend on the stimulus, but is independent of the gains. The noise model Eq. (6) can equivalently be characterised by (1) the noise variance of spike count na is given by σ2ha(s)2α, and (2) the noise correlation coefficient between na and nb is given by Σab(s), independently of the neural gains. Thus the strength (standard deviation) of noise scales sublinearly, as gain to the power α. Note that the case σ = 2α = 1 corresponds to Poisson-like noise with variance of na equal to its mean; in App. B.2, we show that, under mild conditions, a true Poisson noise model (with no noise correlations) leads to the same results as we derive below for the Gaussian noise model, Eq. (6), with σ = 2α = 1 and Σ(s) = I.

In addition to the gains, the objective ℒ (g), Eq. (4), also depends tacitly on (i.e., is a functional of) the stimulus distribution P, the representational curves Ω, and the stimulus-dependent noise correlation matrix Σ. As shown in App. B.1, for our class of noise models, Eq. (6), ℒ depends on P, Ω, and Σ only through the following collection of signal and noise statistics. (By “signal” we refer to the noise-free trial-averaged population response, h(s); under variations of stimulus, h(s) can be thought of as a multivariate random variable with associated statistics.)

The units’ normalised average responses These are related to the units’ mean responses, ra, via ra := 𝔼[na] = 𝔼[ha(s)] = gaωa.

The relative size of stimulus-dependent signal variations, as quantified by the coefficients of variation of the units’ trial-average responses, CVa := SD[ha(s)]/𝔼[ha(s)].

Signal correlation structure, as quantified by the matrix, ρ, of correlation coefficients of the trial-averaged responses, ρab := Cov[ha(s), hb(s)] / (SD[ha(s)]SD[ha(s)]).

Noise correlation structure, as captured by the average normalised noise covariance matrix (Note that for Poisson-like noise scaling, α = 1/2, this matrix is actually a correlation matrix, in the sense that it is positive-definite with diagonal elements equal to 1.), W, defined by where denotes the tuning curve normalised by its stimulus average.

Note that all these statistics are invariant with respect to arbitrary rescalings of the tuning curves, and are thus independent of the gains — in particular, in all the definitions above, the ha could be replaced with the Ωa. In terms of these quantities, the objective function, Eq. (4), is given by and — in line with the convention introduced above — CV and r denote K × K diagonal matrices with CVa and ra = ωaga on their diagonals, respectively. Thus the objective depends on the gains, g, only via the average responses r. In addition, it also depends on the gain-independent quantities, CV, ρ, and W, which capture aspects of signal and noise statistics.

We now present the results of numerical optimisation of the objective Eq. (9) for different noise distributions within the general family of noise models, Eq. (6), for a population of K = 10, 000 units. To capture the notion of adaptation to changes in the environment or context, we simulated a family of environment models smoothly parameterised by a single parameter ν ∈ [0, 1]. As we have just seen, the objective depends only on ω, ρ, W, and CV. Changes in these quantities can arise due to changes in the stimulus density, P (s), the representational curves of the population, Ω(s), or both. We specify environment models only through the parameters on which the objective depends (ω, ρ, W, and CV), remaining agnostic to the underlying cause of changes to these quantities.

For simplicity, we fixed the response coefficient of variation CVa = 1, the noise scale σ2 = 1, and the information-energy trade-off parameter µ = 5/(1 − α) in all environments. (Below, in Sec. 2.5, we will discuss the biological basis for these parameter choices.) On the other hand, the normalised average response, ωa, and the signal correlation matrix, ρ, depended smoothly on the environment parameter ν, as follows. For each unit a = 1, …, K, the normalised average responses in the extreme environments, ωa(ν = 0) and ωa(ν = 1), were drawn independently from a wide distribution (specifically the beta distribution Beta(6, 1)). The normalised average response at intermediate values of ν, ωa(ν), was then obtained by linear interpolation between the independently sampled boundary values. This method for sampling environments is illustrated in Fig. 4. Similarly, to construct the signal correlation matrix ρ(ν), we first constructed the boundary values ρ(ν = 0) and ρ(ν = 1) with independent random structures and smoothly interpolated between them (see Sec. 4.1 for the details). We did this in such a way that, for all ν, the correlation matrices, ρ(ν), were derived from covariance matrices with a 1/n power-law eigenspectrum (i.e., the ranked eigenvalues of the covariance matrix fall off inversely with their rank), in line with the findings of Stringer et al. (2019) in the primary visual cortex. Finally, the stimulus-averaged noise correlation matrix, W, is determined by our choice of the noise model subfamily used in each simulation. We now present simulation results for three 1-parameter subfamilies of the general family of noise models, Eq. (6), in turn.

The structure of pre-modulated average population responses in the simulated environments.

We simulated a one-parameter family of environments parameterised by ν ∈ [0, 1]. In the environments on the two extremes, for each unit, the gain-independent average normalised responses ωa(ν = 0) and ωa(ν = 1) are randomly and independently sampled from a Beta(1, 6) distribution, with density shown on the left and right panels. For intermediate environments (i.e., for 0 < ν < 1), ωa(ν) is then obtained by linearly interpolating these values (middle panel). The middle panel shows K = 10000 such interpolations, with every 500th interpolation (ordered by initial value) coloured black.

Uncorrelated power-law noise

Consider first the case of no average noise correlations, in the sense that W = I (In terms of the objective function, Eq. (9), this is mathematically equivalent to having zero stimulus-conditioned noise correlations, i.e., Σ(s) = 1, but with CV’s redefined not to denote the standard coefficient of variation, but rather where ; for Poisson-like scaling, α = 1/2, this reduces to the standard coefficient of variation.). We investigated the adaptation behavior of optimally gain-modulated population responses as a function of the noise scaling exponent α. Fig. 5A shows the scaling of spike count variance as a function of mean spike count for different values of the noise scaling exponent α. By construction, in the absence of adaptation, an environmental shift (ν = 0 → ν = 1) results in a wide range of average firing rates (Fig. 5B, middle panel). But following adaptation, the firing rates become tightly concentrated around a single value (Fig. 5B, bottom panel), which matches the value they were concentrated around in the original environment (Fig. 5B, top panel). Thus, adaptation results in (1) uniformisation of average firing rates across the population in a fixed environment, and (2) firing rate homeostasis, where following adaptation the mean firing rates of the units returns to their value before the environmental shift. These same uniformisation and homeostatic effects are seen for other values of α (Fig. 5B and Supp. Fig. 14). In Fig. 5C we quantify deviation from perfect homeostasis as the average % difference in the units’ mean firing rates between the initial environment (ν = 0) and environments at increasing ν. As shown, the faster the spike count variance grows with mean spike count (i.e., the greater the noise scaling exponent α is) the greater the deviation from firing rate homeostasis. But even for super-poissonian noise with α = 0.75 the average relative change in firing rates between the environments never exceeds 3% (Fig. 5C).

Optimal gains under uncorrelated power-law noise lead to homeostasis of rates.

(A) Illustration of power-law noise. The trial-to-trial variance of spike counts, as a function of the trial-averaged spike count, is shown for different values of the noise scaling parameter α. (B) Distribution of average firing rates before and after a discrete environmental shift from ν = 0 to ν = 1 for the case α = 1/2. Top: the histogram of firing rates adapted to the original environment (ν = 0) before the shift, as determined by the optimal gains Eq. (5) in this environment, . Middle: the histogram of firing rates immediately following the environmental shift to ν = 1, but before gain adaptation; these are proportional to . Bottom: the histogram of firing rates after adaptation to the new environment, proportional to . We have normalised firing rates such that the pre-adaptation firing rate distribution has mean 1. (C) Deviation from firing rate homeostasis for different values of α as a function of environmental shift. For each environment (parametrised by ν), we compute the optimal gains and find the adapted firing rate under these gains. The average relative deviation of the (post-adaptation) firing rates from their value before the environmental shift, , is plotted as a function of ν.

Even under super-poissonian noise (α = 0.75 in Fig. 5C), the average relative change in firing rates between environments never gets above 3%. Fig. 5C additionally shows that the slower the spike count variance grows with mean spike count (i.e., the smaller the noise scaling exponent α is) the more precise the firing rate homeostasis.

Uniform noise correlations

In the second sub-family of noise models we allow for noise correlations, but consider only Poisson-like scaling of noise (i.e., a noise scaling exponent α = 1/2). More precisely, we assume that the effective noise correlation coefficients between all pairs of units (corresponding to off-diagonal elements of the matrix W, Eq. (8)) are the same, equal to p (Fig. 6A; The positivity of the noise correlation matrix, W, requires that . Assuming large K, the right side of the inequality goes to zero; hence we only simulated cases with positive p.). As shown in Fig. 6B, optimal gain adaptation leads to uniformisation and homeostasis of mean firing rates in this noise model as well. In fact, in the presence of uniform noise correlations, homeostasis is even stronger compared to the previous noise model with zero noise correlations; in this case, mean relative deviations in average firing rates never exceed 1% (Fig. 6C). Further, as noise correlations get stronger (i.e., as p increases) we see tighter homeostasis (see also Supp. Fig. 15).

Firing rate homeostasis under Poissonian noise with uniform noise correlations.

(A) Illustration of an effective noise correlation matrix with uniform correlation coefficients; the (green) off-diagonal elements have the same value, p < 1, while the (black) diagonal elements are 1. Panels B and C have the same format as in Fig. 5.

Aligned correlated noise

In the final noise model we considered, we once again fixed α = 1/2 (Poisson-like scaling), but allowed for heterogeneous effective noise correlation coefficients. Specifically, we took the effective noise correlation matrix W to have approximately the same eigenbasis as ρ, the signal correlation matrix — hence “aligned noise”—, but with a different eigenvalue spectrum. As described above, for all noise sub-families, the signal correlation structure ρ(ν) was obtained by normalising a covariance matrix with 1/n spectrum (corresponding to the findings of Stringer et al. (2019) in V1); in the current case, we obtained the noise correlation structure W (ν) by normalising a covariance matrix with the same eigenbasis but with a 1/nγ spectrum (i.e., with the n-th largest eigenvalue of W scaling as 1/nγ); see Fig. 7A, and Sec. 4.1 for further details. Fig. 7B demonstrates that, as with the other noise sub-families, optimal gain adaptation in the presence of aligned correlated noise also leads to homeostasis and uniformisation of mean firing rates (see Supp. Fig. 16 for rate histograms for other values of γ). Across different values of γ, homeostasis is again tighter compared to the zero noise correlation case, with the average firing rate shift never exceeding 1% (Fig. 5C). Further, with the exception of γ = 1, we see more homeostasis for higher values of γ, corresponding to lower dimensional noise. The special case of γ = 1 corresponds to so-called information-limiting noise (Moreno-Bote et al., 2014; Kanitscheider et al., 2015), where noise and signal correlation structures are fully aligned: W = ρ. In this case, homeostasis is perfect, and the mean relative error in firing rates is 0. We provide an analytic proof that in this case we have perfect homeostasis and uniformisation in App. B.5, and discuss this situation further in Sec. 2.5 and Sec. 2.6.

Firing rate homeostasis under signal-aligned noise.

(A) Eigenspectra of noise covariance matrix for the aligned noise model for various values of γ; in this model, the n-th eigenvalue of the average noise correlation matrix, W, is proportional to 1/nγ. Panels B and C have the same format as in Fig. 5.

We have shown that, according to our efficient coding framework, optimal adaptation of response gains (with the goal of combating neural noise at minimal metabolic cost) can robustly account for the firing rate homeostasis of individual units within a population, despite environmental shifts or changes in neural representations (tuning curve shapes), under a diverse family of noise models. To shed light on these results, in Sec. 2.4, we analytically investigate sufficient conditions on the parameters, Eq. (7)–(8), of our model under which we expect homeostasis to emerge. Following that, in Sec. 2.5, we will show that biological estimates for these parameters obtained in cortex (which are determined by natural stimulus statistics, as well as the stimulus dependence of high-dimensional population responses) are consistent with these conditions.

A clustered population accounts for the diversity of firing rates seen in cortex

In addition to homeostasis, the above simulation results also predict uniformisation of average firing rates across the population, in any given environment. By contrast, cortical neurons display a wide range of average firing rates spanning a few orders of magnitude (Buzsáki and Mizuseki, 2014). But a cortical neuron can have very similar response tuning to other neurons in the same area, and neurons can therefore be grouped based on the similarity of their tuning curve shapes. We can thus interpret the units in our preceding numerical experiments not as single neurons, but as clusters of similarly tuned neurons. In this interpretation, uniformisation occurs at the cluster level, with potentially a wide distribution of single-cell average firing rates within each cluster. We found that our model is indeed consistent with such an interpretation. In particular, in the case of uncorrelated noise with Poisson-like mean-variance scaling, we show that a simple extension of our model gives rise to both diversity of firing rates as seen in cortex, as well as homeostasis and uniformisation at the level of clusters of similarly tuned neurons.

To gain intuition and allow for analytic solutions, we study a toy model in which the N neurons in the population are sorted into K such clusters. In this toy model, neurons within a cluster have very similar stimulus tuning, but neurons in different clusters are tuned differently (Fig. 8A); the model thus exaggerates a more realistic situation in which there is a more gradual transition from similar to dissimilar tuning curves within the entire population. A perturbation strength parameter, ϵ ≥ 0, controls how much the tuning curve shapes of single neurons in a cluster deviate from the stereotype tuning curve shape for that cluster (see Eq. (25) in Methods). When ϵ = 0, we have perfect clusters, i.e., neurons in a cluster have identical representational curves. As the perturbation strength ϵ increases, within-cluster deviations become stronger.

Optimal gain adaptation in a clustered population accounts for the broad distribution of cortical single-neuron firing rates.

(A) A cartoon description of the clustered neural population model. Top: The single neurons in the population belong to different functional groups or clusters (shown by different colors). Bottom: The normalised tuning curves of the single neurons belonging to the same cluster (thin lines of the same color) are very similar, and close to their cluster’s stereotype normalised tuning curve (solid lines) which can be very different across clusters. (B) Each panel shows the distribution of log firing rates of the active single neurons for different choices of cluster size k and effective dimensionality, D, of the within-cluster tuning curve variations. The distributions shown are for the active neurons with non-zero firing rate, with their percentage of the total population given in each panel (purple text). For all simulations we fixed the cluster firing rate to k × 5 Hz (corresponding to µ = 10, if we assume a coding interval of 0.1 s), such that the mean rate of single neurons is 5 Hz on average.

We assume single-neuron gains adapt to optimise the same efficient coding objective defined above in Eq. (9), but with units now clearly interpreted as single neurons. Note that optimisation in terms of the neural gains is equivalent to optimisation in terms of the neurons’ mean firing rates (because the objective depends on the neurons’ gains only via their mean fire rates each neuron’s mean rate is proportional to its gain by a fixed constant of proportionality, given by the mean value of its fixed and unoptimised representational curve, Eq. (7)). For ϵ = 0 (identical tuning curve shapes within each cluster), the efficient coding objective function is blind to the precise distribution of single-cell firing rates among the neurons of a cluster, as long as the total rate of the cluster is held fixed. This is because for clusters of identically-tuned and independently-firing neurons with Poisson-like noise, the cluster can be considered as a single coherent unit whose firing rate is given by the sum of the firing rates of its neurons. Thus at zeroth order, the distribution of individual neuron rates are completely free and undetermined by the optimisation, as long as they give rise to optimal total cluster rates.

For small but non-zero perturbation strength ϵ (corresponding to non-identical but sufficiently similar within-cluster tuning curve shapes), we showed that the efficient coding optimisation breaks down into two optimisation problems: one at the level of clusters, wherein the mean cluster rates are optimised, and one at the level of individual neurons within each cluster (Methods Sec. 4.2 and App. B.3). The cluster-level problem remains decoupled from the within-cluster distribution of single-neuron rates; thus the optimisation of cluster firing rates reduces to the same optimisation problem considered in the previous sections, Eqs. (4)–(5), with the units now interpreted not as neurons but as clusters of similarly tuned neurons. On the other hand, within-cluster optimisation of individual neuron rates decouples across clusters. Within each cluster, we show that this optimisation is approximately given by a quadratic program, subject to the constraint that the individual neuron rates sum up to the optimal total cluster rate found by solving the cluster-level problem (Eqs. (28)–(30) in Sec. 4.2).

The within-cluster quadratic program optimisation is specified by the signal covariance structure of the normalised tuning curves in the cluster, or more precisely, the covariance matrix of their deviations from the cluster stereotype normalised tuning curve (see Eq. (30)). A key characteristic of any covariance matrix is its effective dimensionality (corresponding roughly to the number of principal components with significant variance). Instead of making arbitrary choices about the signal covariance matrix defining the within-cluster optimisation, in our simulations we sampled them from a simple random ensemble which was controlled only by an effective dimensionality parameter, D, and the cluster size, k (see Sec. 4.3 for details). Figure 8 (see also Supp. Fig. 18) shows the resulting firing rate distributions for neurons aggregated across all clusters (corresponding to optimisations with independently sampled covariance matrices), for different values of k and D. Firstly, note that, depending on these parameters, a significant fraction of neurons can be silent, i.e., with zero firing rate in the optimal solution; the figure shows the percentage of and the rate distributions for the active neurons. Our theory therefore predicts that a significant fraction of cortical neurons are silent in any given sensory environment, and that which neurons are silent can shuffle following shifts in environmental stimulus statistics (specifically shifts that result in significant changes in the signal correlation structure of the cluster neurons). Secondly, note that the distributions span multiple orders of magnitude, in agreement with empirical observations in the cortex (see e.g. Slomowitz et al. (2015); Hengen et al. (2013); Maffei and Turrigiano (2008)), and is approximately log-normal (albeit with a skewed tail towards small log-rates), also consistent with empirical findings (Buzsáki and Mizuseki, 2014). Lastly, for fixed k, as the effective tuning curve dimension, D, increases, the fraction of silent neurons decreases (Supp. Fig. 18).

To summarise, we have shown that the objective ℒ considered in Sec. 2.2, reinterpreted to correspond to the cluster level, can arise from optimisation of a similar objective, defined at the level of single neurons in a functionally clustered population (Eqs. (26)–(27)). Furthermore, the correction to the cluster-level objective (arising from within-cluster variations in neural tuning) serves to break the symmetry within clusters, generating a diverse range of firing rates which matches the distribution of cortical single-neuron firing rates. In the rest of the paper, we will pursue the problem at the coarser level of clusters.

Analytical insight on the optimality of homeostasis

The numerical results of Sec. 2.2 showed that, across a range of neural noise models, optimal gain modulation for combating noise can give rise to firing rate homeostasis and uniformisation across neural clusters defined based on similarity of stimulus tuning. To shed light on these results, we analytically investigated how and in what parameter regimes these results arise. We found that homeostasis and uniformisation emerge in the regime where the matrix is small (according to an appropriate matrix norm); here β = 2 − 2α. The intuitive meaning of this matrix is that its inverse, Δ−1, characterises the strength of the signal-to-noise ratio (SNR) along different directions in the code space, i.e., the space of cluster responses. Therefore the above condition can be roughly interpreted as the requirement that signal-to-noise ratio is strong (This is a self-consistent statement, as Δ−1, more precisely, encodes the high-dimensional SNR structure (which in general depends on the gains) under the approximate homeostatic solution Eq. (11). But since largeness of Δ−1 justifies that approximation, the largeness of SNR in a population with the homeostatic gains Eq. (11) self-consistently justifies that homeostatic approximation.). We will discuss this condition, and the constraints it imposes on various model parameters, further in Sec. 2.5, and will provide evidence that it is likely to hold in cortex. Here, we focus on its consequences.

We developed a perturbative expansion in Δ, to obtain approximate analytical solutions for the cluster gains that maximise our objective ℒ, Eq. (9) (see App. B.4). In the leading approximation, i.e., to zeroth order in Δ, this expansion yields We denote this approximate solution by . To see the significance of this result, note that the average response of cluster a, i.e., its stimulus-averaged total spike count is given by 𝔼[ha(s)] = gaωa. Thus, in the leading approximation, the stimulus-average spike count of cluster a is given by the constant . Thus, since βµ is constant across both clusters and environments, the zeroth order solution yields homeostasis and uniformisation of cluster firing rates.

In Sec. 2.6, we numerically quantify the performance of the homeostatic approximation, g(0), in terms of maximising the efficient coding objective, under different noise models. Here, we also note the following exact analytical result regarding homeostasis that holds even for non-small Δ. It follows form the structure of the objective function, Eq. (9), that, irrespective of how pre-modulation mean rates, ωa, change between two stimulus environments, as long as Δ and W, the effective noise covariance, remain invariant across the environments, optimal gains imply exact homeostasis of each cluster’s average firing rate across the two environments. However, even in this case, unless Δ is small, in general the optimal solution does not result in uniformisation of firing rates across clusters.

Conditions for the validity of the homeostatic solution hold in cortex

We asked whether cortical populations satisfy the conditions under which the homeostatic solution, Eq. (11), is expected to provide a good approximation to optimal gains according to efficient coding. As noted above, the general mathematical condition is that the matrix Δ, defined in Eq. (10), is in some sense small (recall that through its inverse, Δ represents the high-dimensional structure of signal-to-noise ratio in the coding space, under the homeostatic solution Eq. (11)). We can make this statement more precise by looking at the size of the leading-order correction to the approximate homeostatic solution Eq. (11), in the perturbative expansion. We found that the first-order correction modifies the approximate homeostatic solution for the gain ga by a multiplicative factor 1 − Δaa (see Eq. (106) in App. B.4). In other words, to leading order, the relative error in the homeostatic approximation to the optimal gain (and thus the optimal mean response) for unit a is given by |Δaa|; we can thus reformulate the condition for the accuracy of the homeostatic approximation as for all a. Conversely, in the first approximation, deviations from homeostasis depend on the SNR structure, as captured by Δaa, and how that structure shifts between environments.

Before proceeding to empirical estimates of Δaa, let us first delineate the dependence of Δ on different model parameters. Firstly, note that Δ scales linearly with the noise strength σ2; thus, if noise strength is scaled up, so is Δ. Next, consider increasing the noise scaling exponent α. This decreases (βµ)β provided α 1 ≤ − e/2µ (which holds in our simulations above). Moreover, increasing α increases the diagonal elements of W (see Eq. (8)). Intuitively, this increases the magnitude of W. Through these two mechanisms, increasing the noise scaling exponent α will generally increase the size of Δ. Next, note that Δ has an inverse relationship with the coefficients of variation, CV, which increase as neurons become more stimulus selective; accordingly, the more selective the tuning curves, the smaller the Δ. Lastly, Δ has a complex dependency on the spectra of both CVρ CV and the noise correlation structure W as well as the alignment of their eigenbases (which represent the geometric structure of signal and noise in the neural coding space, respectively). In Sec. 2.6, we will numerically investigate these dependencies, and their effects on the validity of the homeostatic approximation.

Translated to biological quantities, it follows from the above dependencies that the smallness of Δ (or Δaa) requires that (1) the cluster firing rates are sufficiently high, (2) noise variance scales sufficiently slowly with mean response, (3) individual clusters are sufficiently selective in responding to stimuli, and (4) the neural representation has a high-dimensional geometry in the population response space. We used the available biological data to estimate the size of these variables for cortical populations; see Sec. 4.4. Here we summarise the resulting estimates.

Sufficiently high firing rate

In the homeostatic approximation, Eq. (11), βµ is equal to the average total spike count of a cluster over the rate-coding time interval. Condition Eq. (12) thus requires that the stimulus-averaged firing rate of all clusters are sufficiently high. Based on the available data (see Sec. 4.4), we estimate the mean firing rate in rodent V1 to be around 5 Hz. Assuming a coding interval of 0.1 seconds, and a cluster size of k = 20, this yields βµ ~ 10.

Sufficiently slow scaling of noise variance

We can obtain an estimate for the prefactor σ2/(βµ)β, by fitting the following mean-variance relationship for cluster spike counts: variance = σ2mean2−β. Empirical estimates of this relationship lead to estimates of σ2/(βµ)β ranging from 0.05 (Shadlen and Newsome, 1998) to 0.31 (Koyama, 2015), with a median value 0.14 (see Sec. 4.4 for details). This is close to the value for Poissonian noise (σ2 = 1, β = 1), which is 0.1.

Sufficiently selective responses

The coefficient of variation CV for neural responses is a measure of neural selectivity. Accordingly, it is related to another measure of stimulus selectivity, namely the coding level (i.e., the fraction of stimuli to which a neuron responds). Using empirical estimates of the latter quantity for cortical neurons (Lennie, 2003), we estimated CV2 ≈ 20. Below we take CV2 ≈ 10 as a more conservative estimate.

High-dimensional signal geometry

Assuming, for simplicity, that the coefficients of variation are approximately the same for all clusters, we see from Eq. (12) that Δaa are proportional to the diagonal elements of the matrix ρ−1W. Assuming further, that these elements are all comparable in size (This is expected to be valid when the eigenbasis of ρ−1W is not aligned with the standard (cluster) basis, which, in turn, corresponds to distributed coding.), we estimated their order of magnitude based on the known structure of the signal correlation matrix, ρ, in V1 (see below), and for the aligned noise family of average noise covariance matrices, W, studied in Sec. 2.2. This family includes two biologically relevant cases: (1) an uncorrelated noise model corresponding to W = I, and (2) a W with so-called information-limiting noise correlations (Moreno-Bote et al., 2014; Kanitscheider et al., 2015). In the first case Δaa is proportional to [ρ−1]aa. It can be shown that [ρ−1]aa ≥ 1, with equality if and only if cluster a has zero signal signal correlation with every other cluster. [ρ−1]aa can thus be seen as a measure of representational redundancy. Many traditional efficient coding accounts predict zero signal correlation between neurons, at least in the low noise limit (Barlow, 2012; Atick and Redlich, 1990; Nadal and Parga, 1994, 1999), providing a normative justification for low signal correlations. However, complete lack of signal correlations is not necessary for [ρ−1]aa to be small. Geometrically, small signal correlations correspond to neural responses forming a representation with high intrinsic dimensionality. Stringer et al. (Stringer et al., 2019) found that the signal covariance matrix of mouse V1 neurons responding to natural stimuli possesses an approximately 1/n eigen-spectrum. In this case, for K large, we obtain the estimate [ρ−1]aa ≈ ln(K)/2 (see App. B.6). Given the slow logarithmic increase with K, we expect that our condition can hold even for very large neural populations. For example, suppose we take the entire human V1 as our neural population, to obtain an arguably generous upper bound for K. This contains roughly 1.5 × 108 neurons (Wandell, 1995), leading to 7.5 × 106 clusters of 20 neurons, and therefore an average value of [ρ−1]aa just under 8. This, together with our other estimates above, yields the following overall estimate for Δaa We found that in the case of information-limiting noise correlations the |Δaa| tend to be even smaller than the above estimate. Information-limiting noise correlations refer to the case where directions of strong noise in the coding space align with those of strong signal variation. The extreme case, i.e., perfect alignment of noise and signal geometry, corresponds to ρ = W, or equivalently ρ−1W = I, for which Δaa ∝ [ρ−1W]aa = 1 does not grow with K at all. In App. B.6 we analysed less extreme forms of alignment (corresponding to the model numerically studied in Fig. 7, Sec. 2.2), and found (unsurprisingly) that for those cases, the estimates of Δaa lie between the estimates for the two above cases. Accordingly, even partial signal-noise alignment can (potentially strongly) reduce the size of Δ below the estimate Eq. (13), further encouraging homeostatic coding.

The above analysis makes it clear when we should expect homeostasis in general – when cluster firing rates are high enough, the noise scaling of cluster responses is sufficiently slow, responses are highly selective, and signal correlation structure corresponds to a high-dimensional geometry (Stringer et al., 2019), possibly with information-limiting noise correlations (Moreno-Bote et al., 2014; Kanitscheider et al., 2015). On the other hand, when these conditions are violated, the optimal gain configuration can potentially deviate strongly from the homeostatic solution. Moreover, from the above estimates, leading to Eq. (13), we see that in V1 and possibly other cortical areas, we can expect the leading corrections to the homeostatic solution to not exceed 10%, in relative terms.

Numerical comparison of homeostatic gains to optimal gains

We now return to the numerical simulations performed in Sec. 2.2, to evaluate the performance and accuracy of the approximate homeostatic solution, , Eq. (11), and the first-order correction to it, , relative to the true optimal solution of Eq. (9). We also evaluate the performance and accuracy of the best possible homeostatic solution, i.e., the best configuration of cluster gains, , that, by construction, are guaranteed to yield homeostasis and uniformisation. In other words, under such gains the mean responses ra = gaωa must equal a constant (i.e., independent of a, the cluster index, and constant across environments). Denoting this a priori unknown constant by χ, we thus have under which the average spike count of clusters in the coding interval is given by χ. The variable χ is then chosen to optimise the expectation of ℒ, Eq. (9), across a relevant family of environments (note that we need to optimise the average objective across environments, since otherwise the optimal χ will in general depend on the environment, invalidating homeostasis). With such an optimal χ, ghom will, by definition, perform better than g(0) in terms of our objective function, ℒ (Further, it can be shown (see App. B.7) that χ < βµ, i.e., the mean cluster firing rate under the optimal homeostatic solution is strictly smaller than that predicted by g(0), and that the average spike-count of the entire population of clusters is approximately the same under this solution as it is under the first-order approximation g(1).). In fact, as we will show below, in our simulations (with model parameters within the biologically plausible ranges identified in the previous section), the homeostatic ghom performed very close to the true optimal gains.

For each of the different noise models considered in Sec. 2.2 we compare the performance of the homeostatic gains g(0) and ghom, as well as the perturbative correction, g(1), to the numerically optimised gains. To examine the adaptation properties of these gains, we once again work with the sequence of environments indexed by ν ∈ [0, 1], with the same specifications as in the numerical simulations of Sec. 2.2. To measure the accuracy of the approximations to the optimal gains, for each environment ν, we calculated the mean relative errors for each of the approximate solutions, denoted by gapp(ν), with gopt(ν), the numerically optimised solution in environment ν. To measure the performance of an approximate solution, we use the following relative improvement measure, defined by which can be interpreted as the improvement in the objective ℒ (·; ν) (the efficient coding objective Eq. (9) according to the statistics of environment ν) achieved by the adaptive gains gapp(ν) over the unadapted gains from the original environment gapp(0), relative to the improvement obtained by the optimally adaptive gains gopt(ν). The better the approximation the closer Capp(ν) will be to 1. Note that, in general, the relative improvement worsens when ν goes to 0 (see Fig. 9B, D and F); this follows from the definition of C(ν) and the fact that when environmental change is weak, using gains optimised in the original environment without adapting them in the new environment is superior to adaptive but only approximately optimal gains. We now summarise the results for the different noise models discussed in Sec. 2.2 (see that section for the detailed specification of the three noise models).

Accuracy of homeostatic approximation to optimal gains.

(A) The relative error, Eq. (15), of different approximate solutions for the optimal gains, averaged across environments (the standard deviations of these values across environments were negligible relative to the average), as a function of the noise scaling parameter α. To give a sense of the scale of variation of the optimal gains across environments, the black line shows the relative change in the optimal gains between the most extreme environments (a measure of effect size), . (B) Relative improvement in the objective for the homeostatic approximation ghom as measured by Eq. (16). Panels C and E (panels F and D) are the same as panel A (panel B), but for the uniformly correlated noise and the aligned noise models, respectively.

Uncorrelated power-law noise

First, we consider the uncorrelated noise model with general power-law scaling of noise strength with mean responses. In this case, the homeostatic solution ghom, and the first order correction, g(1), had very low (< 3%) relative errors for all noise scaling exponents, α, while the relative error in g(0) becomes large for significantly super-Poissonian noise scaling, i.e., α = 0.75 (Fig. 9A; For the case of uncorrelated power-law noise model, as well as the aligned correlated noise model, we used an analytical approximation to the optimal χ (see App. B.7) used in the homeostatic approximation , Eq. (14). This means that the performance measures in the plots of Fig. 9B and F are lower-bounds for the performance of with the truly optimal χ.). The relative improvement measure Chom(ν) for the homeostatic solution is also close to one, the optimal value, for all α, except for 0.75 (Fig. 9B). For the other approximate solutions too (Supp. Fig. 17A,B), the relative improvement decreases with increasing noise scaling. This is consistent with our analysis, since increasing α increases the size of Δ (see Sec. 2.4) and hence the influence of terms of higher order in Δ.

Uniform noise correlation coefficients

We now examine the effect of noise correlations on the accuracy of the approximate homeostatic solutions, first within the noise family with uniform noise correlation coefficients, denoted by p, and Poisson-like scaling (α = 1/2). For all p, the relative deviation of the homeostatic solution ghom from optimal gains was below 1% (Fig. 9C), while g(0) and g(1) were less accurate overall. Note that the accuracy of the homeostatic solutions ghom and g(0) increases with p, suggesting that positive noise correlations improve the optimality of homeostatic gains. The same trend can be seen in the relative improvement on the objective, Chom, for homeostatic gain adaptation, which increases with p (Fig. 9D).

Aligned noise

Lastly, we considered the algined noise model, in which noise has Poisson-like scaling (i.e., α = 1/2), and is correlated across clusters with the noise and signal covariance matrices sharing the same eigenbasis. In this noise family, the parameter, γ, controls the scaling of the eigenvalues of the noise covariance matrix with their rank, n, as 1/nγ. As in the previous cases, for all γ, the homeostatic solution ghom was very close to the optimal gains gopt (Fig. 9E), and preformed nearly equally (Fig. 9F). The case γ = 1, which corresponds to perfect alignment between signal and noise correlations, W = ρ, is particularly interesting. As noted in the previous section, this case corresponds to that of so-called information-limiting noise correlations (Moreno-Bote et al., 2014; Kanitscheider et al., 2015). As we show in App. B.5, in this case the optimal gains, without any approximation, lead to homeostasis and uniformisation of cluster firing rates. Thus, in this case gopt = ghom exactly, leading to zero relative error (Fig. 9E) for ghom, and perfect relative improvement Chom (Fig. 9F). For other cases, we can see that for increasing γ (at least for the values we simulated), which corresponds to more correlated and lower-dimensional noise (as noise power falls off faster as a function of eigenvalue rank), the homeostatic gains ghom tend to perform better.

Our numerical simulations across the three noise subfamilies (with varying noise power scaling, noise correlations, and noise spectrum) demonstrate that homeostatic strategies can - robustly and across a wide range of noise distributions - combat coding noise by effectively navigating the trade-off between energetic costs and coding fidelity.

Finally, note that adaptation in ghom uses only information local to each neuron, i.e., its firing rate. This implies that good performance does not require the use of complex regulation mechanisms which take into account how the encoding of each cluster relates to the population at large; homeostatic regulation, implemented by purely local mechanisms, is sufficient. In fact, homeostatic regulation can be implemented via a simple mechanism of synaptic scaling (Turrigiano et al., 1998; Turrigiano, 2008). Synaptic scaling involves the normalisation of synaptic weights onto a neuron in order to keep the total synaptic mass constant. In Appendix B.8, we demonstrate that, in a linear feedforward network, synaptic normalisation, via synaptic scaling, is both necessary and sufficient to allow for the propagation of homeostatic coding from the first layer to the subsequent ones. Thus, homeostatic coding in feedforward networks provides an additional normative interpretation of synaptic normalisation.

Homeostatic noisy DDCs

When exposed to over- or under-represented stimuli, neurons across sensory systems exhibit stimulus-specific adaptation whereby they decrease or increase their gains only for a limited subset of stimuli (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Nelken, 2014). In particular, homeostatic adaptation in the primary visual cortex (V1) has been observed to be stimulus-specific (Benucci et al., 2013). In this section and the next (Sec. 2.8), we combine our efficient coding theory of homeostatic adaptation with a specific computational theory of neural representations. Then, in Sec. 2.9, we show that the resulting theory is able to account for stimulus-specific adaptation effects observed in V1, which have been difficult to account for under other efficient coding frameworks (see our discussion of previous attempts in Sec. 3). This theory thus provides a unified normative explanation for these two types of adaptation.

Recall that up to this point we have made no assumptions about the nature of cortical representations (beyond rate coding, and the condition described by Eq. (12)), and thus our framework has been independent of the computational goals of the circuit. We now apply our framework to a specific theory of neural representation and computation, namely the distributive distributional code (DDC) (Sahani and Dayan, 2003; Vertes and Sahani, 2018), introducing what we term a homeostatic DDC.

While there is ample behavioural and psychophysical evidence for the hypothesis that animal perception and decision making approximate optimal Bayesian probabilistic inference (Kersten et al., 2004; Stocker and Simoncelli, 2006; van Bergen et al., 2015; Geurts et al., 2022), our knowledge of the neural implementation of computations serving Bayesian inference (such as computations of posterior distributions and posterior expectations) are rudimentary at best. The DDC framework is one proposal for neural implementations of Bayesian computations underlying perception. In a DDC model, neural responses directly encode posterior expectations of the inferred latent causes of sensory inputs according to an internal generative model. This internal model mathematically relates the latent causes or latent variables, which we denote by z, to the sensory inputs or stimuli, s, via a family of conditional distributions, f (s | z), as well as a prior distribution, π(z), over the latent variables. The sensory system is assumed to implement a so-called recognition model corresponding to this generative model. Namely, given a stimulus s, the task of the sensory system is to invert the generative process by calculating and representing the posterior distribution of latent variables given the current sensory input (see the schema in Fig. 10).

Schema of the generative and recognition models underlying a DDC.

According to the internal generative model, sensory inputs, s, are caused by latent causes, z, that occur in the environment according to a prior distribution π(z). A conditional distribution f (s | z), the so-called likelihood function, describes how the latent causes generate or give rise to the sensory inputs s. The task of the brain is to invert this generative process by inferring the latent causes based on the current sensory input, which is done by computing the posterior distribution Π(z|s).

In a DDC, neural responses are specifically postulated to directly represent the posterior expectations of the latent variables, z, or rather, the posterior expectations of a sufficiently rich (and fixed) class of so-called kernel functions of z. Specifically, in the standard DDC, each neuron, say neuron a, is assigned to a so-called kernel function of the latent variables, ϕa(z), and its response is given by the posterior expectation 𝔼[ϕa(z) | s]. Here, we modify the standard DDC scheme in two ways. First, we assume single-trial neuron responses are corrupted by noise and represent the posterior expectations of the kernel functions only on average. Second, we adopt the key premise of our general efficient coding framework, namely that neural gains adapt in order to optimally mitigate the effect of neural noise on representational fidelity. Specifically, we assume this gain adaptation takes the homeostatic form, Eq. (14) (which we found approximates the optimal solution of the efficient coding problem for noise mitigation). These modifications lead to a new coding scheme which we call noisy homeostatic DDC. In this scheme, as in the standard DDC, the tuning curve of a neuron, with assigned kernel function ϕa(z), is proportional to the posterior expectation 𝔼[ϕa(z) | s]. However, unlike the standard DDC, the corresponding constants of proportionality are not absolute constants but adapt (independently for different neurons) in such a way as to give rise to firing rate homeostasis. In other words, we assume the representational curve of neuron a is proportional to 𝔼[ϕa(z) | s], while its gain adapts across contexts or environments so as to maintain homeostasis. Since (under an ideal-observer internal model) the posterior expectation of a quantity averaged over stimuli is equal to that quantity’s prior expectation, we have 𝔼[𝔼[ϕa(z) | s]] = 𝔼[ϕa(z)]. Therefore, keeping average firing rates constant at level χ (independent of the kernel function or the environmental statistics) requires the gains to be inversely proportional to the prior expectations, 𝔼[ϕa(z)], and thus the tuning curve of neuron a to be given by In other words, in noisy homeostatic DDCs, the neural tuning curves are (up to the absolute constant χ) the ratio of the posterior to the prior expectations of the neurons’ assigned kernel functions. Note that the prior expectation in the denominator, which by definition does not depend on the stimulus, s, does depend implicitly on the environmental stimulus statistics. Indeed, this is also true of the posterior expectation in the numerator (as it combines information about the current stimulus with prior expectations about the latent causes of the stimulus input which can depend on the environment). Thus, in homeostatic DDCs, not only the gains, but also the shapes of tuning curves can adapt to changes in stimulus statistics, and thus vary across environments. This corresponds to a mixture of the two special scenarios sketched in the two panels of Fig. 1: as the stimulus distribution changes (Fig. 1A second row), (1) the tuning curve shapes change (Fig. 1B top row), as dictated by the computational goal of representing posterior expectations of latent variables under the changed statistics, and (2) the gains adapt in order to mitigate the effect of noise at minimal metabolic cost, resulting in mean response homeostasis (right columns). As we will see in Sec. 2.9, this combined adaptation can explain stimulus-specific adaptation.

Homeostatic DDCs can be implemented by divisive normalisation

We now show that for the special case when the DDC kernel function of neuron a is approximately a delta function, ϕa(z) ≈ δ(z − za), the homeostatic DDC can be implemented by divisive normalisation with adaptive weights (Westrick et al., 2016). For ϕa(z) ≈ δ(z − za), Eq. (17) becomes where in the last equality we used Bayes’ rule: Π(z | s) = f (s | z)π(z)/P (s). Thus, in this case, the tuning curve of neuron a is the ratio of the posterior distribution to the prior distribution both evaluated at that neuron’s preferred latent variable value, za. Accordingly, we call this coding scheme Bayes-ratio coding.

Divisive normalisation has been dubbed a canonical neural operation (Carandini and Heeger, 2012), as it appears in multiple brain regions serving different computational goals. Given the feedforward inputs, Fa(s), the divisive normalisation model computes the tuning curve of neuron a as Here, wb are a collection of so-called normalisation weights, and α ≥ 0 and n ≥ 1 are constants. To show that Bayes-ratio coding can be implemented by a divisive normalisation model, we note that P (s), which appears in Eq. (18), is given by the marginalisation of the joint density P (s) = ∫ P (s, z)dz = ∫ f (s|z)π(z)dz. We approximate the integral by a sum as P (s) ≈ ∑b f (s|zb)π(zb)Δzb =b wbf (s|zb), where we defined the weights by wb = π(zb)δzb (We have assumed the latent variable space is partitioned into cells, one for each neuron, with δzb denoting the volume of the cell centred on neuron b’s preferred latent variable zb.). Substituting this approximation in Eq. (18), and regularising the denominator by an additive small constant α > 0, we obtain By Eq. (19) this corresponds to a divisive normalisation model with n = 1. Not only does this result show that implementing Bayes-ratio coding is biologically plausible, it also provides a normative computational interpretation to both the feedforward inputs and the normalisation weights of the divisive normalisation model, as the internal generative model’s likelihoods and prior probabilities, respectively. Note also that the normalisation weights of this model are adaptive, and vary between environments that have different latent variable distributions, π(z).

Finally, we consider the propagation of Bayes-ratio coding between populations. Suppose the generative model has the hierarchical structure z(2) z(1) → s, and consider a feedforward network with two populations or layers that receives the stimulus and computes a Bayes-ratio encoding of the posterior distributions over z(1) and z(2). Assume that the first layer is implementing Bayes-ratio coding for the lower-level latent variables z(1) (e.g., using divisive normalisation Eq. (20)). We show in App. B.9 that if the feedforward synaptic weights into the second layer are given by , the downstream population implements Bayes-ratio coding for z(2) with linear neurons. This result is significant for three reasons. Firstly, while the downstream population is part of a recognition model that represents the posterior distribution of z(2), the synaptic weights needed to implement this representation only require knowledge of the generative probabilities (of z(1) given z(2)). Secondly, the synaptic weights make no reference to the prior distribution over z(2). There is therefore no need to adapt the weights upon environmental changes that only affect the statistics of the higher-level latent causes z(2) but not the generative process (as encoded by f). Thirdly, this result can clearly be extended to a multi-level generative model by induction.

Homeostatic DDCs account for stimulus specific adaptation

We found that homeostatic DDCs can provide a normative account of stimulus specific adaptation (SSA). Here, we start by showing this in the special case of Bayes-ratio coding, which is mathematically simpler and provides good intuitions for a more general subset of homeostatic DDCs. We then discuss this more general case and build a homeostatic DDC model of empirically observed SSA in the primary visual cortex (V1) (Benucci et al., 2013).

According to Eq. (18), for Bayes-ratio codes the tuning curve of neuron a is given by ha(s) = χf (s | za)/P (s). Suppose now that the stimulus distribution in the environment, P (s), changes due to a change in the statistics of the latent causes, π(z), while the causal processes linking the latent causes and observed stimuli, as captured by f (s | z), remain stable. In this case, by Eq. (18), the tuning curve of all neurons are modified by a multiplicative factor given by the ratio of the P (s) in the old environment to that in the new environment. This will lead to a multiplicative suppression of the response of all neurons to stimuli that are over-represented in the new environment relative to the old one, in a way that is independent of the neurons’ identity or preferred stimulus. Such a suppression thus constitutes a pure form of stimulus specific adaptation. In particular, neurons do not reduce their responsiveness to stimuli that are not over-represented in the new environment. Thus, tuning curves are suppressed only near over-represented stimuli (and may in fact be enhanced near under-represented stimuli), leading to a repulsion of tuning curve peak from over-represented stimuli. This repulsion is a typical manifestation of stimulus specific adaptation; in V1, for example, orientation tuning curves display a repulsion from an over-represented orientation (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013).

Heretofore we have implicitly made a so-called ideal observer assumption by assuming that the internal generative model underlying the DDC or Bayes-ratio code is a veridical model of the stimulus distribution in the environment. We now provide a significant generalization of the above results to the more general and realistic case of a non-ideal-observer internal model. As we will see, in this case, the effects of adaptation on tuning curves are captured by both a stimulus-specific factor as well as a neuron-specific factor. For a non-ideal observer model, there will be a discrepancy between the environment’s true stimulus distribution, PE(s), and the marginal stimulus distribution predicted by the internal generative model, i.e., In App. B.10, we demonstrate that the Bayes-ratio tuning curves in this case are given by (instead of Eq. (18)) where Note that, for ideal-observer internal models, PE = PI, in which case the integral yields Fa = 1 due to the normalisation of f (s | z), and we recover Eq. (18). As we can see, the tuning curves neatly decompose into a “base tuning curve” χf (s | za) multiplied by a stimulus-specific factor PI (s)−1 and a neuron-specific factor . Consequently, if we consider a pair of environments indexed by ν = 0, 1, and assume the likelihood function of the internal model does not change between environments, the transformation of the tuning curves due to adaptation will be given by Eq. (24), Thus, adaptation from environment ν = 0 to environment ν = 1 causes the tuning curves to transform via multiplication by a stimulus specific factor PI (s; ν = 0)/PI (s; ν = 1) and a neuron specific factor Fa(ν = 0)/Fa(ν = 1). Now suppose an environmental change results in an increase in the prevalence of an adaptor stimulus. As long as the internal model is a sufficiently good (if not perfect) model of PE(s), it should adapt its prior distribution, π(z), in a way that results in an increase of its predicted stimulus distribution, PI (s), around the adaptor stimulus. This results in a stimulus specific factor that goes below 1 around the adaptor s, thus causing suppression and repulsion of tuning curves away from the adaptor. Now consider a neuron whose preferred stimulus (roughly represented by za) is close to the adaptor stimulus. We further make the assumption that the change in the internal model, in response to the environmental change, is sufficiently conservative such that it leads to a smaller increase in PI (s) around the adaptor, compared with the increase in the veridical PE(s) (this is a reasonable assumption especially for artificially extreme adaptation protocols used in experiments). In this case, we expect the ratio PE(s)/PI (s) to increase in the support of f (s | za), which (by our assumption about the preferred stimulus za) is near the adaptor. According to Eq. (23) this will lead to an increase in Fa, and hence a suppressive neuron-specific factor Fa(v = 0)/Fa(v = 1) for neurons with preferred stimulus near the adaptor. In accordance with this picture, Benucci et al. (2013) found that homeostatic and stimulus-specific adaptation in V1 can indeed be modelled via separable stimulus-specific and neuron-specific factors.

Note that the assumption of constancy of the likelihood across the two environments need not be framed as a veridical assumption about the objective environmental change, but rather as an assumption about the inductive biases of the internal generative model, according to which it tends to model changes in the observed stimulus distribution, PE(s), as primarily resulting from changes in the statistics of the latent causes in the environment (which the model captures in π(z)), rather than in the causal process itself, as modelled by f (s | z). As long as this is good enough an assumption to result in a PI (s) which changes in similar ways to PE(s) (and in particular exhibits an increase around adaptor stimuli), the conclusions reached above will hold.

In the case of DDC codes that use more general kernel functions, other than the delta functions underlying Bayes-ratio codes, we arrive at an analogous transformation, given by Eq. (140) in App. B.10. However, in this case, neuron-specific and stimulus-specific effects are mixed, and the wider the DDC kernel the stronger the mixing. Nevertheless, for kernels that are unimodal and sufficiently narrow, we expect an approximate factorisation of the effect of adaptation into a stimulus-specific and a neuron-specific suppression factor, as seen in Eq. (24).

We now apply this result to the experiments performed by Benucci et al. (2013). These experiments examined the effects of adaptation on orientation-tuned neurons in the cat primary visual cortex. Anaesthetised cats were shown a sequence of oriented gratings chosen randomly from a distribution that had an increased prevalence (3 to 5 times more likely) of one particular orientation (arbitrarily defined to be to 0°). A control group was exposed to a uniform distribution of gratings. After about 2 seconds (on the order of 50 stimulus presentations), V1 tuning curves had adapted, and exhibited both the suppressive and repulsive effects mentioned above.

To model these findings we constructed a homeostatic DDC model as follows, and fit it to the data from Benucci et al. (2013). We took the model’s stimulus and latent variable spaces to be the circular orientation space [−90, 90), and we took the model’s generative likelihoods to be given by translation invariant functions f (s | z) = f (s − z), proportional to the Gaussian distribution with standard deviation σf, normalised over the circle (More rigorously, we ought to use a wrapped (periodic) normal distribution. However, since the standard deviations are small compared to the length of the circle, the normalisation constant is approximately 1, and we can treat the density as a normal Gaussian.). Likewise, we take the DDC kernel functions ϕa(z) = ϕ (z − za) proportional to the Gaussian distribution with standard deviation σϕ.

The distributions of stimulus orientations used in the experiment is highly artificial, in the way they jump discontinuously near the adaptor orientation. The internal prior distribution employed by the brain is more likely to be smooth, reflecting the brain’s inductive bias adapted to natural environments. Thus in our model we chose the internal prior to be a smooth distribution. The orientation distribution used in the experiment can be understood as a mixture of a uniform distribution, and a spike concentrated at the adaptor (Fig. 11A, blue curve). To obtain the smooth internal prior in our model, we replace the spike component of the experimental orientation distribution with a Gaussian distribution centred at the adaptor with standard deviation σπ (Fig. 11A, red curve). Thus, our model has only three free parameters, σf, σϕ and σπ, i.e., the widths of the generative likelihood, kernel functions, and the internal orientation prior, respectively.

Components of the model for stimulus-specific adaptation in V1.

(A) The distribution of stimulus orientations used in the experiments (Benucci et al., 2013), for an adaptor probability of 30% is shown in blue. We assume V1’s internal model use continuous prior distributions and thus we used instead a mixture of the uniform distribution with a smooth Gaussian, with standard deviation σπ, centered at the adaptor orientation (red curve). (B) The baseline tuning curves, i.e., the tuning curves adapted to the uniform orientation distribution. The blue curve was obtained by averaging the experimentally measured tuning curves adapted to the uniform orientation distribution, after centering those curves at 0°. The model’s baseline tuning curves (red) are given by the convolution of the Gaussian likelihood function, with width σf, of the postulated internal generative model, and the Gaussian DDC kernel, with width σϕ, and thus are themselves Gaussian with width . We fit this quantity to the experimentally observed baseline curves as described in Sec. 4.5.

Using the dataset from Benucci et al. (2013), we found the baseline tuning curves, i.e., the tuning curves of the population adapted to the uniform orientation distribution, by assuming that neural response is a function only of the difference between the preferred orientation and stimulus orientation (see Sec. 4.5). We then calculate the changes in the neurons’ preferred orientations (due to adaptation and the resulting tuning curve repulsion) in each of the experimental conditions (corresponding to different levels of over-representation of the adaptor stimulus). We then fit our model’s three parameters to both the unadapted tuning curves (Fig. 11B) and the changes in preferred orientation (see Sec. 4.5 for further details). The best fit values for the model parameters were found to be σπ = 18°, σf = 14.2°, and σϕ = 20°. The predictions of our model are compared to experimental data in Fig. 12. We found that the adaptive tuning curves within our model display suppression and repulsion away from the adaptor stimulus (Fig. 12A,D). In both the experimental data and our model, unadapted tuning curves exhibit a heightened average firing rate near the adaptor stimulus (Fig. 12B,E, blue curves). However, adaptation suppresses responses near the adaptor (Fig. 12A,D, red curves) to just the right degree to return the firing rates to the value before the environmental shift, leading to homeostasis of firing rates (as demonstrated by the uniformity of the red curves in Fig. 12B,E). Lastly, our model recapitulates the repulsion of preferred orientations found experimentally (Fig. 12C,F). Repulsion here is reflected by the change in preferred orientation having the same sign as the difference between the pre-adaptation preferred orientation and the adaptor orientation. As seen, in both the experiment and the model, repulsion is strongest for neurons with pre-adaptation preferred orientation about 30° from the adaptor.

A homeostatic DDC model accounts for the observations of stimulus-specific adaptation in V1 Benucci et al. (2013).

Panels A-C (left column) and E-F (right column) correspond to the experimental results and the predictions of our model, respectively. (A, D) The tuning curves for the adapted (red) and unadapted (blue) populations, averaged across experimental conditions. (B, E) The trial-averaged firing rates of the adapted (red) and unadapted (blue) populations exposed to the non-uniform stimulus ensemble. The trial-averaged population responses to the uniform stimulus ensemble were normalised to 1. (C, F) The repulsion of preferred orientations, obtained from the average tuning curves in panel A, as a function of the deviation of the neurons’ preferred orientation from the adaptor orientation.

In sum, the application of our homeostatic coding framework to a specific DDC model of Bayesian representations (which approximates what we have termed Bayes-ratio coding) explains quantitatively the empirical observations of both stimulus-specific and homeostatic adaptation in V1 (Benucci et al., 2013).

Discussion

We developed a theory of optimal gain modulation for combating noise in neural representations serving arbitrary computations. We demonstrated that — when mean neural firing rates are not too small; neurons are sufficiently selective and sparsely activated; and their responses form a high-dimensional geometry — the trade-off between coding fidelity and metabolic cost is optimised by gains that react to shifts in environmental stimulus statistics to yield firing rate homeostasis. Specifically, our theory predicts the homeostasis of the mean firing rate for clusters of similarly tuned neurons, while, at the same time, it accounts for the optimality of a broad distribution of firing rates within such clusters. By examining empirical estimates of parameters characterising cortical representations, we argued that the conditions necessary for the homeostatic adaptation to be optimal are expected to hold in the visual cortex and potentially other cortical areas. We further validated our approximation by demonstrating numerically that it performs well compared to the calculated optimal gains. Having developed a normative theory of neural homeostasis divorced from the computational aims of the neural population, we next showed how our theory can account for stimulus specific adaptation when coupled with distributed distributional codes (DDC). In particular, we focused on a special case of homeostatic DDC codes which we termed Bayes-ratio coding and showed that this model can account for stimulus specific adaptation effects empirically observed in the primary visual cortex. In the following, we will situate our work within the broader context of efficient coding theory, Bayesian neural representations, and alternative theories of stimulus specific adaptation. We will also discuss the shortcomings of the present work and possible directions for extension.

Within efficient coding theory, our work is closely related to that of Ganguli and Simoncelli (2014), but differs from it in important ways. Ganguli and Simoncelli (2014) considered a population of tuning curves, defined on a one-dimensional stimulus space, which was homogeneous and translationally invariant up to a warping (i.e., reparametrization) of the stimulus space and possible variations in gain. Their normative framework then optimised the neural gains as well as the stimulus warping (i.e., the reparametrization mapping from the original stimulus parameter to one over which the tuning curve population is homogeneous) to maximise a similar objective function to ours. Specifically, their objective employed the Fisher Information lower bound on the mutual information (Brunel and Nadal, 1998), as opposed to the Gaussian upper bound we used; our choice of the Gaussian upper bound was motivated by increased analytic tractability in the multidimensional stimulus case covered by our theory. In the case of unimodal tuning curves, they showed that the optimal warping amounts to a concentration and sharpening of tuning curves around stimulus values that are over-represented in the environment. On the other hand, in their setting, the optimal gains stay constant and do not adapt. Nevertheless, their solution also exhibits homeostasis of single-neuron firing rates.

Our framework extends and generalises that of Ganguli and Simoncelli (2014) in different ways. In their framework, the stimulus space is one-dimensional, and rigid constraints are placed on the shape and arrangement of the tuning curve population (they consider only unimodal or sigmoidal tuning curves, and assume that, up to a reparameterization of the stimulus space, all tuning curves have the same shape and are homogeneously placed). Wang et al. (2016) and Yerxa et al. (2020) have generalised the framework of Ganguli and Simoncelli (2014) to the case of alternative coding fidelity objective functions and high-dimensional stimuli, respectively; however, both of these studies maintain the same rigid constraint over the shape and arrangement of the tuning curve population. In contrast, our framework is applicable to tuning curves with heterogeneous (in particular multi-modal) shapes and arbitrary arrangement over a high-dimensional stimulus space; our homeostatic solution only requires sufficiently high signal-to-noise ratio (which is also the condition for the tightness of the mutual information lower-bound employed in Ganguli and Simoncelli (2014) or the approximation to the reconstruction-error loss used by Wang et al. (2016)). On the other hand, we only optimised the neural gains, and not the shapes or arrangement of the tuning curves, as we assumed the latter are determined by the computational goals of the circuit and not by the aim of optimally combating coding noise; in this way, our framework is agnostic to the computational goals of the population. Additionally, Ganguli and Simoncelli (2014) only considered the case of uncorrelated Poissonian noise. In contrast, our framework can handle correlated noise with general (non-Poissonian) power-law scaling. Furthermore, we analytically calculated first-order corrections to the homeostatic solution that arises in the high signal-to-noise ratio limit.

As noted above, the theory of Ganguli and Simoncelli (2014) predicts that tuning curves optimised to provide an efficient code under a stimulus distribution with over-represented stimuli cluster around those stimuli. This has been used to explain asymmetries in neural selectivities, such as the over-abundance of V1 neurons with preferred orientations near cardinal orientations, which are more abundant in natural scenes; or more generally, the aggregation of tuning curves around stimuli that are more prevalent in the natural environment. Many of these asymmetries have likely resulted from adaptation at very long (e.g., ontogenetic, or possibly evolutionary) timescales; however, long-term exposure on the order of few minutes to adaptor stimuli has also been found to result in the attraction of tuning curves towards adaptor stimuli (Ghisovan et al., 2009). This is in contrast to the repulsive effects seen in short-term adaptation (Movshon and Lennie, 1979; Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013). It is likely that the effects of long-term exposure are the results of mechanisms which operate in parallel to those underlying short-term adaptation, with both types co-existing at different timescales. We can therefore interpret Ganguli and Simoncelli (2014) as a model of adaptation to natural stimulus statistics at long timescales. Our results obtained here are applicable to adaptation on shorter timescales (Müller et al., 1999; Dragoi et al., 2000, 2002; Benucci et al., 2013). In particular, when married with Bayesian theories of neural representation, our framework predicts repulsive effects around an adaptor stimulus, as shown in Sec. 2.9.

Westrick et al. (2016) also propose a model of the experimental findings of Benucci et al. (2013). Their model (which is not explicitly normative or based on efficient coding) uses divisive normalisation (Carandini and Heeger, 2012) with adaptive weights to achieve homeostasis and stimulus specific repulsion. As discussed above in Sec. 2.8, Bayes-ratio coding (a special case of homeostatic DDCs, which we showed can account for the findings of Benucci et al. (2013)) can be accomplished by such a divisive normalisation scheme. Our framework therefore yields a normative interpretation of the model of Westrick et al. (2016), and links divisive normalisation with Bayesian representations and efficient coding theory. In particular, our theory provides an interpretation of the normalisation weights of the divisive normalisation model as the internal prior over latent causes of sensory inputs. This prior should naturally adapt as stimulus statistics change across environments. Additionally, the feedforward inputs to the divisive normalisation model are interpreted as the likelihood of stimuli given latent causes.

Snow et al. (2016) developed two normative models of temporal adaptation effects in V1, based on statistics of dynamic natural visual scenes (natural movies). These models are based on generative models of natural movies in the class of mixtures of Gaussian scale mixture (MGSM) distributions. In MGSMs the outputs of linear oriented filters applied to video frames are assumed to be Gaussian variables, multiplied by positive scale variables. These scale variables can be shared within pools of filter outputs at different times (video frames) and orientations. The two models of Snow et al. (2016) differ according to which pools of filter outputs are able to share scale variables. In both models, V1 neural responses are assumed to represent the inferred Gaussian latent variables of the corresponding MGSM. While each model was able to account for some of the findings of Benucci et al. (2013), neither one could account for all. In particular, each of their two models could only account for either stimulus-specific or neuron-specific adaptation factors found by Benucci et al. (2013). In particular, only the model that accounted for neuron-specific adaptation was able to produce firing rate homeostasis. As we showed in Sec. 2.9, our framework, which combines homeostatic gain modulation with the DDC theory of representation, robustly accounts both for stimulus-specific adaptation and firing rate homeostasis. Furthermore, when there is a discrepancy between the internal model’s stimulus prior and the true environmental stimulus distribution, it additionally accounts for a neuron-specific adaptation factor as well (see Eq. (24)).

Tring et al. (2023) replicates the experiment performed in Benucci et al. (2013), this time in awake mice (rather than anesthetised cats). However, in contrast to the results of Benucci et al. (2013), Tring et al. (2023) do not find firing rate homeostasis in mouse V1. Additionally, Tring et al. (2023) find that adaptation mainly changes the scale of the population response vector, while minimally affecting its direction. Mathematically, this amounts to pure stimulus-specific adaptation without the neuron-specific factor found in by Benucci et al. (2013) - see Eq. (24). As we discuss in Sec. 2.7, a Bayes-ratio code based on an ideal-observer generative model also displays pure stimulus-specific adaptation with no neuron-specific factor. Our final model for Benucci’s data did contain a neuronal factor, because we used a non-ideal observer DDC (specifically, we assumed a smoother prior distribution over orientations compared to the distribution used in the experiment — which has a very sharp peak — as it is more natural given the inductive biases we expect in the brain). The resultant neuron-specific factor suppresses the tuning curves tuned to the adaptor stimulus. Interestingly, when gain adaptation is incomplete, and happens to a weaker degree compared to what is necessary for firing rate homeostasis, an additional neuronal factor emerges that is greater than one (i.e., facilitatory) for neurons tuned to the adaptor stimulus. These two multiplicative neural factors can potentially cancel each other approximately; such a theory would thus predict both deviation from homeostasis and approximately pure stimulus-specific adaptation. We plan to explore this possibility in future work.

Benucci et al. (2013) find that in cases where the adaptor probability is particularly extreme (i.e., 50%), homeostasis is imperfect, and neurons tuned to the adaptor display higher firing rates. These results cannot be accounted for within our homeostatic DDC framework, since we take as a starting point homeostatic regulation of firing rates in our model. However, note that our efficient coding framework does predict deviation from homeostasis when signal-to-noise ratio falls (see Sec. 2.4). Thus, a more complete model in which we both specify DDC representational curves and numerically optimise gains according to our efficient coding objective may be able to account for the observed deviation from homeostasis.

Our approach has relied on a number of assumptions and simplifications; relaxing or generalising these assumptions provide opportunities for future research. Firstly, our toy model (see Sec. 2.3) imposed a cluster structure on the neural tuning curves, whereby neurons within a cluster are similarly tuned, while neurons in different clusters have distinct tuning. We found that total firings rates are equalised across clusters (Sec. 2.2), but the optimal rates of individual neurons within clusters span a broad range (Sec. 2.3). This results from the fact that the efficient coding objective function depends sharply on total cluster firing rates, but changes only slightly when the partition of the cluster firing rate among the cluster’s neurons changes. In reality, tuning similarity has a more graded distribution. As the distinction between clusters with different tuning (or similarity of tuning within clusters) weakens, higher order terms in our expansion of the objective Eq. (27) cannot be neglected, and hence the justification for dividing the problem into two separate, cluster-level vs. within-cluster, problems breaks down. However, arguably, in a more realistic model with a more graded variation in neural signal correlations, the optimisation landscape will still contain shallow directions (i.e., directions over which the objective changes slowly), corresponding to partitions of firing among similarly tuned neurons, and non-shallow directions corresponding to total firings of such groups of neurons; if so, we would expect a similar solution to emerge, whereby similarly tuned neurons display heterogeneous firing rates, with the total firing rate of such a group of neurons displaying approximate homeostasis.

Secondly, we showed analytically that, at the level of cluster responses, homeostasis emerges universally in the high signal-to-noise regime. In general, firing rate homeostasis is no longer optimal in the low signal-to-noise ratio limit, as correction terms to our solutions become large. However, as we argued in Sec. 2.5, based on empirical values we expect the signal-to-noise ratio within cortex to be sufficiently high for our results to hold. Generally, within efficient coding theory (except for simple cases such as linear Gaussian models), the regime of arbitrary signal-to-noise ratio is analytically intractable. Many approaches are therefore limited to the regime of high signal-to-noise ratio. For example, the Fisher Information Lower Bound used frequently within efficient coding theory breaks down outside of this regime (Yarrow et al., 2012; Bethge et al., 2002; Wei and Stocker, 2016). Thus, inevitably, exploring the low signal-to-noise ratio regime would require numerical simulation. However, such numerical simulations require concrete models of signal and noise structure. The manner in which optimal coding deviates from homeostatic coding will not be universal across these models, but depend on each model’s specific details. This limits our ability to draw general conclusions from numerical simulations, which is why we have chosen not to pursue that strategy extensively here.

Thirdly, our analysis here considered a specific class of noise models. These noise models include those of particular biological relevance, such as information-limiting noise correlations (Moreno-Bote et al., 2014; Kanitscheider et al., 2015), signatures of which have been found in the cortex (Rumyantsev et al., 2020), and power-law variability (Goris et al., 2014) (see Sec. 2.6). As we noted, our homeostatic solution is particularly well suited to the case of information-limiting noise correlations, performing optimally in that case (see App. B.5). The core ideas of our framework and derivation could potentially be applied to an even wider class of noise models.

Fourthly, up to Sec. 2.7, we make no assumptions about what the neural population is attempting to represent via the representational curves. We then apply our framework to a specific theory of Bayesian encoding (namely DDCs), and develop the new idea of Bayes-ratio coding. There is therefore an opportunity to apply our framework to other theories of neural representation, and in particular to alternative theories of Bayesian representations, such as Probabilistic Population Codes (PPCs) (Beck et al., 2007).

Finally, it is possible to generalise our theoretical framework (both the general efficient coding framework and its combination with frameworks for Bayesian inference) to the case of temporally varying stimuli and dynamic Bayesian inference. In particular, Młynarski and Tkacik (2022) have applied optimal gain modulation, according to efficient coding (to reduce metabolic cost), to sparse coding (an example of a probabilistic model with latent variable representation), in order to model gain modulation by attention. Accordingly, their efficient coding objective was task-dependent (i.e., depended on the choice of latent variables that are desired to be in the focus of attention). However, we are not aware of similar work that has applied such a combination to model the dynamics of task-independent (i.e., bottom up, rather than top-down) adaptation.

In summary, we showed that homeostatic coding can arise from optimal gain modulation for fighting corruption by noise in neural representations. Based on this coding scheme, we derived a novel implementation of probabilistic inference in neural populations, known as Bayes-ratio coding, which can be achieved by divisive normalisation with adaptive weights to invert generative probabilistic models in an adaptive manner. This coding scheme accounts for adaptation effects that are otherwise hard to explain exclusively based on efficient coding theory. These contributions provide important connections between Bayesian representation, efficient coding theory, and neural adaptation.

Methods

Specification of environments for simulations

For our numerical simulations, we used K = 10, 000 units. To generate the K × K signal correlation matrix, ρ(ν), for the different environments in our numerical simulations, in Sec. 2.2, we first generate a covariance matrix Σ(ν), and let ρ(ν) be the corresponding correlation matrix. The covariance matrix Σ(ν) is in turn constructed such that it has (1) the same eigenspectrum for all ν, with eigenvalues that decrease inversely with rank, as observed in V1 (Stringer et al., 2019), and (2) an eigenbasis that varies smoothly with ν. Thus we defined Σ(ν) = U (ν)Λ1U (ν)T, where Λ1 = diag(1, 1/2, 1/3, …, 1/K) and U (ν) is an orthogonal matrix that was generated as follows. We randomly and independently sample two K×K random iid Gaussian matrices R0, R1 ~ 𝒩 K×K(0, IK×K) and obtain symmetric matrices and . As is well known, the eigen-basis (represented by an orthogonal matrix) of such a random Gaussian matrix is distributed uniformly (i.e., according to the corresponding Haar measure) over the orthogonal group O(K). For the extreme environments, ν = 0 and 1, we let U (ν) be the matrix of eigenvectors of S0 and S1, when ordered according to eigenvalue rank. For 0 < ν < 1, we let U (ν) be the matrix of the eigenvectors (again ordered according to eigenvalue rank) of the interpolated symmetric matrix S(ν) = (1 − ν)S0 + νS1. Thus, for the extreme environments the eigenbases of the covariance matrices Σ(0) and Σ(1) are sampled independently and uniformly (i.e., from the Haar measure), and the eigenbases of Σ(ν) for intermediate environments (i.e., for 0 < ν < 1) smoothly interpolate between these.

In the case of simulations for aligned noise, we also generate a noise correlation matrix W (ν) which has an approximately 1/nγ spectrum. Ideally, W and ρ would have the same eigen-basis. However, this is impossible, since W and ρ are both correlation matrices. Instead, we generate W (ν) by normalising the positive definite matrix U (ν)ΛγU (ν)T where Λγ = diag(1, 1/2γ, 1/3γ, …, 1/Kγ)

Analysis of the clustered population

Here, we provide a summary of the analysis of the clustered model introduced in Sec. 2.3 (for derivations and a more detailed account see App. B.3). In this model, a population of N neurons are sorted into K clusters. Neurons within a cluster have very similar stimulus tuning, but neurons in different clusters are tuned differently. More concretely, we will assume that the representational curves of the neurons, indexed by i, belonging to cluster a, are small perturbations to the same representational curve characterising that cluster; i.e., where ϵ is the perturbation strength parameter controlling the deviation from perfect within-cluster similarity. Below, hatted symbols denote variables defined analogously to Eq. (7)–(8) but for single neurons, with Ωa(s) replaced by .

We assume single-neuron gains adapt to optimise an objective ℒ pop, defined analogously to the objective ℒ, Eq. (9), but for individual neurons rather than clusters. In the case of uncorrelated noise with Poisson-like scaling, this objective is given by where are the mean firing rates of individual neurons, given by ; note that maximising ℒpop in terms of the single-neuron gains is equivalent to maximising it in terms of . In App. B.3, we show that this objective can be expanded in ϵ as Here, the leading order term, ℒ (r = ωg), is a cluster-level objective and is given byLEq. (9), now viewed as a function of the cluster rates ra, i.e., the total mean firing rate of neurons in a cluster; thus where Ca is the set of neurons in cluster a. Note that this term is a function only of the cluster firing rates. Thus, for ϵ = 0 (perfect similarity within clusters), the efficient coding objective function is blind to the precise distribution of single-cell firing rates, , among the neurons of a cluster, as long as the total rate of the cluster is held fixed. This is because for clusters of identically-tuned and independently-firing neurons with Poisson-like noise, the cluster can be considered as a single coherent unit whose firing rate is given by the sum of the firing rates of its neurons. Thus at zeroth order, the distribution of individual neuron rates are free and undetermined by the optimisation, as long as they give rise to optimal cluster rates.

According to Eq. (27), at small but non-zero is still dominated by ℒ; we thus expect this term to play the dominant role in determining the cluster rates, with terms of order ϵ2 and higher having negligible effect. However, the perturbation term breaks the invariance of the loss with respect to changes in the distribution of single neuron gains (or rates) within clusters. We will therefore approximate the maximisation of the total objective function ℒpop as follows. We first specify the cluster rates by maximising ℒ (r), obtaining the optimal cluster rates . We then specify the rates of individual neurons within the population by maximising the leading correction term in the objective, , subject to the constraint that the single-neuron rates in each cluster sum to that cluster’s optimal rate. We reinterpret the optimisation in the previous sections to correspond to the first stage of this two-stage optimisation, with population units there corresponding to clusters of similarly tuned neurons. As we have seen in Sec. 2.2, optimisation of the cluster-level objective, ℒ, can result in homeostasis and uniformisation across clusters.

As for the distribution of individual neuron rates, we show in App. B.3.5 (see the derivation of Eq. (87)) that, in the parameter regime (see Secs. 2.4 and 2.5) in which cluster-level homeostasis and uniformisation occurs, the perturbation objective ℒ pert decomposes into a sum of objectives over different clusters. Thus, in the second stage of optimisation the problem decouples across clusters, and (as shown in App. B.3.5) the optimal single-neuron rates in cluster a maximise the objective subject to the constraint Here ΔΩi is defined to be the centered, zero-mean version of δΩi: (We can interpret ΔΩi as the direction of the perturbed tuning curve in excess of the cluster tuning curve Ωa.) Note that the optimisation problem, Eqs. (28)–(29), is a quadratic program. Moreover, the terms in the objective for this quadratic program are homogeneous in the sense that if the information-energy trade-off parameter µ is scaled a factor, the optimal solutions for scale with the same factor. In our cluster-level simulations Sec. 2.2 we used µ = 10 (in agreement with our biological estimates in Sec. 4.4 for mean spike count of clusters of 20 V1 neurons in a coding interval of 0.1 seconds). That dimensionless µ (which in the homeostatic approximation determines the mean spike count of clusters over the coding interval) thus corresponds to a total cluster rate of 100 Hz, and using this value for µ in Eqs. (28)–(29) the solutions for can be interpreted as being firing rates in units of Hz.

Within-cluster optimisation

To calculate the single-neuron firing rate distributions in Fig. 8, we solved the optimisation problem Eqs. (28)–(29) numerically for 10,000 randomly generated covariance matrices (note that different instances of the optimisation can be thought of as characterising different clusters within the same population). In general, the covariance matrix, (which defines this optimisation problem), depends on the stimulus distribution, the cluster tuning curve, and properties (e.g., degree of smoothness) of the intra-cluster variations in single-cell tuning curves. In our simulations, instead of making specific arbitrary choices for these different factors, we used a minimal toy model where the covariance matrix was generated as the matrix of inner products of k random vectors in a D-dimensional space; specifically, we generated this matrix as the the Gram matrix of k random D-dimensional vectors with isotropic Gaussians distributions, i.e., , where δwi ~ 𝒩 (0, ID). In App. B.3.6 we show that D can be interpreted as the dimensionality of the space of independent perturbations to the cluster-wide tuning curve Ωa which vary significantly on the portion of stimulus space to which the cluster tuning curve responds (below, we will refer to D as the effective tuning curve dimension). Additionally, we fixed the cluster rate to µ = k × 5Hz, such that the average firing rate per single neuron is 5Hz, and fixed σ2 = 1 (corresponding to a Fano factor of 1). For various values of k and D, we used the above method to generate 10, 000 samples of the covariance matrix Cov(ΔΩ). For each such sample, we then solved the problem (28–29) numerically using the cvxpy package in Python (Diamond and Boyd, 2016). Neurons with firing rates less than 10−4Hz had their firing rate set to zero. We aggregated the set of non-zero rates obtained from each of these optimisations to create the histograms in Fig. 8.

Estimation of biological parameters justifying the homeostatic approximation

In this section we provide the details for our estimates of cortical mean firing rates and noise scaling used in Sec. 2.5.

Firing rate estimates

A wide range of mean firing rates for individual neurons have been reported in cortex. Here we focus on firing rates in rodent V1 during free behaviour (which tends to be lower compared to rates in cat or monkey cortex). Reported values tend to range from 0.4 Hz (Greenberg et al., 2008) to 14 Hz (Parker et al., 2022), depending on the cortical layer or area or the brain state, with other values lying in a tighter range of 4-7 Hz (Hengen et al., 2013; Szuts et al., 2011; Torrado Pacheco et al., 2019). Therefore, for rodent V1, we take a mean firing rate of 5 Hz to be a reasonable rough estimate.

Scaling of noise variance with mean response

A number of studies (Shadlen and Newsome, 1998; Gershon et al., 1998; Moshitch and Nelken, 2014; Koyama, 2015) have characterised the relationship between the trial-to-trial mean and variance of neural spike counts as a power law of the form and have provided best fit estimates of and β in different species and cortical areas. We have summarised these in Table 1.

While the power-law exponent β is directly identifiable with the same parameter in our noise model, we need to adjust the empirical estimates of in Table 1 in order to obtain estimates for our model parameter σ, for two reasons. First, the duration of the time interval in which spikes were counted in the above papers differs from our assumed coding interval of 0.1s. We therefore must adjust the reported values to be for an interval of this size. We do this by assuming temporal homogeneity over the timescales of interest; specifically we assume that mean and variance are linear in the time interval. Second, we must also adjust the data to account for the fact that our responses are for clusters of k = 20 similarly tuned neurons. We do this by assuming uncorrelated response noise within each cluster, making mean and variance both linear functions of cluster size. Our adjustments do not affect the value of β, but do change the value of the pre-factor Specifically, we perform the transformation: where δt is the reported time interval, and Δt is our coding interval, Δt = 0.1s. Fig. 13 shows the adjusted values on a scatter plot. Using the adjusted values, σ2, and assuming that µβ = 10 (to maintain a firing rate of 5Hz per neuron), we can find the value of σ2/(βµ)β. These are reported in the furthest right column of Table 1.

Empirically estimated values of the noise scaling parameter β and the base noise-level σ2, after adjusting for coding interval duration and cluster size using the formula Eq. (31).

The dashed lines show the values for Poissonian noise (σ2 = 1, β = 1).

Sufficiently selective responses

The coefficient of variation CV is a measure of neural selectivity, and therefore is related also to measures of sparseness of neural responses, or so-called coding level. To see this, consider a toy model in which the cluster responds at a fixed level to a fraction p of stimuli and is silent otherwise. In this case, CV2 = (1−p)/p ≈1/p for small p. Our condition therefore requires that neurons are sufficiently selective in their responses and respond only to a small fraction of stimuli. Lennie (2003) places the fraction of simultaneously active neurons (which we use as a proxy for the response probability of a single cluster) at under 5%. For our toy model, this yields the estimate CV2 ≈ 20. The binary distribution in the toy model is particularly sparse, and so we take CV2 ≈ 10 as a more conservative estimate.

High-dimensional signal geometry

Assuming, for simplicity, that the coefficients of variation are approximately the same for all clusters, we see that the expression for Δaa are proportional to the diagonal elements of ρ−1W. To estimate the latter, we will assume that these elements are all comparable in size (This is expected to be valid when the eigenbasis of ρ−1W is not aligned with the standard (cluster) basis, which, in turn, corresponds to distributed coding.), and will therefore estimate their average value, which is given by the normalised trace . By the invariance of trace, this can equally be characterised as the mean eigenvalue of ρ−1W. We estimated this trace for the aligned noise family (corresponding to information-limiting noise correlations, but also including uncorrelated noise) which we numerically studied in Sec. 2.2 and Sec. 2.6, and a biologically-motivated form of ρ used there. In the aligned noise model, W and ρ share the same eigenbasis, but have different eigenvalue spectra. Our choice of ρ has spectrum that scales like 1/n with the rank of the eigenvalue (based on the findings of Stringer et al. (2019) that the signal covariance matrix of mouse V1 neural populations responding to natural stimuli possesses an approximately 1/n spectrum), while the spectrum of W scales like 1/nγ for a positive exponent parameter γ.

We show in App. B.6 that, for 0 ≥ γ ≥ 1 (relatively high-dimensional and weakly correlated noise), we have . The upper bound is saturated in the case of uncorrelated noise (which correspond to γ = 0 yielding W = I). In the other extreme where γ = 1, we have W = ρ or ρ−1W, which yields a normalised trace of 1; which does not scale with K. On the other hand, for γ > 1 (relatively low-dimensional and strongly correlated noise), tr(ρ−1W) decays like K− min(γ−1,1) as K grows. Thus, in this case, the estimate of Δaa is smaller than that of Eq. (13). However, this case is arguably not biologically relevant, due to the fact that neural responses always contain some amount of “private” uncorrelated noise. Mathematically this would contribute an additive correction to W proportional to the identity matric. In the case γ > 1 (where noise eigenvalues decay faster than signal eigenvalues) this term, even when the size of the correction is small, prevents the inverse power-law scaling with K. Therefore we did not use this case in the Results section in our estimates of Δaa.

Fitting a homeostatic DDC to Benucci et al. 2013

The data we obtained from Benucci et al. (2013) comprised 11 datasets corresponding to different experimental sessions. In each dataset, the recorded neural population was clustered into 12 groups based on preferred orientation. The datasets contained the trial-averaged responses of these clusters to gratings of different orientation (which took one of 6 or 12 possible values depending on the dataset). In each experimental context, the grating orientations were randomly drawn from a distribution that was either uniform or biased; in the latter case, one particular grating (arbitrarily defined as 0°) had higher prevalence, by either 30%, 35%, 40%, or 50%. We discarded all datasets with a 50% prevalence, since (Benucciet al., 2013) report that homeostasis was less perfect in this case, and they similarly discarded it from their main analysis. After exposure to the distribution of gratings, the responses (i.e., spike count) of each cluster to a test set of 20 oriented gratings were then measured. In each dataset, the trial-averaged tuning curves had been normalised by an affine transform (see Benucci et al. (2013) for details) such that tuning curves in the uniform stimulus ensemble context ranged from 0 to 1. We further interpolated and up-sampled all normalised tuning curves using a cubic spline (Fig. 12A).

To fit our model parameters (σf, σπ, and σϕ), we started by centering the cluster tuning curves in the uniform ensemble context at 0° and averaged across clusters with different preferred orientations, yielding a single average tuning curve (Fig. 11B, blue). We fit the resultant curve with a Gaussian (Fig. 11B, red) with a standard deviation matched to 1/4 the width of this curve at height e−2. Subsequently, (which is the model’s prediction for the standard deviation of the Gaussian tuning curves in the uniform ensemble context) was constrained to equal this standard deviation. This reduces the number of free model parameters to 2, which we choose to be σπ and σf. We found the best value of these parameters by fitting the curves for repulsion of tuning curves (adaptation-induced change in preferred orientation vs. pre-adaptation preferred orientation – Fig. 12C and F) between the model and experimental data. Specifically, we performed a grid search over values of σπ and σf and chose the pair of values which minimised the sum of the absolute differences between the empirical and model curves.

We obtained the empirical repulsion curves (Fig. 12C) as follows. For each preferred orientation cluster and context (i.e., exposure to a uniform or biased ensemble), a smoothing kernel was applied before using cubic interpolation to generate up-sampled tuning curves. The smoothing kernel was applied to ensure that none of the tuning curves were multimodal (multimodality was a minor effect — reasonably attributable to finite-trial averaging noise — but could have nevertheless introduced relatively large noise in the estimated preferred orientations). The argmax of these tuning curves was found to give the preferred orientation. The preferred orientation was then compared across conditions (uniform ensemble vs. biased ensemble) to give a change in preferred orientation. These were then averaged across contexts with the same adaptor probability to obtain the repulsion curve for each adaptor probability.